### 2014-10-10 brb-wip-oct.Rmd
brb presentation

---
title: "A Critical examination of the ROC curve plot"
author: "Michael Sachs"
date: "`r Sys.Date()`"
output: 
    ioslides_presentation:
        widescreen: true
        self_contained: false
        fig_align: center
        fig_width: 6
---

<style type="text/css">
blockquote {
  background: #f9f9f9;
  border-left: 10px solid #ccc;
  margin: 1.5em 10px;
  padding: 0.5em 10px;
  quotes: "\201C""\201D""\2018""\2019";
  font-size: 18px;
}
blockquote p {
  display: inline;
}
</style>

```{r init, include = FALSE}
library(ggplot2)
library(ggthemes)
library(knitr)

bltab <- function(x, name){

  t1 <- data.frame(as.list(table(x)))
  rownames(t1)[1] <- name
  kable(t1)

}

rocme <- function(M, D){

c <- sort(M)
TPF <- sapply(c, function(x) mean(M[D == 1] > x))
FPF <- sapply(c, function(x) mean(M[D == 0] > x))

data.frame(cbind(c, TPF, FPF))

}
```

# Introduction

## What is an ROC curve?

Binary biomarker and Binary outcome

<style>
td { border: 1px #aaa solid;
     padding: .2em; 
     }
</style>

<table align="center" style="text-align:center; border:1px; background:transparent; font-size: medium;">
<tr>
<td colspan="2" style="border:none;"></td>
<td colspan="2" style="background:#eeeebb;"><b>Condition<br />
(as determined by "Gold standard")</b></td>
</tr>
<tr>
<td style="border:none;"></td>
<td style="background:#ddddcc;">Total population</td>
<td style="background:#ffffcc;">Condition positive</td>
<td style="background:#ddddaa;">Condition negative</td>
<td style="background:#ddddcc;">Prevalence =<br />
Σ&#160;Condition positive
<div style="border-top:1px solid;">Σ&#160;Total population</div>
</td>
</tr>
<tr>
<td rowspan="2" style="background:#bbeeee;"><b>Test<br />
outcome</b></td>
<td style="background:#ccffff;">Test<br />
outcome<br />
positive</td>
<td style="background:#ccffcc;"><span style="color:#006600;"><b>True positive</b></span></td>
<td style="background:#eedddd;"><span style="color:#cc0000;"><b>False positive</b></span><br />
</td>
<td style="background:#ccffff;"> Positive predictive value (PPV) =<br />
Σ True positive
<div style="border-top:1px solid;">Σ&#160;Test&#160;outcome&#160;positive</div>
</td>
</tr>
<tr>
<td style="background:#aadddd;">Test<br />
outcome<br />
negative</td>
<td style="background:#eedddd;"><span style="color:#cc0000;"><b>False negative</b></span><br />
</td>
<td style="background:#bbeebb;"><span style="color:#006600;"><b>True negative</b></span></td>
<td style="background:#aadddd;">Negative predictive value (NPV) =<br />
Σ True negative
<div style="border-top:1px solid;">Σ&#160;Test&#160;outcome&#160;negative</div>
</td>
</tr>
<tr>
<td style="border:none;"></td>
<td style="border:none;"></td>
<td style="background:#ffffcc;">True positive fraction (TPF, Sensitivity) =<br />
Σ True positive
<div style="border-top:1px solid;">Σ&#160;Condition&#160;positive</div>
</td>
<td style="background:#ddddaa;">False positive fraction (FPF) =<br />
Σ False positive
<div style="border-top:1px solid;">Σ&#160;Condition&#160;negative</div>
</td>
<td style="background:#ddddcc;">Accuracy (ACC) =<br />
Σ&#160;True positive + Σ True negative
<div style="border-top:1px solid;">Σ&#160;Total population</div>
</td>
</tr>
<tr>
<td style="border:none;"></td>
<td style="border:none;"></td>
<td style="background:#ffffcc;">False negative fraction (FNF) =<br />
Σ False negative
<div style="border-top:1px solid;">Σ&#160;Condition&#160;positive</div>
</td>
<td style="background:#ddddaa;">True negative fraction</a> (TNF, Specificity) =<br />
Σ True negative
<div style="border-top:1px solid;">Σ&#160;Condition&#160;negative</div>
</td>
</tr>
<tr>
<td style="border:none;"></td>
</tr>
</table>

## Continuous test

 - *test positive* is defined as $M > c$
 - Consider measures as functions of $c$, i.e.
    $$ TPF(c) = Pr\{M > c | D = 1\} $$
    $$ FPF(c) = Pr\{M > c | D = 0\} $$
 - Plot $FPF(c)$ versus $TPF(c)$

## ROC Curve

```{r rocmin, fig.width = 5, fig.height = 5, echo = FALSE}
D <- rbinom(250, 1, .3)
M <- rnorm(250, mean = D, sd = .75)

tour <- rocme(M, D)
p1 <- ggplot(tour, aes(x = FPF, y = TPF)) + geom_path() + theme_classic() + 
  scale_x_continuous("False positive fraction at c") + scale_y_continuous("True positive fraction at c") 
p1
```

## Design considerations

**Grid lines**

```{r des1, echo = FALSE}
p2 <- p1 + theme_bw()
p2
```

## Design considerations

**Reference line**

```{r des2, echo = FALSE}
p2b <- p2 + geom_abline(intercept = 0, slope = 1, lty = 2)
p2b
```

## Design considerations

**Labels/Legend**

```{r des3, echo = FALSE, fig.width = 4.5}
y1 <- max(tour$TPF[tour$FPF <= .25]) - .1
p3 <- p2b + annotate("text", x = c(.25, .5), y  = c(y1, .4), label = c("Marker", "Reference"), angle = 45)
p3leg <- p2b + geom_line(data = NULL, aes(x = c(0, 0), y = c(0, 0), linetype = c("Marker", "Reference"))) + scale_linetype_discrete("Type")
p3
p3leg
```

## Design considerations

**Cutoff values**

```{r cut, echo = FALSE}
subdat <- tour[seq(1, nrow(tour), length.out = 15),]
p4 <- p3 + geom_point(data = subdat, aes(x = FPF, y = TPF)) + 
  geom_text(data = subdat, aes(x = FPF - .01, y = TPF + .01, label = round(c, 1), hjust = 1, vjust = 0))
p4
```

## Design considerations

**Other annotations**

```{r auc, echo = FALSE}
p4 + annotate("text", x = .75, y = .25, label = "AUC = 0.71 (0.62 to 0.80)")
```

## Goals

**Evaluate the state of ROC curve plots**

+ How are they used
+ What stories do they tell
+ What tools are there to create them

## A bit of history

A common academic rumor is that ROC curves were created during WWII to assess the ability of radar operators to distinguish flocks of birds from aircraft.

I can't find any primary evidence of that.

  - ROC pioneers in signal detection: [Green and Swets, 1966](http://www.amazon.com/Signal-Detection-Theory-Psychophysics-Marvin/dp/0932146236), and [Egan 1975](http://books.google.com/books?id=V40oAAAAYAAJ&q=egan+roc+analysis+1975&dq=egan+roc+analysis+1975&hl=en&sa=X&ei=vE01VPvnDo-wyATwlYKYDg&ved=0CB0Q6AEwAA)
  - and radiology/medical decision making: [Lusted, 1978](http://www.sciencedirect.com/science/article/pii/S0001299878800154), [Swets and Pickett, 1982](http://www.amazon.com/Evaluation-Diagnostic-Systems-John-Swets/dp/0124122515)

## What were ROC curves used for?

> ROC curve describes possible relationships among the probabilities of the various types of correct and incorrect decisions. Therefore, it plays a central role in optimizing diagnostic strategies

> Conventional ROC curve describes the compromises that can be made between TPF and FPF as a decision threshold is varied. By appropriate choice of the decision threshold, a decision maker or observer can operate at (or near) any desired compromise that lies on the curve.

![old](figures/old-school.png)

# Literature review

## Methods

**Goal:** Assess the usage of the ROC plot in oncology studies

**Search criteria:** Clinical trials or observational studies in humans reported in major oncology journals for the past 10 years.

**Query:** "ROC Curve" OR "ROC Analysis" OR "Receiver operating characteristic curve"

**Results:** 57 Papers, 47 figures extracted and reviewed

[Pubmed query](http://www.ncbi.nlm.nih.gov/pubmed/?term=Cancer%5BText+Word%5D+AND+(ROC+Curve%5BText+Word%5D+OR+ROC+analysis%5BText+Word%5D+OR+Receiver+operating+characteristic+curve%5BText+Word%5D)+AND+((Clinical+Trial%5Bptyp%5D+OR+Observational+Study%5Bptyp%5D)+AND+%22loattrfull+text%22%5Bsb%5D+AND+%222004%2F10%2F05%22%5BPDat%5D+%3A+%222014%2F10%2F02%22%5BPDat%5D+AND+%22humans%22%5BMeSH+Terms%5D+AND+jsubsetaim%5Btext%5D))

## Results

```{r fig1, echo = FALSE, fig.height = 3}
rev <- read.csv("../Literature Review/rocusage.csv", stringsAsFactors = FALSE)
rev <- subset(rev, flag != "comedy")

ggplot(rev, aes(x = factor(number.of.curves))) + geom_bar(fill = "blue", color = "black") + 
  theme_igray() + scale_x_discrete("Number of curves on plot")
```

```{r tab1, echo = FALSE, results = 'asis'}
bltab(rev$type, "Purpose")
```

## Results: axis labels

 - Sensitivity = <emph class="blue">True</emph> <emph class="red">positive</emph> fraction
 - 1 - Specificity = <emph class="blue">False</emph> <emph class="red">positive</emph> fraction
 - $Pr\{\color{red}{M > c} | \color{blue}{D = 1}\}$
 - $Pr\{\color{red}{M > c} | \color{blue}{D = 0}\}$

```{r jargon, echo = FALSE, results = 'asis'}
bltab(rev$axis.names, "Axis type")
```

## Results: other features

```{r features, echo = FALSE, results = 'asis'}
kable(rbind(
  "Diagonal Guide" = table(rev$guide),
  "Grid lines" = table(rev$minor.axes),
  "AUC given" = table(rev$auc)  ))

bltab(rev$cutoffs, "Cut points")
bltab(rev$labels, "Curve labels")
```

# Highlights and lowlights

## Extreme minimalist

![min](figures/min1.gif)
![min](figures/min2.gif)

## Spaghetti plots

![spag](figures/spaghetti.jpg)

## Spaghetti plots

![spag](figures/spaghetti2.png)

## Tables in diguise

![table](figures/table.png)

## Almost useful

![good](figures/good1.jpg)

## Getting there

![good](figures/good2.png)

# Analysis

## Pure comparison of tests

  - Value of a biomarker
    + AUC/Partial AUC?
    + Most north-western point = Total Gain, maximum Youden Index

### ROC plots encode this information poorly

```{r ex, fig.width = 5, fig.height = 3, echo = FALSE}
D <- rbinom(250, 1, .3)
M1 <- rnorm(250, mean = D, sd = .75)
M2 <- rnorm(250, mean = D, sd = 1.75)

tour2 <- rbind(rocme(M1, D), rocme(M2, D))
tour2$Marker <- c(rep("New great", 250), rep("Old bad", 250))

ggplot(tour2, aes(x = FPF, y = TPF, color = Marker)) + geom_path() + geom_abline(intercept = 0, slope = 1, lty = 2) + 
  theme_igray() + scale_color_colorblind() + 
  scale_x_continuous("False positive fraction at c") + scale_y_continuous("True positive fraction at c") 
```

## A more direct way

```{r line, fig.width = 7, fig.height = 3, echo = FALSE}
dlin <- data.frame(AUC = sort(runif(5)), marker = sort(sample(LETTERS, 5)))
ggplot(dlin, aes(y = AUC, x = marker)) + geom_point(size = 3) + 
  geom_segment(aes(y = pmax(AUC - .12, 0), yend = pmin(AUC + .12, 1), x = marker, xend = marker)) + 
  scale_y_continuous("AUC", limits = c(0, 1)) + theme_igray() + coord_flip()
```

See the works of [Tufte](http://www.edwardtufte.com/tufte/books_vdqi) and [Cleveland](http://dl.acm.org/citation.cfm?id=529269)

## What good is the ROC plot then?

Enable the observer to

  - Accurately discern the operating charactersitics at any cutoff
  - Guide the choice of "optimal" cutoff
  - Determine what the cutoff is!

# Tools for ROC plot creation

## Currently available

#### Several R packages

  - [`ROCR`](http://rocr.bioinf.mpi-sb.mpg.de/) (Encodes cutoff values as color scale along curves)
  - [`pROC`](http://web.expasy.org/pROC/) (Basically standard R plot with CIs)
  - Others for specific estimation methods
  - Roll your own

#### Others

  - Excel!
  - SPSS
  - SAS?

Apparently these don't have sensible defaults

## My solution

```{r test-a, fig.keep='none', results = 'asis', echo = 1, fig.width=6, fig.height=6}
library(plotROC)
D.ex <- rbinom(100, 1, .5)
rocdata <- calculate_roc(rnorm(100, mean = D.ex), D.ex)
myrocplot <- ggroc(rocdata) + theme_igray()
cat(svgize_roc(myrocplot, cutoffs = rocdata$c, font.size = "12px", prefix = "a"))
```

## What it does

```{r example, eval = FALSE}
rocres <- calculate_roc(marker, outcome)
rocplot <- ggroc(rocres)

standalone_svg_roc(rocres) ## runs interactive plot in Rstudio or browser
svgize_roc(rocplot) ## generates code to include interactive plot in webpage or knitr
printize_roc(rocplot) ## generates roc plot suitable for print
```

## Print version

```{r, echo = FALSE, fig.width = 5, fig.height = 5}
printize_roc(myrocplot, rocdata)
```

# How it works

## `ggplot` + `gridSVG`

- `ggplot2` to construct an ROC curve chart with sensible defaults
    + Points at each cutoff are plotted invisibly
- `gridSVG` converts ggplot object to svg, and garnishes with data:

![ggtosvg](figures/gridtosvg.png)

## `d3.js`

- Compute [Voronoi tesselation](http://en.wikipedia.org/wiki/Voronoi_diagram) over points that define ROC
- Bind hover events to polygons: display cutpoint and label

![vornoi](figures/voronoi.png)

## Is anyone going to use it?

Maybe: [if we make it easy](http://brbweb1.nci.nih.gov:3838/plotROC/)

## How to contribute

http://github.com/sachsmc/plotROC

- Documentation, how tos
- Testing, bugs
- Features
- Criticism 
- Help write it up (JSS, R Journal, somewhere else?)

## Acknowledgements

**Giants**

- Hadley Wickham [`ggplot2`](http://ggplot2.org/)
- Mike Bostock [`d3.js`](http://d3js.org)
- Simon Potter [`gridSVG`](http://sjp.co.nz/projects/gridsvg/)

**For help with the review**

- Erin Gabriel

**For help with the server**

- Yingdong Zhao

### 2014-10-13 brb-wip-oct.Rmd
edits

@@
 > ROC curve describes possible relationships among the probabilities of the various types of correct and incorrect decisions. Therefore, it plays a central role in optimizing diagnostic strategies
+- Lusted, 1971
 
@@
 By appropriate choice of the decision threshold, a decision maker or observer can operate at (or near) any desired compromise that lies on the curve.
+- Metz, 1978
 
@@
 
-  - [`ROCR`](http://rocr.bioinf.mpi-sb.mpg.de/) (Encodes cutoff values as color scale along curves)
+  - [`ROCR`](http://rocr.bioinf.mpi-sb.mpg.de/) (Encodes cutoff values as color scale along curve)
   - [`pROC`](http://web.expasy.org/pROC/) (Basically standard R plot with CIs)

### 2015-01-27 Paper/sachs2014plotROC.Rmd
drafting paper for jss

---
title: '\pkg{plotROC}: An \proglang{R} Package to Improve the State of ROC Curve Plots in the Medical Literature'
shorttitle: '\pkg{plotROC}: An \proglang{R} Package for ROC Curve Plots'
plaintitle: 'plotROC: An R Package to Improve the State of ROC Curve Plots in the Medical Literature'
author:
- name: Michael C. Sachs
  affil: Biometric Research Branch, Division of Cancer Treatment and Diagnosis, National Cancer Institute
address:
- Michael C. Sachs
- 9609 Medical Center Drive, MSC 9735
- Bethesda, MD 20892
- "telephone: 240-276-7888"
email: michael.sachs@nih.gov
abstract: 'Plots of the receiver operating characteristic (ROC) curve are ubiquitous in medical research. Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers. We evaluate current trends in the design of ROC curve plots with a literature review. Our review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display. We describe our new \proglang{R} package that was created to address the shortcomings of existing tools. The package has functions to create informative ROC curve plots, with sensible defaults, for use in print or as an interactive web-based plot. A web application was developed to reach a broader audience of scientists who do not use R.'
keywords: ROC curves; graphics; interactive; plots
bibliography: plotroc
...

```{r setup, echo = FALSE}
library(knitr)
knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})
opts_chunk$set(dev="tikz", comment = NA)
```

# Introduction

## About ROC Curves

The Receiver Operating Characteristic (ROC) curve is used to assess the accuracy of a continuous measurement for predicting a binary outcome. In medicine, ROC curves have a long history of use for evaluating diagnostic tests in radiology and general diagnostics. ROC curves have also been used for a long time in signal detection theory.

The accuracy of a diagnostic test can be evaluated by considering the two possible types of errors: false positives, and false negatives. For a continuous measurement that we denote as $M$, convention dictates that a test positive is defined as $M$ exceeding some fixed threshold $c$: $M > c$. In reference to the binary outcome that we denote as $D$, a good outcome of the test is when the test is positive among an individual who truly has a disease: $D = 1$. A bad outcome is when the test is positive among an individual who does not have the disease $D = 0$.

Formally, for a fixed cutoff $c$, the true positive fractions is the probability of a test positive among the diseased population:

$$ TPF(c) = P\{ M > c | D = 1 \} $$

and the false positive fraction is the probability of a test positive among the healthy population:

$$ FPF(c) = P\{ M > c | D = 0 \} $$

Since the cutoff $c$ is not usually fixed in advance, we can plot the TPF against the FPF for all possible values of $c$. This is exactly what the ROC curve is, $FPF(c)$ on the $x$ axis and $TPF(c)$ along the $y$ axis. A useless test that is not informative at all in regards to the disease status has $TPF(c) = FPF(c)$ for all $c$. The ROC plot of a useless test is thus the diagonal line. A perfect test that is completely informative about disease status has $TPF(c) = 1$ and $FPF(c) = 0$ for all $c$.

Given a sample of test and disease status pairs, $(M_1, D_1), \ldots, (M_n, D_n)$, we can estimate the ROC curve by computing proportions in the diseased and healthy subgroups separately. Specifically, given a fixed cutoff $c$, an estimate of the $TPF(c)$ is

$$ \widehat{TPF(c)} = \frac{\sum_{i = 1}^n 1\{M_i > c\} \cdot 1\{D_i = 1\}}{\sum_{i=1}^n 1\{D_i = 1\}}, $$

where $1\{\cdot\}$ is the indicator function. An estimate for $FPF(c)$ is given by a similar expression with $D_i = 1$ replaced with $D_i = 0$. Calculating these proportions for $c$ equal to each unique value of the observed $M_i$ yields what is known as the empirical ROC curve estimate. The empirical estimate is a step function. Other methods exist to estimate the ROC curve, such as the binormal parametric estimate which can be used to get a smooth curve. There are also extensions that allow for estimation with time-to-event outcomes subject to censoring. For a more thorough reference on the methods and theory surrounding ROC curves, we refer interested readers to @pepe2003statistical.

A common way to summarize the value of a test for classifying disease status is to calculate the area under the ROC curve (AUC). The greater the AUC, the more informative the test. The AUC summarizes the complexities of the ROC curve into a single number and therefore is widely used to facilitate comparisons between tests and populations. It has been criticized for the same reason because it does not fully characterize the trade-offs between false- and true-positives.

## Design Considerations

The main purpose of visually displaying the ROC curve is to show the trade-off between the FPF and TPF as the cutoff $c$ varies. This can be useful for aiding viewers in choosing an optimal cutoff for decision making, for comparing a small number of candidate tests, and for generally illustrating the performance of the test as a classifier. In practice, once the FPF and TPF are computed for each unique observed cutoff value, they can be plotted as a simple line chart or scatter plot using standard statistical plotting tools. This often leads to the unfortunate design choice of obscuring the critical and useful third dimension, the range of cutoff values $c$.

Another key design element is the use of a diagonal guideline for comparison. The diagonal guideline serves as a baseline for comparison, allowing observers to roughly estimate the area between the diagonal and the estimated ROC curve, which serves as a proxy for estimating the value of the test for classification above a coin-flip. Likewise, gridlines inside the plotting region and carefully selected axes allow for accurate assessment of the TPF and FPF at particular points along the ROC curve. Many medical studies use ROC curves to compare a multitude of candidate tests to each other and to the null diagonal. In those cases, curves need to be distinguished by using a legend combined with different colors or line types, or direct labels inside the plotting region.

In the medical literature, FPF and TPF are usually referred to in terms of the jargon terms Sensitivity and Specificity. Sensitivity is equivalent to the true positive fraction. Specificity is 1 - FPF, the true negative fraction. Sometimes, the FPF and TPF are incorrectly referred to as rates, using the abbreviations FPR and TPR. These are probabilities and their estimates are proportions, therefore we prefer the use of the term fraction as opposed to rate.

## Existing Plotting Software

The ROC curve plot is, at the most basic level, a line graph. Therefore, once the appropriate statistics are estimated, existing plotting functions can be used to create an ROC curve plot. The addition of axis labels, titles, legends, and so on, indicate a chart as such. In our literature review, we observed plots with the distinctive characteristics of the base plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions.

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF versus TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is place along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions.

## Motivation

Anyone giving a cursory look at any of the major medical journals is likely to find at least one ROC curve plot. We sought assess the usage of ROC curve plots and to evaluate the design choices made in the current oncology literature by conducting a small literature review. We searched Pubmed for  clinical trials or observational studies in humans reported in major oncology journals for the past 10 years for the terms "ROC Curve" OR "ROC Analysis" OR "Receiver operating characteristic curve". The search was conducted on October 8, 2014 and returned 54 papers. From those papers, 47 images were extracted and reviewed. The exact specifications for the Pubmed query are available in the supplementary materials.

Each image consisted of a single ROC curve plot or a panel of multiple plots. Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of diagonal guide line, whether any cutpoint were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the area under the curve. The numerical results of the survey are summarized in table \ref{table1}.

```{r review, echo = FALSE, results = 'asis'}
library(xtable)

rev <- read.csv("../../Literature Review/rocusage.csv", stringsAsFactors = FALSE)
rev <- subset(rev, flag != "comedy")
rev$axis.names <- gsub("fpr/tpr", "FPF/TPF", rev$axis.names, fixed = TRUE)

mytable <- function(var, groups = sort(unique(var)), tab = FALSE, name = NULL) {

    if(tab) groupsa <- paste0("\t", groups) else groupsa <- groups
    blah <- data.frame(lab = groupsa, `percent (count)`  = NA, check.names = FALSE, stringsAsFactors = FALSE)
    N <- sum(!is.na(var))
    for(k in 1:length(groups)){

      nk <- sum(var == groups[k])
      perc <- nk/N
      cnt <- nk

      blah[k, 2] <- paste0(format(100 * perc, digits = 1, nsmall = 1), " (", cnt, ")")

    }

    if(all(groups == c("no", "yes"))) {

      c(name, paste0(format(100 * perc, digits = 1, nsmall = 1), " (", cnt, ")"))

      } else blah

  }

rev$numcurves <- with(rev, ifelse(number.of.curves >= 4, "4+", as.character(number.of.curves)))
rev$cutyesno <- with(rev, ifelse(cutoffs != "none" & cutoffs != "points", "yes", "no"))

a1 <- rbind(c("Number of curves", ""), mytable(rev$numcurves, tab = TRUE))
a2 <- c("\tAverage (SD)", paste0(round(mean(rev$number.of.curves), 1), " (", round(sd(rev$number.of.curves), 1), ")"))

b1 <- rbind(c("Axis labels", ""), mytable(rev$axis.names, tab = TRUE))
clist <- with(rev, list(`Diagonal Guide` = guide, Gridlines = minor.axes, `Cutoffs indicated` = cutyesno, `AUC indicated` = auc))
cname <- names(clist)
c1 <- as.data.frame(do.call(rbind, lapply(1:length(clist), 
                            function(x) mytable(clist[[x]], groups = c("no", "yes"), tab = FALSE, name = cname[x]))))
colnames(c1) <- colnames(a1)
d1 <- rbind(c("Curve Labels", ""), mytable(rev$labels, tab = TRUE))

table1 <- rbind(a1, a2, b1, c1, d1)
colnames(table1) <- c("", colnames(table1)[2])
print(xtable(table1, label = "table1", 
             caption = "Results of a literature review of major oncology journals for ROC curve plots. The rows indicate the frequency and count of key design elements of an ROC curve plot. FPR = False positive rate; TPR = True positive rate; sens = Sensitivity; spec = Specificity; AUC = Area under the Curve"), 
      include.rownames = FALSE, 
      comment = FALSE, 
      sanitize.text.function = function(x) gsub("\t", "$\\quad$", x, fixed = TRUE))

```

The small minority of the figures make any attempt to indicate the values of the test cutoffs, which is an integral component of the ROC curve. We conjecture that this is mainly due to the use of default plotting procedures in statistical software. The software, by default, treats the ROC curve as a 2 dimensional object, obscuring the cutoff dimension. Gridlines and direct labels are also somewhat out of the ordinary. The absence of these features make accurate determination and comparison of the values more difficult. Many of the plots included large tables containing estimates and inference for AUCs, while the ROC curves themselves, numerous and without clear labels or reference lines, merely served as decoration. We aim to solve some of these problems by providing an easy-to-use plotting interface for the ROC curve that provides sensible defaults.

The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements. We favor the use of gridline and reference lines to facilitate accurate readings off of the axes. Direct labels are preferred over legends because they omit the additional cognitive step of matching line types to labels. Our \pkg{plotROC} package additionally provides plotting of cutoff values, using hover events in interactive use, and direct labels for print use. Exact confidence regions for points on the ROC curve are optionally calculated and displayed. Additionally, we use axis scales are adjusted to be denser near the margins 0 and 1. In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins. The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.

```{r figure1, fig.width = 6, fig.height = 6, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves. Panel A shows a sparse ROC curve, with no design additions inside the plotting region. The plot results in more white space than anything else. It is difficult to accurately determine values without reference lines. Panel B shows a plot comparing 2 curves, with different line types and a legend. AUCs are also given in the legend. Panels B and C add gridlines, diagonal reference lines, and direct labels. \\label{figure1}"}
library(plotROC)

set.seed(520)
D0 <- rbinom(100, size = 1, prob = .5)
M1 <- rnorm(100, mean = D0)
M2 <- rnorm(100, mean = D0, sd = 1.5)

R1 <- calculate_roc(M1, D0)
R2 <- calculate_roc(M2, D0)

par(mfrow = c(2,2))

# A
plot(TPF ~ FPF, data = R1, type = 's', xlab = "1 - Specificity", ylab = "Sensitivity", main = "A")
# B
plot(TPF ~ FPF, data = R1, type = 's', xlab = "1 - Specificity", ylab = "Sensitivity", main = "B")
lines(TPF ~ FPF, data = R2, type = 's', lty = 2)
legend("bottomright", legend = c("Test A: AUC = 0.81", "Test B: AUC = 0.64"), lty = 1:2)

# C
plot(TPF ~ FPF, data = R1, type = 's', lwd  = 2, xlab = "False positive fraction", ylab = "True positive fraction", main = "C")
abline(v = seq(0, 1, by = .2), lwd = 1, col = "grey80")
abline(h = seq(0, 1, by = .2), lwd = 1, col = "grey80")
abline(0, 1, col = "grey80")
lines(TPF ~ FPF, data = R1, type = 's', lwd = 2)
text(.2, .8, "Test A")

# D
plot(TPF ~ FPF, data = R1, type = 's', lwd  = 2, xlab = "False positive fraction", ylab = "True positive fraction", main = "D")
abline(v = seq(0, 1, by = .2), lwd = 1, col = "grey80")
abline(h = seq(0, 1, by = .2), lwd = 1, col = "grey80")
abline(0, 1, col = "grey80")
lines(TPF ~ FPF, data = R1, type = 's', lwd = 2)
lines(TPF ~ FPF, data = R2, type = 's', lwd = 2)
text(.2, .8, "A")
text(.3, .4, "B")
```

# Usage of the Package

## Shiny application

We created a \pkg{shiny} application [@shiny] in order to make the features more accessible to non-\proglang{R} users. A limited subset of the functions of the \pkg{plotROC} can be performed on an example dataset or on data that users upload to the website. Resulting plots can be saved to the users' machine as a pdf or as a stand-alone html file.  It can be used in any modern web browser with no other dependencies at the website here: http://sachsmc.shinyapps.io/plotROC.

## Quick start
After installing, the interactive Shiny application can be run locally. 
```{r shiny, eval = FALSE}
shiny_plotROC()
```

## Command line basic usage

We start by creating an example data set. The marker we generate is moderately accurate for predicting disease status.

```{r dataset}
library(plotROC)
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
```

Next we use the `calculate_roc` function to compute the empirical ROC curve. The disease status need not be coded as 0/1, but if it is not, `plotROC` assumes (with a warning) that the lowest value in sort order signifies disease-free status. This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.

```{r calc}
rocdata <- calculate_roc(M.ex, D.ex)
str(rocdata)
```

The same data.frame `rocdata` can be used to generate an interactive plot to view in Rstudio viewer or web browser.

```{r inter, eval = FALSE}
plot_interactive_roc(rocdata)
```

The `rocdata` is passed to the `ggroc` function with an optional label. This creates a ggplot object of the ROC curve using the \pkg{ggplot2} package [@ggplot2].

```{r test-a}
myrocplot <- ggroc(rocdata, label = "Example")
```

We can create an interactive ROC plot using the `export_interactive_roc` function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}. The key interactive feature of these plots is that the cutoff values nearest to the mouse cursor are displayed when hovering over the figure. Clicking makes the cutoff label stick until the next click. The character string can be copy-pasted into an html document, or better yet, using \pkg{knitr} [@knitr], we can `cat` the results and use the option `results = 'asis'` so that the interactive plot is displayed correctly. For examples of interactive plots and how to incorporate them into `knitr` documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the webpage https://sachsmc.github.io/plotROC/. Of note is the `prefix` option, which allows the user to assign each plot a unique identifier. This is necessary to prevent conflicts with multiple plots in a single webpage.

```{r int-no, eval = FALSE}
cat(
  export_interactive_roc(myrocplot, cutoffs = rocdata$c, 
                         font.size = "12px", prefix = "a")
  )
```

The same `ggroc` object that we called `myrocplot` can be used to generate an ROC plot suitable for use in print. It annotates the cutoff values and is completely in black and white. A simple example with the default options is shown in figure \ref{first}.

```{r print, fig.width = 6, fig.height = 6, fig.cap = "Illustration of ROC curve plot generated by \\pkg{plotROC} for use in print. \\label{first}"}
plot_journal_roc(myrocplot, rocdata)
```

## Advanced options

### Click to view confidence region

We use the `ci = TRUE` option in `calculcate_roc` and `ggroc` to compute confidence regions for points on the ROC curve using the @clopper1934use exact method. The significance level can be specified using the `alpha` option.

```{r test-a-ci}
rocdata <- calculate_roc(M.ex, D.ex, ci = TRUE, alpha = 0.05)
myrocplot <- ggroc(rocdata, label = "Example", ci = TRUE)
```

For interactive plots, the confidence regions are automatically detected. When the user clicks on the ROC curve, the confidence region for the TPF and FPF is overlaid using a grey rectangle. The label and region stick until the next click.

```{r int-nob, eval = FALSE}
cat(
  export_interactive_roc(myrocplot, cutoffs = rocdata$c, 
                         font.size = "12px", prefix = "aci")
  )
```

For use in print, we pass a small vector of cutoff locations at which to display the confidence regions. This is shown in figure \ref{conf}.

```{r printci, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} plot with exact confidence regions. \\label{conf}"}
plot_journal_roc(myrocplot, rocdata, n.cuts = 10, 
                 ci.at = c(-.5, .5, 2.1))
```

### Multiple ROC curves

If you have multiple tests of different types on the same subjects, you can use the `calculate_multi_roc` function to compute the empirical ROC curve for each test.  Then the `multi_ggroc` function creates the appropriate type of `ggplot` object. Confidence regions are not supported for multiple curves at the time of writing.

```{r multistart}
D.ex <- rbinom(100, 1, .5)

fakedata <- data.frame(M1 = rnorm(100, mean = D.ex), 
                       M2 = rnorm(100, mean = D.ex, sd = .4), 
                       M3 = runif(100), D = D.ex)

datalist <- calculate_multi_roc(fakedata, c("M1", "M2", "M3"), "D")
rocplot <- multi_ggroc(datalist)
```

This multi ggroc object can be passed to the `plot_journal_roc` and the `export_interactive_roc` functions, as desired.

Labels can be added easily with the `label` option. The length of the label element should match the number of plotted curves. The resulting plot is shown in figure \ref{multi}.

```{r multi2, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} plot with multiple curves. \\label{multi} "}
rocplot <- multi_ggroc(datalist, label = c("M1", "M2", "M3"))
plot_journal_roc(rocplot, datalist)
```

### Themes and annotations

`plotROC` uses the `ggplot2` package to create ggplot objects. Therefore, themes and annotations can be added to `ggroc` objects in the usual `ggplot` way. A `plotROC` figure with a new theme, title, axis label, and AUC annotation is shown in figure \ref{annotate}.

```{r print2, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
library(ggplot2)
plot_journal_roc(myrocplot, rocdata) + 
  theme_grey() + 
  geom_abline(intercept = 0, slope = 1, color = "white") + 
  ggtitle("Themes and annotations") + 
  annotate("text", x = .75, y = .25, 
           label = "AUC = 0.80") + 
  ylab("Sensitivity")
```

### Other estimation methods

By default `calculate_roc` computes the empirical ROC curve. There are other estimation methods out there, as we have summarized in the introduction.  Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned. Then you can simply pass those values in a data frame to the `ggroc` function. For example, let us use the binormal method to create a smooth curve. This approach assumes that the test distribution is normal conditional on disease status.

```{r binormalsetup}
D.ex <- rbinom(100, 1, .5)
M.ex <- rnorm(100, mean = D.ex, sd = .5)

mu1 <- mean(M.ex[D.ex == 1])
mu0 <- mean(M.ex[D.ex == 0])
s1 <- sd(M.ex[D.ex == 1])
s0 <- sd(M.ex[D.ex == 0])
c.ex <- seq(min(M.ex), max(M.ex), length.out = 300)

binorm_rocdata <- data.frame(c = c.ex, 
                             FPF = pnorm((mu0 - c.ex)/s0), 
                             TPF = pnorm((mu1 - c.ex)/s1)
                             )
```

Then we can pass this data.frame to the `ggroc` function as before. The example is shown in figure \ref{binorm}.

```{r binormal, echo = TRUE, fig.width=6, fig.height=6, fig.cap = "Illustration of smooth binormal ROC curve. \\label{binorm}"}
binorm_plot <- ggroc(binorm_rocdata, label = "Binormal")
plot_journal_roc(binorm_plot, binorm_rocdata)
```

# How it Works

# Discussion

Bibliography 
===============

### 2015-01-27 Paper/sachs2014plotROC.tex
drafting paper for jss

--- 2015-01-27 Paper/sachs2014plotROC.Rmd
@@
----
-title: '\pkg{plotROC}: An \proglang{R} Package to Improve the State of ROC Curve Plots in the Medical Literature'
-shorttitle: '\pkg{plotROC}: An \proglang{R} Package for ROC Curve Plots'
-plaintitle: 'plotROC: An R Package to Improve the State of ROC Curve Plots in the Medical Literature'
-author:
-- name: Michael C. Sachs
-  affil: Biometric Research Branch, Division of Cancer Treatment and Diagnosis, National Cancer Institute
-address:
-- Michael C. Sachs
-- 9609 Medical Center Drive, MSC 9735
-- Bethesda, MD 20892
-- "telephone: 240-276-7888"
-email: michael.sachs@nih.gov
-abstract: 'Plots of the receiver operating characteristic (ROC) curve are ubiquitous in medical research.
-Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers.
-We evaluate current trends in the design of ROC curve plots with a literature review.
-Our review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display.
-We describe our new \proglang{R} package that was created to address the shortcomings of existing tools.
-The package has functions to create informative ROC curve plots, with sensible defaults, for use in print or as an interactive web-based plot.
-A web application was developed to reach a broader audience of scientists who do not use R.'
-keywords: ROC curves; graphics; interactive; plots
-bibliography: plotroc
-...
+\documentclass[article]{jss}
 
-```{r setup, echo = FALSE}
-library(knitr)
-knit_hooks$set(plot = function(x, options) {
-  if ('tikz' %in% options$dev && !options$external) {
-    hook_plot_tex(x, options)
-  } else hook_plot_md(x, options)
-})
-opts_chunk$set(dev="tikz", comment = NA)
-```
+\usepackage[utf8]{inputenc}
+\usepackage{tikz}
+\usepackage{natbib}
 
-# Introduction
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-## About ROC Curves
+%% almost as usual
+\author{		Michael C. Sachs\\Biometric Research Branch, Division of Cancer Treatment and Diagnosis,
+National Cancer Institute		}
 
-The Receiver Operating Characteristic (ROC) curve is used to assess the accuracy of a continuous measurement for predicting a binary outcome.
-In medicine, ROC curves have a long history of use for evaluating diagnostic tests in radiology and general diagnostics.
-ROC curves have also been used for a long time in signal detection theory.
+\title{\pkg{plotROC}: An \proglang{R} Package to Improve the State of ROC Curve
+Plots in the Medical Literature}
 
-The accuracy of a diagnostic test can be evaluated by considering the two possible types of errors: false positives, and false negatives.
-For a continuous measurement that we denote as $M$, convention dictates that a test positive is defined as $M$ exceeding some fixed threshold $c$: $M > c$.
-In reference to the binary outcome that we denote as $D$, a good outcome of the test is when the test is positive among an individual who truly has a disease: $D = 1$.
-A bad outcome is when the test is positive among an individual who does not have the disease $D = 0$.
+%% for pretty printing and a nice hypersummary also set:
+\Plainauthor{ Michael C. Sachs} %% comma-separated
+\Plaintitle{plotROC: An R Package to Improve the State of ROC Curve Plots in the
+Medical Literature} %% without formatting
 
-Formally, for a fixed cutoff $c$, the true positive fractions is the probability of a test positive among the diseased population:
+\Shorttitle{\pkg{plotROC}: An \proglang{R} Package for ROC Curve Plots} %% a short title (if necessary)
+%% an abstract and keywords
+\Abstract{
+  Plots of the receiver operating characteristic (ROC) curve are
+  ubiquitous in medical research. Designed to simultaneously display the
+  operating characteristics at every possible value of a continuous
+  diagnostic test, ROC curves are used in oncology to evaluate screening,
+  diagnostic, prognostic and predictive biomarkers. We evaluate current
+  trends in the design of ROC curve plots with a literature review. Our
+  review suggests that ROC curve plots are often ineffective as
+  statistical charts and that poor design obscures the relevant
+  information the chart is intended to display. We describe our new
+  \proglang{R} package that was created to address the shortcomings of
+  existing tools. The package has functions to create informative ROC
+  curve plots, with sensible defaults, for use in print or as an
+  interactive web-based plot. A web application was developed to reach a
+  broader audience of scientists who do not use R.
+}
+\Keywords{ROC curves; graphics; interactive; plots}
+\Plainkeywords{ROC curves; graphics; interactive; plots} %% without formatting
+%% at least one keyword must be supplied
 
-$$ TPF(c) = P\{ M > c | D = 1 \} $$
+%% publication information
+%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
+%% \Volume{50}
+%% \Issue{9}
+%% \Month{June}
+%% \Year{2012}
+%% \Submitdate{2012-06-04}
+%% \Acceptdate{2012-06-04}
 
-and the false positive fraction is the probability of a test positive among the healthy population:
+%% The address of (at least) one author should be given
+%% in the following format:
+\Address{
+    Michael C. Sachs\\
+    9609 Medical Center Drive, MSC 9735\\
+    Bethesda, MD 20892\\
+    telephone: 240-276-7888  \\\email{michael.sachs@nih.gov}
+}
+%% It is also possible to add a telephone and fax number
+%% before the e-mail in the following format:
+%% Telephone: +43/512/507-7103
+%% Fax: +43/512/507-2851
 
-$$ FPF(c) = P\{ M > c | D = 0 \} $$
+%% for those who use Sweave please include the following line (with % symbols):
+%% need no \usepackage{Sweave.sty}
 
-Since the cutoff $c$ is not usually fixed in advance, we can plot the TPF against the FPF for all possible values of $c$.
-This is exactly what the ROC curve is, $FPF(c)$ on the $x$ axis and $TPF(c)$ along the $y$ axis.
-A useless test that is not informative at all in regards to the disease status has $TPF(c) = FPF(c)$ for all $c$.
-The ROC plot of a useless test is thus the diagonal line.
-A perfect test that is completely informative about disease status has $TPF(c) = 1$ and $FPF(c) = 0$ for all $c$.
+%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Given a sample of test and disease status pairs, $(M_1, D_1), \ldots, (M_n, D_n)$, we can estimate the ROC curve by computing proportions in the diseased and healthy subgroups separately.
-Specifically, given a fixed cutoff $c$, an estimate of the $TPF(c)$ is
+\begin{document}
 
-$$ \widehat{TPF(c)} = \frac{\sum_{i = 1}^n 1\{M_i > c\} \cdot 1\{D_i = 1\}}{\sum_{i=1}^n 1\{D_i = 1\}}, $$
+\section{Introduction}\label{introduction}
 
-where $1\{\cdot\}$ is the indicator function.
-An estimate for $FPF(c)$ is given by a similar expression with $D_i = 1$ replaced with $D_i = 0$.
-Calculating these proportions for $c$ equal to each unique value of the observed $M_i$ yields what is known as the empirical ROC curve estimate.
-The empirical estimate is a step function.
-Other methods exist to estimate the ROC curve, such as the binormal parametric estimate which can be used to get a smooth curve.
-There are also extensions that allow for estimation with time-to-event outcomes subject to censoring.
-For a more thorough reference on the methods and theory surrounding ROC curves, we refer interested readers to @pepe2003statistical.
+\subsection{About ROC Curves}\label{about-roc-curves}
 
-A common way to summarize the value of a test for classifying disease status is to calculate the area under the ROC curve (AUC).
-The greater the AUC, the more informative the test.
-The AUC summarizes the complexities of the ROC curve into a single number and therefore is widely used to facilitate comparisons between tests and populations.
-It has been criticized for the same reason because it does not fully characterize the trade-offs between false- and true-positives.
+The Receiver Operating Characteristic (ROC) curve is used to assess the
+accuracy of a continuous measurement for predicting a binary outcome. In
+medicine, ROC curves have a long history of use for evaluating
+diagnostic tests in radiology and general diagnostics. ROC curves have
+also been used for a long time in signal detection theory.
 
-## Design Considerations
+The accuracy of a diagnostic test can be evaluated by considering the
+two possible types of errors: false positives, and false negatives. For
+a continuous measurement that we denote as \(M\), convention dictates
+that a test positive is defined as \(M\) exceeding some fixed threshold
+\(c\): \(M > c\). In reference to the binary outcome that we denote as
+\(D\), a good outcome of the test is when the test is positive among an
+individual who truly has a disease: \(D = 1\). A bad outcome is when the
+test is positive among an individual who does not have the disease
+\(D = 0\).
 
-The main purpose of visually displaying the ROC curve is to show the trade-off between the FPF and TPF as the cutoff $c$ varies.
-This can be useful for aiding viewers in choosing an optimal cutoff for decision making, for comparing a small number of candidate tests, and for generally illustrating the performance of the test as a classifier.
-In practice, once the FPF and TPF are computed for each unique observed cutoff value, they can be plotted as a simple line chart or scatter plot using standard statistical plotting tools.
-This often leads to the unfortunate design choice of obscuring the critical and useful third dimension, the range of cutoff values $c$.
+Formally, for a fixed cutoff \(c\), the true positive fractions is the
+probability of a test positive among the diseased population:
 
-Another key design element is the use of a diagonal guideline for comparison.
-The diagonal guideline serves as a baseline for comparison, allowing observers to roughly estimate the area between the diagonal and the estimated ROC curve, which serves as a proxy for estimating the value of the test for classification above a coin-flip.
-Likewise, gridlines inside the plotting region and carefully selected axes allow for accurate assessment of the TPF and FPF at particular points along the ROC curve.
-Many medical studies use ROC curves to compare a multitude of candidate tests to each other and to the null diagonal.
-In those cases, curves need to be distinguished by using a legend combined with different colors or line types, or direct labels inside the plotting region.
+\[ TPF(c) = P\{ M > c | D = 1 \} \]
 
-In the medical literature, FPF and TPF are usually referred to in terms of the jargon terms Sensitivity and Specificity.
-Sensitivity is equivalent to the true positive fraction.
-Specificity is 1 - FPF, the true negative fraction.
-Sometimes, the FPF and TPF are incorrectly referred to as rates, using the abbreviations FPR and TPR.
-These are probabilities and their estimates are proportions, therefore we prefer the use of the term fraction as opposed to rate.
+and the false positive fraction is the probability of a test positive
+among the healthy population:
 
-## Existing Plotting Software
+\[ FPF(c) = P\{ M > c | D = 0 \} \]
 
-The ROC curve plot is, at the most basic level, a line graph.
-Therefore, once the appropriate statistics are estimated, existing plotting functions can be used to create an ROC curve plot.
-The addition of axis labels, titles, legends, and so on, indicate a chart as such.
-In our literature review, we observed plots with the distinctive characteristics of the base plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions.
+Since the cutoff \(c\) is not usually fixed in advance, we can plot the
+TPF against the FPF for all possible values of \(c\). This is exactly
+what the ROC curve is, \(FPF(c)\) on the \(x\) axis and \(TPF(c)\) along
+the \(y\) axis. A useless test that is not informative at all in regards
+to the disease status has \(TPF(c) = FPF(c)\) for all \(c\). The ROC
+plot of a useless test is thus the diagonal line. A perfect test that is
+completely informative about disease status has \(TPF(c) = 1\) and
+\(FPF(c) = 0\) for all \(c\).
 
-There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions.
-The \pkg{ROCR} package [@rocr] plots the FPF versus TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself.
-A legend for the color scale is place along the vertical axis on the right of the plotting region.
-The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve.
-The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions.
+Given a sample of test and disease status pairs,
+\((M_1, D_1), \ldots, (M_n, D_n)\), we can estimate the ROC curve by
+computing proportions in the diseased and healthy subgroups separately.
+Specifically, given a fixed cutoff \(c\), an estimate of the \(TPF(c)\)
+is
 
-## Motivation
+\[ \widehat{TPF(c)} = \frac{\sum_{i = 1}^n 1\{M_i > c\} \cdot 1\{D_i = 1\}}{\sum_{i=1}^n 1\{D_i = 1\}}, \]
 
-Anyone giving a cursory look at any of the major medical journals is likely to find at least one ROC curve plot.
-We sought assess the usage of ROC curve plots and to evaluate the design choices made in the current oncology literature by conducting a small literature review.
-We searched Pubmed for  clinical trials or observational studies in humans reported in major oncology journals for the past 10 years for the terms "ROC Curve" OR "ROC Analysis" OR "Receiver operating characteristic curve".
-The search was conducted on October 8, 2014 and returned 54 papers.
-From those papers, 47 images were extracted and reviewed.
-The exact specifications for the Pubmed query are available in the supplementary materials.
+where \(1\{\cdot\}\) is the indicator function. An estimate for
+\(FPF(c)\) is given by a similar expression with \(D_i = 1\) replaced
+with \(D_i = 0\). Calculating these proportions for \(c\) equal to each
+unique value of the observed \(M_i\) yields what is known as the
+empirical ROC curve estimate. The empirical estimate is a step function.
+Other methods exist to estimate the ROC curve, such as the binormal
+parametric estimate which can be used to get a smooth curve. There are
+also extensions that allow for estimation with time-to-event outcomes
+subject to censoring. For a more thorough reference on the methods and
+theory surrounding ROC curves, we refer interested readers to
+\citet{pepe2003statistical}.
 
-Each image consisted of a single ROC curve plot or a panel of multiple plots.
-Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of diagonal guide line, whether any cutpoint were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the area under the curve.
-The numerical results of the survey are summarized in table \ref{table1}.
+A common way to summarize the value of a test for classifying disease
+status is to calculate the area under the ROC curve (AUC). The greater
+the AUC, the more informative the test. The AUC summarizes the
+complexities of the ROC curve into a single number and therefore is
+widely used to facilitate comparisons between tests and populations. It
+has been criticized for the same reason because it does not fully
+characterize the trade-offs between false- and true-positives.
 
-```{r review, echo = FALSE, results = 'asis'}
-library(xtable)
+\subsection{Design Considerations}\label{design-considerations}
 
-rev <- read.csv("../../Literature Review/rocusage.csv", stringsAsFactors = FALSE)
-rev <- subset(rev, flag != "comedy")
-rev$axis.names <- gsub("fpr/tpr", "FPF/TPF", rev$axis.names, fixed = TRUE)
+The main purpose of visually displaying the ROC curve is to show the
+trade-off between the FPF and TPF as the cutoff \(c\) varies. This can
+be useful for aiding viewers in choosing an optimal cutoff for decision
+making, for comparing a small number of candidate tests, and for
+generally illustrating the performance of the test as a classifier. In
+practice, once the FPF and TPF are computed for each unique observed
+cutoff value, they can be plotted as a simple line chart or scatter plot
+using standard statistical plotting tools. This often leads to the
+unfortunate design choice of obscuring the critical and useful third
+dimension, the range of cutoff values \(c\).
 
-mytable <- function(var, groups = sort(unique(var)), tab = FALSE, name = NULL) {
+Another key design element is the use of a diagonal guideline for
+comparison. The diagonal guideline serves as a baseline for comparison,
+allowing observers to roughly estimate the area between the diagonal and
+the estimated ROC curve, which serves as a proxy for estimating the
+value of the test for classification above a coin-flip. Likewise,
+gridlines inside the plotting region and carefully selected axes allow
+for accurate assessment of the TPF and FPF at particular points along
+the ROC curve. Many medical studies use ROC curves to compare a
+multitude of candidate tests to each other and to the null diagonal. In
+those cases, curves need to be distinguished by using a legend combined
+with different colors or line types, or direct labels inside the
+plotting region.
 
-    if(tab) groupsa <- paste0("\t", groups) else groupsa <- groups
-    blah <- data.frame(lab = groupsa, `percent (count)`  = NA, check.names = FALSE, stringsAsFactors = FALSE)
-    N <- sum(!is.na(var))
-    for(k in 1:length(groups)){
+In the medical literature, FPF and TPF are usually referred to in terms
+of the jargon terms Sensitivity and Specificity. Sensitivity is
+equivalent to the true positive fraction. Specificity is 1 - FPF, the
+true negative fraction. Sometimes, the FPF and TPF are incorrectly
+referred to as rates, using the abbreviations FPR and TPR. These are
+probabilities and their estimates are proportions, therefore we prefer
+the use of the term fraction as opposed to rate.
 
-      nk <- sum(var == groups[k])
-      perc <- nk/N
-      cnt <- nk
+\subsection{Existing Plotting
+Software}\label{existing-plotting-software}
 
-      blah[k, 2] <- paste0(format(100 * perc, digits = 1, nsmall = 1), " (", cnt, ")")
+The ROC curve plot is, at the most basic level, a line graph. Therefore,
+once the appropriate statistics are estimated, existing plotting
+functions can be used to create an ROC curve plot. The addition of axis
+labels, titles, legends, and so on, indicate a chart as such. In our
+literature review, we observed plots with the distinctive
+characteristics of the base plotting functions from Microsoft Office,
+\proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting
+functions.
 
-    }
+There are several \proglang{R} packages related to ROC curve estimation
+that contain dedicated plotting functions. The \pkg{ROCR} package
+\citep{rocr} plots the FPF versus TPF, as usual, and then takes the
+interesting approach of encoding the cutoff values as a separate color
+scale along the ROC curve itself. A legend for the color scale is place
+along the vertical axis on the right of the plotting region. The
+\pkg{pROC} package \citep{pROC} is mainly focused on estimating
+confidence intervals and regions for restricted ranges of the ROC curve.
+The plotting methods therein use the base \proglang{R} plotting
+functions to create nice displays of the curves along with shaded
+confidence regions.
 
-    if(all(groups == c("no", "yes"))) {
+\subsection{Motivation}\label{motivation}
 
-      c(name, paste0(format(100 * perc, digits = 1, nsmall = 1), " (", cnt, ")"))
+Anyone giving a cursory look at any of the major medical journals is
+likely to find at least one ROC curve plot. We sought assess the usage
+of ROC curve plots and to evaluate the design choices made in the
+current oncology literature by conducting a small literature review. We
+searched Pubmed for clinical trials or observational studies in humans
+reported in major oncology journals for the past 10 years for the terms
+``ROC Curve'' OR ``ROC Analysis'' OR ``Receiver operating characteristic
+curve''. The search was conducted on October 8, 2014 and returned 54
+papers. From those papers, 47 images were extracted and reviewed. The
+exact specifications for the Pubmed query are available in the
+supplementary materials.
 
-      } else blah
+Each image consisted of a single ROC curve plot or a panel of multiple
+plots. Each plot was inspected manually for the following design
+features: the number of curves displayed, the type of axis labels
+(sensitivity/ 1 - specificity or true/false positive fractions),
+presence or absence of grid lines, presence or absence of diagonal guide
+line, whether any cutpoint were indicated, the type of curve label
+(legend or direct label), and presence of other textual annotations such
+as the area under the curve. The numerical results of the survey are
+summarized in table \ref{table1}.
 
-  }
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+  \hline
+ & percent (count) \\ 
+  \hline
+Number of curves &  \\ 
+  $\quad$1 & 19.6 (9) \\ 
+  $\quad$2 & 43.5 (20) \\ 
+  $\quad$3 & 10.9 (5) \\ 
+  $\quad$4+ & 26.1 (12) \\ 
+  $\quad$Average (SD) & 2.6 (1.5) \\ 
+  Axis labels &  \\ 
+  $\quad$FPF/TPF & 13.0 (6) \\ 
+  $\quad$mixed & 2.2 (1) \\ 
+  $\quad$none & 2.2 (1) \\ 
+  $\quad$sens/spec & 82.6 (38) \\ 
+  Diagonal Guide & 43.5 (20) \\ 
+  Gridlines & 17.4 (8) \\ 
+  Cutoffs indicated & 15.2 (7) \\ 
+  AUC indicated & 50.0 (23) \\ 
+  Curve Labels &  \\ 
+  $\quad$direct & 10.9 (5) \\ 
+  $\quad$legend & 63.0 (29) \\ 
+  $\quad$none & 19.6 (9) \\ 
+  $\quad$title & 6.5 (3) \\ 
+   \hline
+\end{tabular}
+\caption{Results of a literature review of major oncology journals for ROC curve plots.
+The rows indicate the frequency and count of key design elements of an ROC curve plot.
+FPR = False positive rate; TPR = True positive rate; sens = Sensitivity; spec = Specificity; AUC = Area under the Curve}
+\label{table1}
+\end{table}
 
-rev$numcurves <- with(rev, ifelse(number.of.curves >= 4, "4+", as.character(number.of.curves)))
-rev$cutyesno <- with(rev, ifelse(cutoffs != "none" & cutoffs != "points", "yes", "no"))
+The small minority of the figures make any attempt to indicate the
+values of the test cutoffs, which is an integral component of the ROC
+curve. We conjecture that this is mainly due to the use of default
+plotting procedures in statistical software. The software, by default,
+treats the ROC curve as a 2 dimensional object, obscuring the cutoff
+dimension. Gridlines and direct labels are also somewhat out of the
+ordinary. The absence of these features make accurate determination and
+comparison of the values more difficult. Many of the plots included
+large tables containing estimates and inference for AUCs, while the ROC
+curves themselves, numerous and without clear labels or reference lines,
+merely served as decoration. We aim to solve some of these problems by
+providing an easy-to-use plotting interface for the ROC curve that
+provides sensible defaults.
 
-a1 <- rbind(c("Number of curves", ""), mytable(rev$numcurves, tab = TRUE))
-a2 <- c("\tAverage (SD)", paste0(round(mean(rev$number.of.curves), 1), " (", round(sd(rev$number.of.curves), 1), ")"))
+The panels of figure \ref{figure1} illustrate the most common styles of
+ROC curve plots, and the associated design elements. We favor the use of
+gridline and reference lines to facilitate accurate readings off of the
+axes. Direct labels are preferred over legends because they omit the
+additional cognitive step of matching line types to labels. Our
+\pkg{plotROC} package additionally provides plotting of cutoff values,
+using hover events in interactive use, and direct labels for print use.
+Exact confidence regions for points on the ROC curve are optionally
+calculated and displayed. Additionally, we use axis scales are adjusted
+to be denser near the margins 0 and 1. In medical applications, it is
+often necessary to have a very low FPR (less than 10\%, for instance),
+therefore the smaller scales are useful for accurately determining
+values near the margins. The next section details the usage of the
+\pkg{plotROC} \proglang{R} package and these features.
 
-b1 <- rbind(c("Axis labels", ""), mytable(rev$axis.names, tab = TRUE))
-clist <- with(rev, list(`Diagonal Guide` = guide, Gridlines = minor.axes, `Cutoffs indicated` = cutyesno, `AUC indicated` = auc))
-cname <- names(clist)
-c1 <- as.data.frame(do.call(rbind, lapply(1:length(clist), 
-                            function(x) mytable(clist[[x]], groups = c("no", "yes"), tab = FALSE, name = cname[x]))))
-colnames(c1) <- colnames(a1)
-d1 <- rbind(c("Curve Labels", ""), mytable(rev$labels, tab = TRUE))
+\begin{figure}[htbp]
+\centering
+\includegraphics{figure/figure1-1.pdf}
+\caption{Illustration of design choices in plotting ROC curves. Panel A
+shows a sparse ROC curve, with no design additions inside the plotting
+region. The plot results in more white space than anything else. It is
+difficult to accurately determine values without reference lines. Panel
+B shows a plot comparing 2 curves, with different line types and a
+legend. AUCs are also given in the legend. Panels B and C add gridlines,
+diagonal reference lines, and direct labels. \label{figure1}}
+\end{figure}
 
-table1 <- rbind(a1, a2, b1, c1, d1)
-colnames(table1) <- c("", colnames(table1)[2])
-print(xtable(table1, label = "table1", 
-             caption = "Results of a literature review of major oncology journals for ROC curve plots.
-The rows indicate the frequency and count of key design elements of an ROC curve plot.
-FPR = False positive rate; TPR = True positive rate; sens = Sensitivity; spec = Specificity; AUC = Area under the Curve"),
-      include.rownames = FALSE, 
-      comment = FALSE, 
-      sanitize.text.function = function(x) gsub("\t", "$\\quad$", x, fixed = TRUE))
+\section{Usage of the Package}\label{usage-of-the-package}
 
-```
+\subsection{Shiny application}\label{shiny-application}
 
-The small minority of the figures make any attempt to indicate the values of the test cutoffs, which is an integral component of the ROC curve.
-We conjecture that this is mainly due to the use of default plotting procedures in statistical software.
-The software, by default, treats the ROC curve as a 2 dimensional object, obscuring the cutoff dimension.
-Gridlines and direct labels are also somewhat out of the ordinary.
-The absence of these features make accurate determination and comparison of the values more difficult.
-Many of the plots included large tables containing estimates and inference for AUCs, while the ROC curves themselves, numerous and without clear labels or reference lines, merely served as decoration.
-We aim to solve some of these problems by providing an easy-to-use plotting interface for the ROC curve that provides sensible defaults.
+We created a \pkg{shiny} application \citep{shiny} in order to make the
+features more accessible to non-\proglang{R} users. A limited subset of
+the functions of the \pkg{plotROC} can be performed on an example
+dataset or on data that users upload to the website. Resulting plots can
+be saved to the users' machine as a pdf or as a stand-alone html file.
+It can be used in any modern web browser with no other dependencies at
+the website here: http://sachsmc.shinyapps.io/plotROC.
 
-The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements.
-We favor the use of gridline and reference lines to facilitate accurate readings off of the axes.
-Direct labels are preferred over legends because they omit the additional cognitive step of matching line types to labels.
-Our \pkg{plotROC} package additionally provides plotting of cutoff values, using hover events in interactive use, and direct labels for print use.
-Exact confidence regions for points on the ROC curve are optionally calculated and displayed.
-Additionally, we use axis scales are adjusted to be denser near the margins 0 and 1.
-In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins.
-The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.
+\subsection{Quick start}\label{quick-start}
 
-```{r figure1, fig.width = 6, fig.height = 6, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves.
-Panel A shows a sparse ROC curve, with no design additions inside the plotting region.
-The plot results in more white space than anything else.
-It is difficult to accurately determine values without reference lines.
-Panel B shows a plot comparing 2 curves, with different line types and a legend.
-AUCs are also given in the legend.
-Panels B and C add gridlines, diagonal reference lines, and direct labels.
-\\label{figure1}"}
-library(plotROC)
+After installing, the interactive Shiny application can be run locally.
 
-set.seed(520)
-D0 <- rbinom(100, size = 1, prob = .5)
-M1 <- rnorm(100, mean = D0)
-M2 <- rnorm(100, mean = D0, sd = 1.5)
+\begin{verbatim}
+shiny_plotROC()
+\end{verbatim}
 
-R1 <- calculate_roc(M1, D0)
-R2 <- calculate_roc(M2, D0)
+\subsection{Command line basic usage}\label{command-line-basic-usage}
 
-par(mfrow = c(2,2))
+We start by creating an example data set. The marker we generate is
+moderately accurate for predicting disease status.
 
-# A
-plot(TPF ~ FPF, data = R1, type = 's', xlab = "1 - Specificity", ylab = "Sensitivity", main = "A")
-# B
-plot(TPF ~ FPF, data = R1, type = 's', xlab = "1 - Specificity", ylab = "Sensitivity", main = "B")
-lines(TPF ~ FPF, data = R2, type = 's', lty = 2)
-legend("bottomright", legend = c("Test A: AUC = 0.81", "Test B: AUC = 0.64"), lty = 1:2)
-
-# C
-plot(TPF ~ FPF, data = R1, type = 's', lwd  = 2, xlab = "False positive fraction", ylab = "True positive fraction", main = "C")
-abline(v = seq(0, 1, by = .2), lwd = 1, col = "grey80")
-abline(h = seq(0, 1, by = .2), lwd = 1, col = "grey80")
-abline(0, 1, col = "grey80")
-lines(TPF ~ FPF, data = R1, type = 's', lwd = 2)
-text(.2, .8, "Test A")
-
-# D
-plot(TPF ~ FPF, data = R1, type = 's', lwd  = 2, xlab = "False positive fraction", ylab = "True positive fraction", main = "D")
-abline(v = seq(0, 1, by = .2), lwd = 1, col = "grey80")
-abline(h = seq(0, 1, by = .2), lwd = 1, col = "grey80")
-abline(0, 1, col = "grey80")
-lines(TPF ~ FPF, data = R1, type = 's', lwd = 2)
-lines(TPF ~ FPF, data = R2, type = 's', lwd = 2)
-text(.2, .8, "A")
-text(.3, .4, "B")
-```
-
-# Usage of the Package
-
-## Shiny application
-
-We created a \pkg{shiny} application [@shiny] in order to make the features more accessible to non-\proglang{R} users.
-A limited subset of the functions of the \pkg{plotROC} can be performed on an example dataset or on data that users upload to the website.
-Resulting plots can be saved to the users' machine as a pdf or as a stand-alone html file.
-It can be used in any modern web browser with no other dependencies at the website here: http://sachsmc.shinyapps.io/plotROC.
-
-## Quick start
-After installing, the interactive Shiny application can be run locally. 
-```{r shiny, eval = FALSE}
-shiny_plotROC()
-```
-
-## Command line basic usage
-
-We start by creating an example data set. The marker we generate is moderately accurate for predicting disease status.
-
-```{r dataset}
+\begin{verbatim}
 library(plotROC)
@@
 M.ex <- rnorm(100, mean = D.ex)
-```
+\end{verbatim}
 
-Next we use the `calculate_roc` function to compute the empirical ROC curve.
-The disease status need not be coded as 0/1, but if it is not, `plotROC` assumes (with a warning) that the lowest value in sort order signifies disease-free status.
-This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.
+Next we use the \texttt{calculate\_roc} function to compute the
+empirical ROC curve. The disease status need not be coded as 0/1, but if
+it is not, \texttt{plotROC} assumes (with a warning) that the lowest
+value in sort order signifies disease-free status. This returns a
+dataframe with three columns: the cutoff values, the TPF and the FPF.
 
-```{r calc}
+\begin{verbatim}
 rocdata <- calculate_roc(M.ex, D.ex)
 str(rocdata)
-```
+\end{verbatim}
 
-The same data.frame `rocdata` can be used to generate an interactive plot to view in Rstudio viewer or web browser.
+\begin{verbatim}
+'data.frame':   100 obs. of  3 variables:
+ $ c  : num  -2.18 -1.67 -1.57 -1.55 -1.54 ...
+ $ TPF: num  1 1 1 1 1 1 1 1 1 1 ...
+ $ FPF: num  0.981 0.962 0.943 0.925 0.906 ...
+\end{verbatim}
 
-```{r inter, eval = FALSE}
+The same data.frame \texttt{rocdata} can be used to generate an
+interactive plot to view in Rstudio viewer or web browser.
+
+\begin{verbatim}
 plot_interactive_roc(rocdata)
-```
+\end{verbatim}
 
-The `rocdata` is passed to the `ggroc` function with an optional label. This creates a ggplot object of the ROC curve using the \pkg{ggplot2} package [@ggplot2].
+The \texttt{rocdata} is passed to the \texttt{ggroc} function with an
+optional label. This creates a ggplot object of the ROC curve using the
+\pkg{ggplot2} package \citep{ggplot2}.
 
-```{r test-a}
+\begin{verbatim}
 myrocplot <- ggroc(rocdata, label = "Example")
-```
+\end{verbatim}
 
-We can create an interactive ROC plot using the `export_interactive_roc` function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}.
-The key interactive feature of these plots is that the cutoff values nearest to the mouse cursor are displayed when hovering over the figure.
-Clicking makes the cutoff label stick until the next click.
-The character string can be copy-pasted into an html document, or better yet, using \pkg{knitr} [@knitr], we can `cat` the results and use the option `results = 'asis'` so that the interactive plot is displayed correctly.
-For examples of interactive plots and how to incorporate them into `knitr` documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the webpage https://sachsmc.github.io/plotROC/.
-Of note is the `prefix` option, which allows the user to assign each plot a unique identifier.
-This is necessary to prevent conflicts with multiple plots in a single webpage.
+We can create an interactive ROC plot using the
+\texttt{export\_interactive\_roc} function, which returns a character
+string containing the necessary \proglang{HTML} and
+\proglang{JavaScript}. The key interactive feature of these plots is
+that the cutoff values nearest to the mouse cursor are displayed when
+hovering over the figure. Clicking makes the cutoff label stick until
+the next click. The character string can be copy-pasted into an html
+document, or better yet, using \pkg{knitr} \citep{knitr}, we can
+\texttt{cat} the results and use the option
+\texttt{results = \textquotesingle{}asis\textquotesingle{}} so that the
+interactive plot is displayed correctly. For examples of interactive
+plots and how to incorporate them into \texttt{knitr} documents, see the
+package vignette (\texttt{vignette("examples", package = "plotROC")}) or
+the webpage https://sachsmc.github.io/plotROC/. Of note is the
+\texttt{prefix} option, which allows the user to assign each plot a
+unique identifier. This is necessary to prevent conflicts with multiple
+plots in a single webpage.
 
-```{r int-no, eval = FALSE}
+\begin{verbatim}
 cat(
@@
   )
-```
+\end{verbatim}
 
-The same `ggroc` object that we called `myrocplot` can be used to generate an ROC plot suitable for use in print.
-It annotates the cutoff values and is completely in black and white.
-A simple example with the default options is shown in figure \ref{first}.
+The same \texttt{ggroc} object that we called \texttt{myrocplot} can be
+used to generate an ROC plot suitable for use in print. It annotates the
+cutoff values and is completely in black and white. A simple example
+with the default options is shown in figure \ref{first}.
 
-```{r print, fig.width = 6, fig.height = 6, fig.cap = "Illustration of ROC curve plot generated by \\pkg{plotROC} for use in print. \\label{first}"}
+\begin{verbatim}
 plot_journal_roc(myrocplot, rocdata)
-```
+\end{verbatim}
 
-## Advanced options
+\begin{figure}[htbp]
+\centering
+\includegraphics{figure/print-1.pdf}
+\caption{Illustration of ROC curve plot generated by \pkg{plotROC} for
+use in print. \label{first}}
+\end{figure}
 
-### Click to view confidence region
+\subsection{Advanced options}\label{advanced-options}
 
-We use the `ci = TRUE` option in `calculcate_roc` and `ggroc` to compute confidence regions for points on the ROC curve using the @clopper1934use exact method.
-The significance level can be specified using the `alpha` option.
+\subsubsection{Click to view confidence
+region}\label{click-to-view-confidence-region}
 
-```{r test-a-ci}
+We use the \texttt{ci = TRUE} option in \texttt{calculcate\_roc} and
+\texttt{ggroc} to compute confidence regions for points on the ROC curve
+using the \citet{clopper1934use} exact method. The significance level
+can be specified using the \texttt{alpha} option.
+
+\begin{verbatim}
 rocdata <- calculate_roc(M.ex, D.ex, ci = TRUE, alpha = 0.05)
 myrocplot <- ggroc(rocdata, label = "Example", ci = TRUE)
-```
+\end{verbatim}
 
-For interactive plots, the confidence regions are automatically detected.
-When the user clicks on the ROC curve, the confidence region for the TPF and FPF is overlaid using a grey rectangle.
-The label and region stick until the next click.
+For interactive plots, the confidence regions are automatically
+detected. When the user clicks on the ROC curve, the confidence region
+for the TPF and FPF is overlaid using a grey rectangle. The label and
+region stick until the next click.
 
-```{r int-nob, eval = FALSE}
+\begin{verbatim}
 cat(
@@
   )
-```
+\end{verbatim}
 
-For use in print, we pass a small vector of cutoff locations at which to display the confidence regions. This is shown in figure \ref{conf}.
+For use in print, we pass a small vector of cutoff locations at which to
+display the confidence regions. This is shown in figure \ref{conf}.
 
-```{r printci, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} plot with exact confidence regions. \\label{conf}"}
+\begin{verbatim}
 plot_journal_roc(myrocplot, rocdata, n.cuts = 10, 
                  ci.at = c(-.5, .5, 2.1))
-```
+\end{verbatim}
 
-### Multiple ROC curves
+\begin{figure}[htbp]
+\centering
+\includegraphics{figure/printci-1.pdf}
+\caption{Illustration of \pkg{plotROC} plot with exact confidence
+regions. \label{conf}}
+\end{figure}
 
-If you have multiple tests of different types on the same subjects, you can use the `calculate_multi_roc` function to compute the empirical ROC curve for each test.
-Then the `multi_ggroc` function creates the appropriate type of `ggplot` object.
-Confidence regions are not supported for multiple curves at the time of writing.
+\subsubsection{Multiple ROC curves}\label{multiple-roc-curves}
 
-```{r multistart}
+If you have multiple tests of different types on the same subjects, you
+can use the \texttt{calculate\_multi\_roc} function to compute the
+empirical ROC curve for each test. Then the \texttt{multi\_ggroc}
+function creates the appropriate type of \texttt{ggplot} object.
+Confidence regions are not supported for multiple curves at the time of
+writing.
+
+\begin{verbatim}
 D.ex <- rbinom(100, 1, .5)
@@
 rocplot <- multi_ggroc(datalist)
-```
+\end{verbatim}
 
-This multi ggroc object can be passed to the `plot_journal_roc` and the `export_interactive_roc` functions, as desired.
+This multi ggroc object can be passed to the \texttt{plot\_journal\_roc}
+and the \texttt{export\_interactive\_roc} functions, as desired.
 
-Labels can be added easily with the `label` option. The length of the label element should match the number of plotted curves. The resulting plot is shown in figure \ref{multi}.
+Labels can be added easily with the \texttt{label} option. The length of
+the label element should match the number of plotted curves. The
+resulting plot is shown in figure \ref{multi}.
 
-```{r multi2, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} plot with multiple curves. \\label{multi} "}
+\begin{verbatim}
 rocplot <- multi_ggroc(datalist, label = c("M1", "M2", "M3"))
 plot_journal_roc(rocplot, datalist)
-```
+\end{verbatim}
 
-### Themes and annotations
+\begin{figure}[htbp]
+\centering
+\includegraphics{figure/multi2-1.pdf}
+\caption{Illustration of \pkg{plotROC} plot with multiple curves.
+\label{multi}}
+\end{figure}
 
-`plotROC` uses the `ggplot2` package to create ggplot objects.
-Therefore, themes and annotations can be added to `ggroc` objects in the usual `ggplot` way.
-A `plotROC` figure with a new theme, title, axis label, and AUC annotation is shown in figure \ref{annotate}.
+\subsubsection{Themes and annotations}\label{themes-and-annotations}
 
-```{r print2, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
+\texttt{plotROC} uses the \texttt{ggplot2} package to create ggplot
+objects. Therefore, themes and annotations can be added to
+\texttt{ggroc} objects in the usual \texttt{ggplot} way. A
+\texttt{plotROC} figure with a new theme, title, axis label, and AUC
+annotation is shown in figure \ref{annotate}.
+
+\begin{verbatim}
 library(ggplot2)
@@
   ylab("Sensitivity")
-```
+\end{verbatim}
 
-### Other estimation methods
+\begin{figure}[htbp]
+\centering
+\includegraphics{figure/print2-1.pdf}
+\caption{Using \pkg{ggplot2} themes and annotations with \pkg{plotROC}
+objects. \label{annotate}}
+\end{figure}
 
-By default `calculate_roc` computes the empirical ROC curve.
-There are other estimation methods out there, as we have summarized in the introduction.
-Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned.
-Then you can simply pass those values in a data frame to the `ggroc` function.
-For example, let us use the binormal method to create a smooth curve.
-This approach assumes that the test distribution is normal conditional on disease status.
+\subsubsection{Other estimation methods}\label{other-estimation-methods}
 
-```{r binormalsetup}
+By default \texttt{calculate\_roc} computes the empirical ROC curve.
+There are other estimation methods out there, as we have summarized in
+the introduction. Any estimation method can be used, as long as the
+cutoff, the TPF and the FPF are returned. Then you can simply pass those
+values in a data frame to the \texttt{ggroc} function. For example, let
+us use the binormal method to create a smooth curve. This approach
+assumes that the test distribution is normal conditional on disease
+status.
+
+\begin{verbatim}
 D.ex <- rbinom(100, 1, .5)
@@
                              )
-```
+\end{verbatim}
 
-Then we can pass this data.frame to the `ggroc` function as before. The example is shown in figure \ref{binorm}.
+Then we can pass this data.frame to the \texttt{ggroc} function as
+before. The example is shown in figure \ref{binorm}.
 
-```{r binormal, echo = TRUE, fig.width=6, fig.height=6, fig.cap = "Illustration of smooth binormal ROC curve. \\label{binorm}"}
+\begin{verbatim}
 binorm_plot <- ggroc(binorm_rocdata, label = "Binormal")
 plot_journal_roc(binorm_plot, binorm_rocdata)
-```
+\end{verbatim}
 
-# How it Works
+\begin{figure}[htbp]
+\centering
+\includegraphics{figure/binormal-1.pdf}
+\caption{Illustration of smooth binormal ROC curve. \label{binorm}}
+\end{figure}
 
-# Discussion
+\section{How it Works}\label{how-it-works}
 
-Bibliography 
-===============
+\section{Discussion}\label{discussion}
+
+%\bibliographystyle{jss}
+\bibliography{plotroc}
+
+\end{document}

### 2015-01-28 Paper/sachs2014plotROC.Rmd
First draft complete

@@
 The same data.frame `rocdata` can be used to generate an interactive plot to view in Rstudio viewer or web browser.
+A screen shot of an interactive plot is shown in figure \ref{interact}.
+Hovering over the display shows the cutoff value at the point nearest to the cursor.
+Clicking makes the cutoff label stick until the next click, and if confidence regions are available, clicks will also display those as grey rectangles.
 
@@
 ```
+
+\begin{figure}[ht]
+\centering
+\includegraphics{screen-shot.pdf}
+\caption{Screen shot of an interactive plot created with \pkg{plotROC} being displayed in the Rstudio viewer.
+Hovering the mouse cursor over the plot causes the cutoff label nearest to the cursor to be displayed.
+Clicking will display a confidence region, if available, and make the label stick until the next click.
+For live examples, see the package vignette, or go to http://sachsmc.github.io/plotROC.
+\label{interact}}
+\end{figure}
 
@@
 
+Another potential use of this approach is for plotting time-dependent ROC curves for time-to-event outcomes estimated as desribed in [@heagerty2000time].
+
 # How it Works
+
+\pkg{plotROC} makes use of \pkg{ggplot2} [@ggplot2], \pkg{gridSVG} [@gridsvg], and \pkg{d3.js} [@bostock2011d3] to create interactive plots.
+The first step in the process is to create `ggplot` objects using the `ggroc` or `multi_ggroc` functions.
+These functions return standard `ggplot` objects that include basic styling, hidden cutoff labels, and hidden confidence regions.
+They can be plotted and inspected in the \proglang{R} console.
+These form the basis for both the print versions and the interactive versions of the plots.
+Creating a print version by using the `plot_journal_roc` function simply makes visible a subset of the hidden cutoff labels and confidence regions, if available.
+
+\pkg{plotROC} makes interactive plots by first converting the `ggplot` object into a scalable vector graphic (svg) object with the `grid.export` function.
+This function maps each element of the plot to a corresponding element of the svg markup language.
+We keep track of the names of the points and labels elements so that we can add interactivity using \pkg{d3.js} and \proglang{JavaScript}.
+The main interactive feature we wanted was to be able to display the cutoff labels at the points on the ROC curve closest to the mouse cursor.
+
+There are many ways to solve this with \pkg{d3.js}, but we decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve.
+The idea is that for the set of cutoff points along the ROC curve, the `d3.geom.voronoi` function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point.
+Hover events are bound to the polygons so that when the mouse cursor moves around the plotting region, the closest point on the ROC curve is made visible.
+Similarly, click events are bound to the polygons so that the appropriate confidence region is made visible upon clicking.
+The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by `export_interactive_roc`.
+Figure \ref{flow} outlines the \pkg{plotROC} process.
+
+```{r diagram, include = FALSE}
+library(DiagrammeR)
+
+DiagrammeR("
+  graph LR;
+    A[ROC Estimate]-->|ggroc|B(ggplot object);
+    B-->|plot_journal_roc|C(Static Plot);
+    B-->|grid.export|D(svg object);
+    D-->E(Interactive Plot);
+    F(d3.js functions)-->E;
+    G(css styles)-->E;
+
+  style A fill:#DCEBE3
+  style C fill:#A2EB86
+  style E fill:#A2EB86
+")
+
+```
+
+\begin{figure}[ht]
+\centering
+\includegraphics{diagram.pdf}
+\caption{Flowchart illustrating the approach that \pkg{plotROC} takes to generate either static plots for print or interactive plots for web-use. \label{flow}}
+\end{figure}
+
+This approach is similar to what is done in the \pkg{gridSVG} `grid.animate` function, which uses the svg `<animate />` tags.
+However, the available features were not sufficient for our needs, which is why we used \pkg{d3.js}.
+There are several other \proglang{R} packages that aim to create interactive figures.
+The authors of \pkg{animint} [@animint] created an extensive \proglang{JavaScript} library that creates plots in a similar way as \pkg{ggplot2}.
+A set of interactive features can be added to plots using \pkg{d3.js}.
+\pkg{ggvis} [@ggvis], \pkg{rCharts} [@rcharts], and the more recently released \pkg{htmlwidgets} [@htmlwidgets] all leverage existing charting libraries written in \proglang{JavaScript}.
+Their general approach is to manipulate the data and create options in \proglang{R}, and then let the charting libraries handle the rendering and interactivity.
+\pkg{plotROC} lets \proglang{R} do the rendering, allowing the figures to be consistent across print and web-based media, and retaining the distinctive \proglang{R} style.
 
@@
 
+Here we have illustrated the usage and described the mechanics of a new \proglang{R} package for creating ROC curve plots.
+The functions are easy to use, even for non-\proglang{R} users _via_ the web application, yet have sufficient flexibility to meet the needs of power users.
+Our approach to creating interactive plots differs from other interactive charting packages.
+We found that existing approaches did not meet the highly specialized needs of plotting ROC curves.
+While ROC curve plots can technically be created with even the most basic plotting tools, we find that specialized functions make the results clearer and more informative.
+
 Bibliography 

### 2015-01-28 Paper/sachs2014plotROC.tex
First draft complete

@@
 The same data.frame \texttt{rocdata} can be used to generate an
-interactive plot to view in Rstudio viewer or web browser.
+interactive plot to view in Rstudio viewer or web browser. A screen shot
+of an interactive plot is shown in figure \ref{interact}. Hovering over
+the display shows the cutoff value at the point nearest to the cursor.
+Clicking makes the cutoff label stick until the next click, and if
+confidence regions are available, clicks will also display those as grey
+rectangles.
 
@@
 \end{verbatim}
+
+\begin{figure}[ht]
+\centering
+\includegraphics{screen-shot.pdf}
+\caption{Screen shot of an interactive plot created with \pkg{plotROC} being displayed in the Rstudio viewer.
+Hovering the mouse cursor over the plot causes the cutoff label nearest to the cursor to be displayed.
+Clicking will display a confidence region, if available, and make the label stick until the next click.
+For live examples, see the package vignette, or go to http://sachsmc.github.io/plotROC.
+\label{interact}}
+\end{figure}
 
@@
 
+Another potential use of this approach is for plotting time-dependent
+ROC curves for time-to-event outcomes estimated as desribed in
+\citep{heagerty2000time}.
+
 \section{How it Works}\label{how-it-works}
 
+\pkg{plotROC} makes use of \pkg{ggplot2} \citep{ggplot2}, \pkg{gridSVG}
+\citep{gridsvg}, and \pkg{d3.js} \citep{bostock2011d3} to create
+interactive plots. The first step in the process is to create
+\texttt{ggplot} objects using the \texttt{ggroc} or
+\texttt{multi\_ggroc} functions. These functions return standard
+\texttt{ggplot} objects that include basic styling, hidden cutoff
+labels, and hidden confidence regions. They can be plotted and inspected
+in the \proglang{R} console. These form the basis for both the print
+versions and the interactive versions of the plots. Creating a print
+version by using the \texttt{plot\_journal\_roc} function simply makes
+visible a subset of the hidden cutoff labels and confidence regions, if
+available.
+
+\pkg{plotROC} makes interactive plots by first converting the
+\texttt{ggplot} object into a scalable vector graphic (svg) object with
+the \texttt{grid.export} function. This function maps each element of
+the plot to a corresponding element of the svg markup language. We keep
+track of the names of the points and labels elements so that we can add
+interactivity using \pkg{d3.js} and \proglang{JavaScript}. The main
+interactive feature we wanted was to be able to display the cutoff
+labels at the points on the ROC curve closest to the mouse cursor.
+
+There are many ways to solve this with \pkg{d3.js}, but we decided to
+use Voronoi polygons to map the cursor location to the nearest point on
+the ROC curve. The idea is that for the set of cutoff points along the
+ROC curve, the \texttt{d3.geom.voronoi} function chain computes a set of
+polygons overlaying the plotting region such that the area of each
+polygon contains the region of the plot closest to it's corresponding
+cutoff point. Hover events are bound to the polygons so that when the
+mouse cursor moves around the plotting region, the closest point on the
+ROC curve is made visible. Similarly, click events are bound to the
+polygons so that the appropriate confidence region is made visible upon
+clicking. The svg code and all necessary \proglang{JavaScript} code is
+returned in the character string provided by
+\texttt{export\_interactive\_roc}. Figure \ref{flow} outlines the
+\pkg{plotROC} process.
+
+\begin{figure}[ht]
+\centering
+\includegraphics{diagram.pdf}
+\caption{Flowchart illustrating the approach that \pkg{plotROC} takes to generate either static plots for print or interactive plots for web-use. \label{flow}}
+\end{figure}
+
+This approach is similar to what is done in the \pkg{gridSVG}
+\texttt{grid.animate} function, which uses the svg
+\texttt{\textless{}animate /\textgreater{}} tags. However, the available
+features were not sufficient for our needs, which is why we used
+\pkg{d3.js}. There are several other \proglang{R} packages that aim to
+create interactive figures. The authors of \pkg{animint} \citep{animint}
+created an extensive \proglang{JavaScript} library that creates plots in
+a similar way as \pkg{ggplot2}. A set of interactive features can be
+added to plots using \pkg{d3.js}. \pkg{ggvis} \citep{ggvis},
+\pkg{rCharts} \citep{rcharts}, and the more recently released
+\pkg{htmlwidgets} \citep{htmlwidgets} all leverage existing charting
+libraries written in \proglang{JavaScript}. Their general approach is to
+manipulate the data and create options in \proglang{R}, and then let the
+charting libraries handle the rendering and interactivity. \pkg{plotROC}
+lets \proglang{R} do the rendering, allowing the figures to be
+consistent across print and web-based media, and retaining the
+distinctive \proglang{R} style.
+
 \section{Discussion}\label{discussion}
+
+Here we have illustrated the usage and described the mechanics of a new
+\proglang{R} package for creating ROC curve plots. The functions are
+easy to use, even for non-\proglang{R} users \emph{via} the web
+application, yet have sufficient flexibility to meet the needs of power
+users. Our approach to creating interactive plots differs from other
+interactive charting packages. We found that existing approaches did not
+meet the highly specialized needs of plotting ROC curves. While ROC
+curve plots can technically be created with even the most basic plotting
+tools, we find that specialized functions make the results clearer and
+more informative.
 

### 2015-01-29 Paper/sachs2014plotROC.Rmd
editing, moved some figures

@@
 ---
-title: '\pkg{plotROC}: An \proglang{R} Package to Improve the State of ROC Curve Plots in the Medical Literature'
-shorttitle: '\pkg{plotROC}: An \proglang{R} Package for ROC Curve Plots'
-plaintitle: 'plotROC: An R Package to Improve the State of ROC Curve Plots in the Medical Literature'
+title: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
+shorttitle: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
+plaintitle: 'plotROC: A Better Tool for Plotting ROC Curves'
 author:
@@
 Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers.
-We evaluate current trends in the design of ROC curve plots with a literature review.
+We reviewed a sample of ROC curve plots from the major oncology journals in order to assess current trends in usage and design elements.
 Our review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display.
@@
 In medicine, ROC curves have a long history of use for evaluating diagnostic tests in radiology and general diagnostics.
-ROC curves have also been used for a long time in signal detection theory.
+ROC curves have also been used for decades in signal detection theory.
 
-The accuracy of a diagnostic test can be evaluated by considering the two possible types of errors: false positives, and false negatives.
+The accuracy of a diagnostic test can be evaluated by considering two possible types of errors: false positives, and false negatives.
 For a continuous measurement that we denote as $M$, convention dictates that a test positive is defined as $M$ exceeding some fixed threshold $c$: $M > c$.
-In reference to the binary outcome that we denote as $D$, a good outcome of the test is when the test is positive among an individual who truly has a disease: $D = 1$.
-A bad outcome is when the test is positive among an individual who does not have the disease $D = 0$.
+In reference to the binary outcome that we denote as $D$, a good outcome of the test is when the test is positive for an individual who truly has a disease: $D = 1$.
+A bad outcome is when the test is positive for an individual who does not have the disease $D = 0$.
 
-Formally, for a fixed cutoff $c$, the true positive fractions is the probability of a test positive among the diseased population:
+Formally, for a fixed cutoff $c$, the true positive fraction is the probability of a test positive in the diseased population:
 
@@
 
-and the false positive fraction is the probability of a test positive among the healthy population:
+and the false positive fraction is the probability of a test positive in the healthy population:
 
@@
 
-Since the cutoff $c$ is not usually fixed in advance, we can plot the TPF against the FPF for all possible values of $c$.
-This is exactly what the ROC curve is, $FPF(c)$ on the $x$ axis and $TPF(c)$ along the $y$ axis.
+Since the cutoff $c$ is not fixed in advance, we can plot the TPF against the FPF for all possible values of $c$.
+This is exactly what the ROC curve is, a plot of $FPF(c)$ on the $x$ axis and $TPF(c)$ along the $y$ axis as $c$ varies.
 A useless test that is not informative at all in regards to the disease status has $TPF(c) = FPF(c)$ for all $c$.
@@
 The greater the AUC, the more informative the test.
-The AUC summarizes the complexities of the ROC curve into a single number and therefore is widely used to facilitate comparisons between tests and populations.
+The AUC summarizes the complexities of the ROC curve into a single number and therefore is widely used to facilitate comparisons between tests and across populations.
 It has been criticized for the same reason because it does not fully characterize the trade-offs between false- and true-positives.
@@
 This can be useful for aiding viewers in choosing an optimal cutoff for decision making, for comparing a small number of candidate tests, and for generally illustrating the performance of the test as a classifier.
-In practice, once the FPF and TPF are computed for each unique observed cutoff value, they can be plotted as a simple line chart or scatter plot using standard statistical plotting tools.
+In practice, once the FPF and TPF are computed for each unique observed cutoff value, they can be plotted as a simple line chart or scatter plot using standard plotting tools.
 This often leads to the unfortunate design choice of obscuring the critical and useful third dimension, the range of cutoff values $c$.
@@
 Another key design element is the use of a diagonal guideline for comparison.
-The diagonal guideline serves as a baseline for comparison, allowing observers to roughly estimate the area between the diagonal and the estimated ROC curve, which serves as a proxy for estimating the value of the test for classification above a coin-flip.
+They allow observers to roughly estimate the area between the diagonal and the estimated ROC curve, which serves as a proxy for estimating the value of the test for classification above a useless test.
 Likewise, gridlines inside the plotting region and carefully selected axes allow for accurate assessment of the TPF and FPF at particular points along the ROC curve.
-Many medical studies use ROC curves to compare a multitude of candidate tests to each other and to the null diagonal.
-In those cases, curves need to be distinguished by using a legend combined with different colors or line types, or direct labels inside the plotting region.
+Many medical studies use ROC curves to compare a multitude of candidate tests to each other.
+In those cases, curves need to be distinguished by using different colors or line types combined with a legend, or direct labels inside the plotting region.
 
-In the medical literature, FPF and TPF are usually referred to in terms of the jargon terms Sensitivity and Specificity.
+In the medical literature, FPF and TPF are usually referred to in terms of the jargon sensitivity and specificity.
 Sensitivity is equivalent to the true positive fraction.
@@
 Therefore, once the appropriate statistics are estimated, existing plotting functions can be used to create an ROC curve plot.
-The addition of axis labels, titles, legends, and so on, indicate a chart as such.
-In our literature review, we observed plots with the distinctive characteristics of the base plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions.
+Viewers can identify ROC plots through context, by observing the shape of the line, and through the addition of axis labels, titles, legends, and so on.
+In our review of the oncology literature, we observed plots with the distinctive characteristics of the plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions [@arr].
 
@@
 The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions.
+My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.
 
@@
 Anyone giving a cursory look at any of the major medical journals is likely to find at least one ROC curve plot.
-We sought assess the usage of ROC curve plots and to evaluate the design choices made in the current oncology literature by conducting a small literature review.
+We sought to assess the usage of ROC curve plots and to evaluate the design choices made in the current oncology literature by conducting a small literature review.
 We searched Pubmed for  clinical trials or observational studies in humans reported in major oncology journals for the past 10 years for the terms "ROC Curve" OR "ROC Analysis" OR "Receiver operating characteristic curve".
@@
 From those papers, 47 images were extracted and reviewed.
-The exact specifications for the Pubmed query are available in the supplementary materials.
+The exact specifications for the Pubmed query are available in the manuscript source files.
 
 Each image consisted of a single ROC curve plot or a panel of multiple plots.
-Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of diagonal guide line, whether any cutpoint were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the area under the curve.
+Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of a diagonal guide line, whether any cutpoints were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the AUC.
 The numerical results of the survey are summarized in table \ref{table1}.
@@
 
-rev <- read.csv("../../Literature Review/rocusage.csv", stringsAsFactors = FALSE)
+rev <- read.csv("litreview/rocusage.csv", stringsAsFactors = FALSE)
 rev <- subset(rev, flag != "comedy")
@@
 Direct labels are preferred over legends because they omit the additional cognitive step of matching line types to labels.
-Our \pkg{plotROC} package additionally provides plotting of cutoff values, using hover events in interactive use, and direct labels for print use.
+Our \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use.
 Exact confidence regions for points on the ROC curve are optionally calculated and displayed.
@@
 lines(TPF ~ FPF, data = R2, type = 's', lty = 2)
-legend("bottomright", legend = c("Test A: AUC = 0.81", "Test B: AUC = 0.64"), lty = 1:2)
+legend("bottomright", legend = c("Test A: AUC = 0.81", "Test B: AUC = 0.64"), lty = 1:2, cex = .75)
 
@@
 
+## Installation and loading
+
+\pkg{plotROC} can be installed from the Comprehensive \proglang{R} Archive Network, or installed from source.
+
+```{r load, eval = FALSE}
+install.packages("plotROC")
+library(plotROC)
+```
+
 ## Quick start
@@
 
-```{r dataset}
+```{r dataset, echo = -1}
 library(plotROC)
@@
 Next we use the `calculate_roc` function to compute the empirical ROC curve.
-The disease status need not be coded as 0/1, but if it is not, `plotROC` assumes (with a warning) that the lowest value in sort order signifies disease-free status.
+The disease status need not be coded as 0/1, but if it is not, `calculate_roc` assumes (with a warning) that the lowest value in sort order signifies disease-free status.
 This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.
@@
 \centering
-\includegraphics{screen-shot.pdf}
+\includegraphics{figure/screen-shot.pdf}
 \caption{Screen shot of an interactive plot created with \pkg{plotROC} being displayed in the Rstudio viewer.
@@
 
-```{r print2, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
+```{r print2, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
 library(ggplot2)
@@
            label = "AUC = 0.80") + 
-  ylab("Sensitivity")
+  scale_y_continuous("Sensitivity") + 
+  scale_x_continuous("1 - Specificity")
 ```
@@
 
+## export as pdf
+
 ```
@@
 \centering
-\includegraphics{diagram.pdf}
+\includegraphics{figure/diagram.pdf}
 \caption{Flowchart illustrating the approach that \pkg{plotROC} takes to generate either static plots for print or interactive plots for web-use. \label{flow}}

### 2015-01-29 Paper/sachs2014plotROC.Rmd
editing and readme

@@
 library(knitr)
+library(stringr)
 knit_hooks$set(plot = function(x, options) {
@@
 opts_chunk$set(dev="tikz", comment = NA)
+knit_hooks$set(source = function(x, options){
+  if (!is.null(options$verbatim) && options$verbatim){
+    opts = gsub(",\\s*verbatim\\s*=\\s*TRUE\\s*", "", options$params.src)
+    opts = gsub(",\\s*eval\\s*=\\s*FALSE\\s*", "", opts)
+    bef = sprintf('\n\n    ```{r %s}\n', opts, "\n")
+    stringr::str_c(
+      bef, 
+      knitr:::indent_block(paste(x, collapse = '\n'), "    "), 
+      "\n    ```\n"
+    )
+  } else {
+    stringr::str_c("\n\n```", tolower(options$engine), "\n", 
+      paste(x, collapse = '\n'), "\n```\n\n"
+    )
+  }
+})
 ```
@@
 There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions.
-The \pkg{ROCR} package [@rocr] plots the FPF versus TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself.
+The \pkg{ROCR} package [@rocr] plots the FPF _versus_ TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself.
 A legend for the color scale is place along the vertical axis on the right of the plotting region.
@@
 The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements.
-We favor the use of gridline and reference lines to facilitate accurate readings off of the axes.
-Direct labels are preferred over legends because they omit the additional cognitive step of matching line types to labels.
+We favor the use of gridlines and a diagonal reference line to facilitate accurate readings off of the axes.
+Direct labels are preferred over legends because they omit the additional cognitive step of matching line types or colors to labels.
 Our \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use.
@@
 
-The same data.frame `rocdata` can be used to generate an interactive plot to view in Rstudio viewer or web browser.
+The same data.frame `rocdata` can be used to generate an interactive plot to view in Rstudio viewer or web browser by passing it to the `plot_interactive_roc` function.
+Give the function an optional path to an html file as an argument called `file` to save the interactive plot as a complete web page.
 A screen shot of an interactive plot is shown in figure \ref{interact}.
@@
 Clicking makes the cutoff label stick until the next click.
-The character string can be copy-pasted into an html document, or better yet, using \pkg{knitr} [@knitr], we can `cat` the results and use the option `results = 'asis'` so that the interactive plot is displayed correctly.
-For examples of interactive plots and how to incorporate them into `knitr` documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the webpage https://sachsmc.github.io/plotROC/.
-Of note is the `prefix` option, which allows the user to assign each plot a unique identifier.
-This is necessary to prevent conflicts with multiple plots in a single webpage.
+The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr].
+In a \pkg{knitr} document, it is necessary to use the `cat` function on the results and use the chunk option `results = 'asis'` so that the interactive plot is displayed correctly.
+For documents that contain multiple interactive plots, it is necessary to assign each plot a unique name using the `prefix` argument of `export_interactive_roc`.
+This is necessary to ensure that the \proglang{JavaScript} code manipulates the correct svg elements.
+For examples of interactive plots and how to incorporate them into `knitr` documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the web page https://sachsmc.github.io/plotROC/.
 
-```{r int-no, eval = FALSE}
+```{r int-no, results = 'asis', eval = FALSE, verbatim = TRUE}
 cat(
@@
 
-`plotROC` uses the `ggplot2` package to create ggplot objects.
-Therefore, themes and annotations can be added to `ggroc` objects in the usual `ggplot` way.
-A `plotROC` figure with a new theme, title, axis label, and AUC annotation is shown in figure \ref{annotate}.
+\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be plotted.
+Therefore, themes and annotations can be added in the usual \pkg{ggplot2} way.
+A `plot_journal_roc` figure with a new theme, title, axis label, and AUC annotation is shown in figure \ref{annotate}.
 

### 2015-01-29 Paper/sachs2014plotROC.tex
editing, moved some figures
editing and readme

@@
 
-\title{\pkg{plotROC}: An \proglang{R} Package to Improve the State of ROC Curve
-Plots in the Medical Literature}
+\title{\pkg{plotROC}: A Better Tool for Plotting ROC Curves}
 
@@
 \Plainauthor{ Michael C. Sachs} %% comma-separated
-\Plaintitle{plotROC: An R Package to Improve the State of ROC Curve Plots in the
-Medical Literature} %% without formatting
+\Plaintitle{plotROC: A Better Tool for Plotting ROC Curves} %% without formatting
 
-\Shorttitle{\pkg{plotROC}: An \proglang{R} Package for ROC Curve Plots} %% a short title (if necessary)
+\Shorttitle{\pkg{plotROC}: A Better Tool for Plotting ROC Curves} %% a short title (if necessary)
 %% an abstract and keywords
@@
   diagnostic test, ROC curves are used in oncology to evaluate screening,
-  diagnostic, prognostic and predictive biomarkers. We evaluate current
-  trends in the design of ROC curve plots with a literature review. Our
-  review suggests that ROC curve plots are often ineffective as
-  statistical charts and that poor design obscures the relevant
-  information the chart is intended to display. We describe our new
-  \proglang{R} package that was created to address the shortcomings of
-  existing tools. The package has functions to create informative ROC
-  curve plots, with sensible defaults, for use in print or as an
-  interactive web-based plot. A web application was developed to reach a
-  broader audience of scientists who do not use R.
+  diagnostic, prognostic and predictive biomarkers. We reviewed a sample
+  of ROC curve plots from the major oncology journals in order to assess
+  current trends in usage and design elements. Our review suggests that
+  ROC curve plots are often ineffective as statistical charts and that
+  poor design obscures the relevant information the chart is intended to
+  display. We describe our new \proglang{R} package that was created to
+  address the shortcomings of existing tools. The package has functions to
+  create informative ROC curve plots, with sensible defaults, for use in
+  print or as an interactive web-based plot. A web application was
+  developed to reach a broader audience of scientists who do not use R.
 }
@@
 diagnostic tests in radiology and general diagnostics. ROC curves have
-also been used for a long time in signal detection theory.
+also been used for decades in signal detection theory.
 
-The accuracy of a diagnostic test can be evaluated by considering the
-two possible types of errors: false positives, and false negatives. For
-a continuous measurement that we denote as \(M\), convention dictates
-that a test positive is defined as \(M\) exceeding some fixed threshold
+The accuracy of a diagnostic test can be evaluated by considering two
+possible types of errors: false positives, and false negatives. For a
+continuous measurement that we denote as \(M\), convention dictates that
+a test positive is defined as \(M\) exceeding some fixed threshold
 \(c\): \(M > c\). In reference to the binary outcome that we denote as
-\(D\), a good outcome of the test is when the test is positive among an
+\(D\), a good outcome of the test is when the test is positive for an
 individual who truly has a disease: \(D = 1\). A bad outcome is when the
-test is positive among an individual who does not have the disease
+test is positive for an individual who does not have the disease
 \(D = 0\).
 
-Formally, for a fixed cutoff \(c\), the true positive fractions is the
-probability of a test positive among the diseased population:
+Formally, for a fixed cutoff \(c\), the true positive fraction is the
+probability of a test positive in the diseased population:
 
@@
 
-and the false positive fraction is the probability of a test positive
-among the healthy population:
+and the false positive fraction is the probability of a test positive in
+the healthy population:
 
@@
 
-Since the cutoff \(c\) is not usually fixed in advance, we can plot the
-TPF against the FPF for all possible values of \(c\). This is exactly
-what the ROC curve is, \(FPF(c)\) on the \(x\) axis and \(TPF(c)\) along
-the \(y\) axis. A useless test that is not informative at all in regards
-to the disease status has \(TPF(c) = FPF(c)\) for all \(c\). The ROC
-plot of a useless test is thus the diagonal line. A perfect test that is
-completely informative about disease status has \(TPF(c) = 1\) and
-\(FPF(c) = 0\) for all \(c\).
+Since the cutoff \(c\) is not fixed in advance, we can plot the TPF
+against the FPF for all possible values of \(c\). This is exactly what
+the ROC curve is, a plot of \(FPF(c)\) on the \(x\) axis and \(TPF(c)\)
+along the \(y\) axis as \(c\) varies. A useless test that is not
+informative at all in regards to the disease status has
+\(TPF(c) = FPF(c)\) for all \(c\). The ROC plot of a useless test is
+thus the diagonal line. A perfect test that is completely informative
+about disease status has \(TPF(c) = 1\) and \(FPF(c) = 0\) for all
+\(c\).
 
@@
 complexities of the ROC curve into a single number and therefore is
-widely used to facilitate comparisons between tests and populations. It
-has been criticized for the same reason because it does not fully
-characterize the trade-offs between false- and true-positives.
+widely used to facilitate comparisons between tests and across
+populations. It has been criticized for the same reason because it does
+not fully characterize the trade-offs between false- and true-positives.
 
@@
 cutoff value, they can be plotted as a simple line chart or scatter plot
-using standard statistical plotting tools. This often leads to the
-unfortunate design choice of obscuring the critical and useful third
-dimension, the range of cutoff values \(c\).
+using standard plotting tools. This often leads to the unfortunate
+design choice of obscuring the critical and useful third dimension, the
+range of cutoff values \(c\).
 
 Another key design element is the use of a diagonal guideline for
-comparison. The diagonal guideline serves as a baseline for comparison,
-allowing observers to roughly estimate the area between the diagonal and
-the estimated ROC curve, which serves as a proxy for estimating the
-value of the test for classification above a coin-flip. Likewise,
-gridlines inside the plotting region and carefully selected axes allow
-for accurate assessment of the TPF and FPF at particular points along
-the ROC curve. Many medical studies use ROC curves to compare a
-multitude of candidate tests to each other and to the null diagonal. In
-those cases, curves need to be distinguished by using a legend combined
-with different colors or line types, or direct labels inside the
-plotting region.
+comparison. They allow observers to roughly estimate the area between
+the diagonal and the estimated ROC curve, which serves as a proxy for
+estimating the value of the test for classification above a useless
+test. Likewise, gridlines inside the plotting region and carefully
+selected axes allow for accurate assessment of the TPF and FPF at
+particular points along the ROC curve. Many medical studies use ROC
+curves to compare a multitude of candidate tests to each other. In those
+cases, curves need to be distinguished by using different colors or line
+types combined with a legend, or direct labels inside the plotting
+region.
 
 In the medical literature, FPF and TPF are usually referred to in terms
-of the jargon terms Sensitivity and Specificity. Sensitivity is
-equivalent to the true positive fraction. Specificity is 1 - FPF, the
-true negative fraction. Sometimes, the FPF and TPF are incorrectly
-referred to as rates, using the abbreviations FPR and TPR. These are
-probabilities and their estimates are proportions, therefore we prefer
-the use of the term fraction as opposed to rate.
+of the jargon sensitivity and specificity. Sensitivity is equivalent to
+the true positive fraction. Specificity is 1 - FPF, the true negative
+fraction. Sometimes, the FPF and TPF are incorrectly referred to as
+rates, using the abbreviations FPR and TPR. These are probabilities and
+their estimates are proportions, therefore we prefer the use of the term
+fraction as opposed to rate.
 
@@
 once the appropriate statistics are estimated, existing plotting
-functions can be used to create an ROC curve plot. The addition of axis
-labels, titles, legends, and so on, indicate a chart as such. In our
-literature review, we observed plots with the distinctive
-characteristics of the base plotting functions from Microsoft Office,
-\proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting
-functions.
+functions can be used to create an ROC curve plot. Viewers can identify
+ROC plots through context, by observing the shape of the line, and
+through the addition of axis labels, titles, legends, and so on. In our
+review of the oncology literature, we observed plots with the
+distinctive characteristics of the plotting functions from Microsoft
+Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R}
+plotting functions \citep{arr}.
 
@@
 that contain dedicated plotting functions. The \pkg{ROCR} package
-\citep{rocr} plots the FPF versus TPF, as usual, and then takes the
-interesting approach of encoding the cutoff values as a separate color
-scale along the ROC curve itself. A legend for the color scale is place
-along the vertical axis on the right of the plotting region. The
+\citep{rocr} plots the FPF \emph{versus} TPF, as usual, and then takes
+the interesting approach of encoding the cutoff values as a separate
+color scale along the ROC curve itself. A legend for the color scale is
+place along the vertical axis on the right of the plotting region. The
 \pkg{pROC} package \citep{pROC} is mainly focused on estimating
@@
 functions to create nice displays of the curves along with shaded
-confidence regions.
+confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2}
+\citep{ggplot2} plotting library to create clear, informative ROC plots,
+with interactive features for use on the web, and sensible defaults for
+use in print.
 
@@
 Anyone giving a cursory look at any of the major medical journals is
-likely to find at least one ROC curve plot. We sought assess the usage
-of ROC curve plots and to evaluate the design choices made in the
+likely to find at least one ROC curve plot. We sought to assess the
+usage of ROC curve plots and to evaluate the design choices made in the
 current oncology literature by conducting a small literature review. We
@@
 exact specifications for the Pubmed query are available in the
-supplementary materials.
+manuscript source files.
 
@@
 (sensitivity/ 1 - specificity or true/false positive fractions),
-presence or absence of grid lines, presence or absence of diagonal guide
-line, whether any cutpoint were indicated, the type of curve label
-(legend or direct label), and presence of other textual annotations such
-as the area under the curve. The numerical results of the survey are
+presence or absence of grid lines, presence or absence of a diagonal
+guide line, whether any cutpoints were indicated, the type of curve
+label (legend or direct label), and presence of other textual
+annotations such as the AUC. The numerical results of the survey are
 summarized in table \ref{table1}.
@@
 ROC curve plots, and the associated design elements. We favor the use of
-gridline and reference lines to facilitate accurate readings off of the
-axes. Direct labels are preferred over legends because they omit the
-additional cognitive step of matching line types to labels. Our
-\pkg{plotROC} package additionally provides plotting of cutoff values,
-using hover events in interactive use, and direct labels for print use.
-Exact confidence regions for points on the ROC curve are optionally
-calculated and displayed. Additionally, we use axis scales are adjusted
-to be denser near the margins 0 and 1. In medical applications, it is
-often necessary to have a very low FPR (less than 10\%, for instance),
-therefore the smaller scales are useful for accurately determining
-values near the margins. The next section details the usage of the
-\pkg{plotROC} \proglang{R} package and these features.
+gridlines and a diagonal reference line to facilitate accurate readings
+off of the axes. Direct labels are preferred over legends because they
+omit the additional cognitive step of matching line types or colors to
+labels. Our \pkg{plotROC} package additionally provides plotting of
+cutoff values, which are displayed interactively with the web-based
+output option, and direct labels for print use. Exact confidence regions
+for points on the ROC curve are optionally calculated and displayed.
+Additionally, we use axis scales are adjusted to be denser near the
+margins 0 and 1. In medical applications, it is often necessary to have
+a very low FPR (less than 10\%, for instance), therefore the smaller
+scales are useful for accurately determining values near the margins.
+The next section details the usage of the \pkg{plotROC} \proglang{R}
+package and these features.
 
@@
 
+\subsection{Installation and loading}\label{installation-and-loading}
+
+\pkg{plotROC} can be installed from the Comprehensive \proglang{R}
+Archive Network, or installed from source.
+
+\begin{verbatim}
+install.packages("plotROC")
+library(plotROC)
+\end{verbatim}
+
 \subsection{Quick start}\label{quick-start}
@@
 \begin{verbatim}
-library(plotROC)
 D.ex <- rbinom(100, size = 1, prob = .5)
@@
 empirical ROC curve. The disease status need not be coded as 0/1, but if
-it is not, \texttt{plotROC} assumes (with a warning) that the lowest
-value in sort order signifies disease-free status. This returns a
+it is not, \texttt{calculate\_roc} assumes (with a warning) that the
+lowest value in sort order signifies disease-free status. This returns a
 dataframe with three columns: the cutoff values, the TPF and the FPF.
@@
 The same data.frame \texttt{rocdata} can be used to generate an
-interactive plot to view in Rstudio viewer or web browser. A screen shot
-of an interactive plot is shown in figure \ref{interact}. Hovering over
-the display shows the cutoff value at the point nearest to the cursor.
+interactive plot to view in Rstudio viewer or web browser by passing it
+to the \texttt{plot\_interactive\_roc} function. Give the function an
+optional path to an html file as an argument called \texttt{file} to
+save the interactive plot as a complete web page. A screen shot of an
+interactive plot is shown in figure \ref{interact}. Hovering over the
+display shows the cutoff value at the point nearest to the cursor.
 Clicking makes the cutoff label stick until the next click, and if
@@
 the next click. The character string can be copy-pasted into an html
-document, or better yet, using \pkg{knitr} \citep{knitr}, we can
-\texttt{cat} the results and use the option
-\texttt{results = \textquotesingle{}asis\textquotesingle{}} so that the
-interactive plot is displayed correctly. For examples of interactive
-plots and how to incorporate them into \texttt{knitr} documents, see the
-package vignette (\texttt{vignette("examples", package = "plotROC")}) or
-the webpage https://sachsmc.github.io/plotROC/. Of note is the
-\texttt{prefix} option, which allows the user to assign each plot a
-unique identifier. This is necessary to prevent conflicts with multiple
-plots in a single webpage.
+document, or better yet, incorporated directly into a dynamic document
+using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it is
+necessary to use the \texttt{cat} function on the results and use the
+chunk option \texttt{results = \textquotesingle{}asis\textquotesingle{}}
+so that the interactive plot is displayed correctly. For documents that
+contain multiple interactive plots, it is necessary to assign each plot
+a unique name using the \texttt{prefix} argument of
+\texttt{export\_interactive\_roc}. This is necessary to ensure that the
+\proglang{JavaScript} code manipulates the correct svg elements. For
+examples of interactive plots and how to incorporate them into
+\texttt{knitr} documents, see the package vignette
+(\texttt{vignette("examples", package = "plotROC")}) or the web page
+https://sachsmc.github.io/plotROC/.
 
 \begin{verbatim}
+```{r int-no, results = 'asis'}
 cat(
@@
   )
+```
 \end{verbatim}
@@
 
-\texttt{plotROC} uses the \texttt{ggplot2} package to create ggplot
-objects. Therefore, themes and annotations can be added to
-\texttt{ggroc} objects in the usual \texttt{ggplot} way. A
-\texttt{plotROC} figure with a new theme, title, axis label, and AUC
-annotation is shown in figure \ref{annotate}.
+\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be
+plotted. Therefore, themes and annotations can be added in the usual
+\pkg{ggplot2} way. A \texttt{plot\_journal\_roc} figure with a new
+theme, title, axis label, and AUC annotation is shown in figure
+\ref{annotate}.
 
@@
            label = "AUC = 0.80") + 
-  ylab("Sensitivity")
+  scale_y_continuous("Sensitivity") + 
+  scale_x_continuous("1 - Specificity")
 \end{verbatim}

### 2015-01-30 Paper/sachs2014plotROC.Rmd
Editing and code changed to reflect improvements to package

@@
 Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers.
-We reviewed a sample of ROC curve plots from the major oncology journals in order to assess current trends in usage and design elements.
-Our review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display.
-We describe our new \proglang{R} package that was created to address the shortcomings of existing tools.
+I reviewed a sample of ROC curve plots from the major oncology journals in order to assess current trends in usage and design elements.
+My review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display.
+I describe my new \proglang{R} package that was created to address the shortcomings of existing tools.
 The package has functions to create informative ROC curve plots, with sensible defaults, for use in print or as an interactive web-based plot.
@@
 The accuracy of a diagnostic test can be evaluated by considering two possible types of errors: false positives, and false negatives.
-For a continuous measurement that we denote as $M$, convention dictates that a test positive is defined as $M$ exceeding some fixed threshold $c$: $M > c$.
-In reference to the binary outcome that we denote as $D$, a good outcome of the test is when the test is positive for an individual who truly has a disease: $D = 1$.
+For a continuous measurement that I denote as $M$, convention dictates that a test positive is defined as $M$ exceeding some fixed threshold $c$: $M > c$.
+In reference to the binary outcome that I denote as $D$, a good outcome of the test is when the test is positive for an individual who truly has a disease: $D = 1$.
 A bad outcome is when the test is positive for an individual who does not have the disease $D = 0$.
@@
 
-Since the cutoff $c$ is not fixed in advance, we can plot the TPF against the FPF for all possible values of $c$.
+Since the cutoff $c$ is not fixed in advance, I can plot the TPF against the FPF for all possible values of $c$.
 This is exactly what the ROC curve is, a plot of $FPF(c)$ on the $x$ axis and $TPF(c)$ along the $y$ axis as $c$ varies.
@@
 
-Given a sample of test and disease status pairs, $(M_1, D_1), \ldots, (M_n, D_n)$, we can estimate the ROC curve by computing proportions in the diseased and healthy subgroups separately.
+Given a sample of test and disease status pairs, $(M_1, D_1), \ldots, (M_n, D_n)$, I can estimate the ROC curve by computing proportions in the diseased and healthy subgroups separately.
 Specifically, given a fixed cutoff $c$, an estimate of the $TPF(c)$ is
@@
 There are also extensions that allow for estimation with time-to-event outcomes subject to censoring.
-For a more thorough reference on the methods and theory surrounding ROC curves, we refer interested readers to @pepe2003statistical.
+For a more thorough reference on the methods and theory surrounding ROC curves, I refer interested readers to @pepe2003statistical.
 
@@
 Sometimes, the FPF and TPF are incorrectly referred to as rates, using the abbreviations FPR and TPR.
-These are probabilities and their estimates are proportions, therefore we prefer the use of the term fraction as opposed to rate.
+These are probabilities and their estimates are proportions, therefore I prefer the use of the term fraction as opposed to rate.
 
@@
 Viewers can identify ROC plots through context, by observing the shape of the line, and through the addition of axis labels, titles, legends, and so on.
-In our review of the oncology literature, we observed plots with the distinctive characteristics of the plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions [@arr].
+In my review of the oncology literature, I observed plots with the distinctive characteristics of the plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions [@arr].
 
@@
 The \pkg{ROCR} package [@rocr] plots the FPF _versus_ TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself.
-A legend for the color scale is place along the vertical axis on the right of the plotting region.
+A legend for the color scale is placed along the vertical axis on the right of the plotting region.
 The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve.
@@
 Anyone giving a cursory look at any of the major medical journals is likely to find at least one ROC curve plot.
-We sought to assess the usage of ROC curve plots and to evaluate the design choices made in the current oncology literature by conducting a small literature review.
-We searched Pubmed for  clinical trials or observational studies in humans reported in major oncology journals for the past 10 years for the terms "ROC Curve" OR "ROC Analysis" OR "Receiver operating characteristic curve".
+I sought to assess the usage of ROC curve plots and to evaluate the design choices made in the current oncology literature by conducting a small literature review.
+I searched Pubmed for  clinical trials or observational studies in humans reported in major oncology journals for the past 10 years for the terms "ROC Curve" OR "ROC Analysis" OR "Receiver operating characteristic curve".
 The search was conducted on October 8, 2014 and returned 54 papers.
@@
 The small minority of the figures make any attempt to indicate the values of the test cutoffs, which is an integral component of the ROC curve.
-We conjecture that this is mainly due to the use of default plotting procedures in statistical software.
+I conjecture that this is mainly due to the use of default plotting procedures in statistical software.
 The software, by default, treats the ROC curve as a 2 dimensional object, obscuring the cutoff dimension.
@@
 Many of the plots included large tables containing estimates and inference for AUCs, while the ROC curves themselves, numerous and without clear labels or reference lines, merely served as decoration.
-We aim to solve some of these problems by providing an easy-to-use plotting interface for the ROC curve that provides sensible defaults.
+I aim to solve some of these problems by providing an easy-to-use plotting interface for the ROC curve that provides sensible defaults.
 
 The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements.
-We favor the use of gridlines and a diagonal reference line to facilitate accurate readings off of the axes.
+I favor the use of gridlines and a diagonal reference line to facilitate accurate readings off of the axes.
 Direct labels are preferred over legends because they omit the additional cognitive step of matching line types or colors to labels.
-Our \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use.
+My \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use.
 Exact confidence regions for points on the ROC curve are optionally calculated and displayed.
-Additionally, we use axis scales are adjusted to be denser near the margins 0 and 1.
+Additionally, I use axis scales adjusted to be denser near the margins 0 and 1.
 In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins.
@@
 
-We created a \pkg{shiny} application [@shiny] in order to make the features more accessible to non-\proglang{R} users.
+I created a \pkg{shiny} application [@shiny] in order to make the features more accessible to non-\proglang{R} users.
 A limited subset of the functions of the \pkg{plotROC} can be performed on an example dataset or on data that users upload to the website.
@@
 
-We start by creating an example data set. The marker we generate is moderately accurate for predicting disease status.
+I start by creating an example data set. The marker I generate is moderately accurate for predicting disease status.
 
@@
 
-Next we use the `calculate_roc` function to compute the empirical ROC curve.
+Next I use the `calculate_roc` function to compute the empirical ROC curve.
 The disease status need not be coded as 0/1, but if it is not, `calculate_roc` assumes (with a warning) that the lowest value in sort order signifies disease-free status.
@@
 ```{r calc}
-rocdata <- calculate_roc(M.ex, D.ex)
-str(rocdata)
+roc.estimate <- calculate_roc(M.ex, D.ex)
+str(roc.estimate)
 ```
 
-The same data.frame `rocdata` can be used to generate an interactive plot to view in Rstudio viewer or web browser by passing it to the `plot_interactive_roc` function.
+The `rocdata` is passed to the `ggroc` function with an optional label. This creates a ggplot object of the ROC curve using the \pkg{ggplot2} package [@ggplot2].
+
+```{r test-a}
+single.rocplot <- ggroc(roc.estimate, label = "Example")
+```
+
+The `myrocplot` object can be used to create an interactive plot and display it in the Rstudio viewer or default web browser by passing it to the `plot_interactive_roc` function.
 Give the function an optional path to an html file as an argument called `file` to save the interactive plot as a complete web page.
@@
 ```{r inter, eval = FALSE}
-plot_interactive_roc(rocdata)
+plot_interactive_roc(single.rocplot)
 ```
@@
 
-The `rocdata` is passed to the `ggroc` function with an optional label. This creates a ggplot object of the ROC curve using the \pkg{ggplot2} package [@ggplot2].
-
-```{r test-a}
-myrocplot <- ggroc(rocdata, label = "Example")
-```
-
-We can create an interactive ROC plot using the `export_interactive_roc` function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}.
-The key interactive feature of these plots is that the cutoff values nearest to the mouse cursor are displayed when hovering over the figure.
-Clicking makes the cutoff label stick until the next click.
+An interactive ROC plot can be exported by using the `export_interactive_roc` function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}.
 The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr].
@@
 This is necessary to ensure that the \proglang{JavaScript} code manipulates the correct svg elements.
-For examples of interactive plots and how to incorporate them into `knitr` documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the web page https://sachsmc.github.io/plotROC/.
+For examples of interactive plots and how to incorporate them into \pkg{knitr} documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the web page https://sachsmc.github.io/plotROC/.
+The next code block shows an example \pkg{knitr} chunk that can be used in an .Rmd document to display an interactive plot.
 
@@
 cat(
-  export_interactive_roc(myrocplot, cutoffs = rocdata$c, 
-                         font.size = "12px", prefix = "a")
+  export_interactive_roc(single.rocplot, 
+                        prefix = "a")
   )
@@
 
-The same `ggroc` object that we called `myrocplot` can be used to generate an ROC plot suitable for use in print.
+The same `ggroc` object that I called `myrocplot` can be used to generate an ROC plot suitable for use in print.
 It annotates the cutoff values and is completely in black and white.
@@
 ```{r print, fig.width = 6, fig.height = 6, fig.cap = "Illustration of ROC curve plot generated by \\pkg{plotROC} for use in print. \\label{first}"}
-plot_journal_roc(myrocplot, rocdata)
+plot_journal_roc(single.rocplot)
+```
+
+### Multiple ROC curves
+
+If you have multiple tests of different types measured on the same subjects, you can use the `calculate_multi_roc` function to compute the empirical ROC curve for each test.
+It returns a list of data frames with the estimates and cutoff values.
+Then the `multi_ggroc` function creates the appropriate type of `ggplot` object.
+Confidence regions are not supported for multiple curves at the time of writing.
+
+```{r multistart}
+D.ex <- rbinom(100, 1, .5)
+
+paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
+                       M2 = rnorm(100, mean = D.ex, sd = .4), 
+                       M3 = runif(100), D = D.ex)
+
+estimate.list <- calculate_multi_roc(paired.data, c("M1", "M2", "M3"), "D")
+```
+
+Labels can be added easily with the `label` option of `multi_ggroc`.
+The length of the label element should match the number of plotted curves.
+The `multi_ggroc` object can be passed to the `plot_journal_roc` and the `export_interactive_roc` functions, as desired.
+The resulting plot is shown in figure \ref{multi}.
+
+```{r multi2, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} plot with multiple curves. \\label{multi} "}
+multi.rocplot <- multi_ggroc(estimate.list, label = c("M1", "M2", "M3"))
+plot_journal_roc(multi.rocplot)
+```
+
+Both `plot_journal_roc` and `export_interactive_roc` support a number of options to customize the look of the plots.
+By default, multiple curves are distinguished by different line types and direct labels.
+For multiple ROC curves that are similar to one another, those defaults can make it difficult to interpret the plots.
+Therefore, we also support colors and legends.
+The x- and y-axis can be changed by passing options to `ggroc` or `multi_ggroc`.
+The next code block illustrates the available options.
+
+```{r multi3, eval = FALSE}
+colorplot <- multi_ggroc(estimate.list, 
+                         xlabel = "1 - Specificity", 
+                         ylabel = "Sensitivity")
+cat(
+  export_interactive_roc(colorplot, lty = rep(1, 3), 
+                         color = c("black", "purple", "orange"), 
+                         legend = TRUE)
+  )
 ```
@@
 
-We use the `ci = TRUE` option in `calculcate_roc` and `ggroc` to compute confidence regions for points on the ROC curve using the @clopper1934use exact method.
+I use the `ci = TRUE` option in `calculcate_roc` and `ggroc` to compute confidence regions for points on the ROC curve using the @clopper1934use exact method.
+Briefly, exact confidence intervals are calculated for the $FPF$ and $TPF$ separately, each at level $1 - \sqrt{1 - \alpha}$.
+Based on result 2.4 from @pepe2003statistical, the cross-product of these intervals yields a $100 * (1 - \alpha)$ percent rectangular confidence region for the pair.
 The significance level can be specified using the `alpha` option.
@@
 ```{r test-a-ci}
-rocdata <- calculate_roc(M.ex, D.ex, ci = TRUE, alpha = 0.05)
-myrocplot <- ggroc(rocdata, label = "Example", ci = TRUE)
+roc.ci <- calculate_roc(paired.data$M1, paired.data$D, ci = TRUE, alpha = 0.05)
+ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
 ```
@@
 cat(
-  export_interactive_roc(myrocplot, cutoffs = rocdata$c, 
-                         font.size = "12px", prefix = "aci")
+  export_interactive_roc(ci.rocplot, 
+                         prefix = "aci")
   )
@@
 
-For use in print, we pass a small vector of cutoff locations at which to display the confidence regions. This is shown in figure \ref{conf}.
+For use in print, I pass a small vector of cutoff locations at which to display the confidence regions. This is shown in figure \ref{conf}.
 
 ```{r printci, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} plot with exact confidence regions. \\label{conf}"}
-plot_journal_roc(myrocplot, rocdata, n.cuts = 10, 
+plot_journal_roc(ci.rocplot, n.cuts = 10, 
                  ci.at = c(-.5, .5, 2.1))
-```
-
-### Multiple ROC curves
-
-If you have multiple tests of different types on the same subjects, you can use the `calculate_multi_roc` function to compute the empirical ROC curve for each test.
-Then the `multi_ggroc` function creates the appropriate type of `ggplot` object.
-Confidence regions are not supported for multiple curves at the time of writing.
-
-```{r multistart}
-D.ex <- rbinom(100, 1, .5)
-
-fakedata <- data.frame(M1 = rnorm(100, mean = D.ex), 
-                       M2 = rnorm(100, mean = D.ex, sd = .4), 
-                       M3 = runif(100), D = D.ex)
-
-datalist <- calculate_multi_roc(fakedata, c("M1", "M2", "M3"), "D")
-rocplot <- multi_ggroc(datalist)
-```
-
-This multi ggroc object can be passed to the `plot_journal_roc` and the `export_interactive_roc` functions, as desired.
-
-Labels can be added easily with the `label` option. The length of the label element should match the number of plotted curves. The resulting plot is shown in figure \ref{multi}.
-
-```{r multi2, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} plot with multiple curves. \\label{multi} "}
-rocplot <- multi_ggroc(datalist, label = c("M1", "M2", "M3"))
-plot_journal_roc(rocplot, datalist)
 ```
@@
 library(ggplot2)
-plot_journal_roc(myrocplot, rocdata) + 
+plot_journal_roc(ci.rocplot, n.cuts = 10, 
+                 ci.at = c(-.5, .5, 2.1)) + 
   theme_grey() + 
@@
            label = "AUC = 0.80") +
-  scale_y_continuous("Sensitivity") + 
-  scale_x_continuous("1 - Specificity")
+  scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
 ```
@@
 By default `calculate_roc` computes the empirical ROC curve.
-There are other estimation methods out there, as we have summarized in the introduction.
+There are other estimation methods out there, as I have summarized in the introduction.
 Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned.
@@
 
-binorm_rocdata <- data.frame(c = c.ex, 
+binorm.roc <- data.frame(c = c.ex, 
                              FPF = pnorm((mu0 - c.ex)/s0), 
@@
 
-Then we can pass this data.frame to the `ggroc` function as before. The example is shown in figure \ref{binorm}.
+Then I can pass this data.frame to the `ggroc` function as before. The example is shown in figure \ref{binorm}.
 
 ```{r binormal, echo = TRUE, fig.width=6, fig.height=6, fig.cap = "Illustration of smooth binormal ROC curve. \\label{binorm}"}
-binorm_plot <- ggroc(binorm_rocdata, label = "Binormal")
-plot_journal_roc(binorm_plot, binorm_rocdata)
+binorm.plot <- ggroc(binorm.roc, label = "Binormal")
+plot_journal_roc(binorm.plot)
 ```
@@
 This function maps each element of the plot to a corresponding element of the svg markup language.
-We keep track of the names of the points and labels elements so that we can add interactivity using \pkg{d3.js} and \proglang{JavaScript}.
-The main interactive feature we wanted was to be able to display the cutoff labels at the points on the ROC curve closest to the mouse cursor.
+I keep track of the names of the points and labels elements so that I can add interactivity using \pkg{d3.js} and \proglang{JavaScript}.
+The main interactive feature I wanted was to be able to display the cutoff labels at the points on the ROC curve closest to the mouse cursor.
 
-There are many ways to solve this with \pkg{d3.js}, but we decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve.
+There are many ways to solve this with \pkg{d3.js}, but I decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve.
 The idea is that for the set of cutoff points along the ROC curve, the `d3.geom.voronoi` function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point.
@@
 This approach is similar to what is done in the \pkg{gridSVG} `grid.animate` function, which uses the svg `<animate />` tags.
-However, the available features were not sufficient for our needs, which is why we used \pkg{d3.js}.
+However, the available features were not sufficient for my needs, which is why I used \pkg{d3.js}.
 There are several other \proglang{R} packages that aim to create interactive figures.
@@
 
-Here we have illustrated the usage and described the mechanics of a new \proglang{R} package for creating ROC curve plots.
+Here I have illustrated the usage and described the mechanics of a new \proglang{R} package for creating ROC curve plots.
 The functions are easy to use, even for non-\proglang{R} users _via_ the web application, yet have sufficient flexibility to meet the needs of power users.
-Our approach to creating interactive plots differs from other interactive charting packages.
-We found that existing approaches did not meet the highly specialized needs of plotting ROC curves.
-While ROC curve plots can technically be created with even the most basic plotting tools, we find that specialized functions make the results clearer and more informative.
+My approach to creating interactive plots differs from other interactive charting packages.
+I found that existing approaches did not meet the highly specialized needs of plotting ROC curves.
+While ROC curve plots can technically be created with even the most basic plotting tools, I find that specialized functions make the results clearer and more informative.
 

### 2015-01-30 Paper/sachs2014plotROC.tex
Editing and code changed to reflect improvements to package

@@
   diagnostic test, ROC curves are used in oncology to evaluate screening,
-  diagnostic, prognostic and predictive biomarkers. We reviewed a sample
-  of ROC curve plots from the major oncology journals in order to assess
-  current trends in usage and design elements. Our review suggests that
-  ROC curve plots are often ineffective as statistical charts and that
-  poor design obscures the relevant information the chart is intended to
-  display. We describe our new \proglang{R} package that was created to
+  diagnostic, prognostic and predictive biomarkers. I reviewed a sample of
+  ROC curve plots from the major oncology journals in order to assess
+  current trends in usage and design elements. My review suggests that ROC
+  curve plots are often ineffective as statistical charts and that poor
+  design obscures the relevant information the chart is intended to
+  display. I describe my new \proglang{R} package that was created to
   address the shortcomings of existing tools. The package has functions to
@@
 possible types of errors: false positives, and false negatives. For a
-continuous measurement that we denote as \(M\), convention dictates that
+continuous measurement that I denote as \(M\), convention dictates that
 a test positive is defined as \(M\) exceeding some fixed threshold
-\(c\): \(M > c\). In reference to the binary outcome that we denote as
+\(c\): \(M > c\). In reference to the binary outcome that I denote as
 \(D\), a good outcome of the test is when the test is positive for an
@@
 
-Since the cutoff \(c\) is not fixed in advance, we can plot the TPF
+Since the cutoff \(c\) is not fixed in advance, I can plot the TPF
 against the FPF for all possible values of \(c\). This is exactly what
@@
 Given a sample of test and disease status pairs,
-\((M_1, D_1), \ldots, (M_n, D_n)\), we can estimate the ROC curve by
+\((M_1, D_1), \ldots, (M_n, D_n)\), I can estimate the ROC curve by
 computing proportions in the diseased and healthy subgroups separately.
@@
 subject to censoring. For a more thorough reference on the methods and
-theory surrounding ROC curves, we refer interested readers to
+theory surrounding ROC curves, I refer interested readers to
 \citet{pepe2003statistical}.
@@
 rates, using the abbreviations FPR and TPR. These are probabilities and
-their estimates are proportions, therefore we prefer the use of the term
+their estimates are proportions, therefore I prefer the use of the term
 fraction as opposed to rate.
@@
 ROC plots through context, by observing the shape of the line, and
-through the addition of axis labels, titles, legends, and so on. In our
-review of the oncology literature, we observed plots with the
-distinctive characteristics of the plotting functions from Microsoft
-Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R}
-plotting functions \citep{arr}.
+through the addition of axis labels, titles, legends, and so on. In my
+review of the oncology literature, I observed plots with the distinctive
+characteristics of the plotting functions from Microsoft Office,
+\proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting
+functions \citep{arr}.
 
@@
 color scale along the ROC curve itself. A legend for the color scale is
-place along the vertical axis on the right of the plotting region. The
+placed along the vertical axis on the right of the plotting region. The
 \pkg{pROC} package \citep{pROC} is mainly focused on estimating
@@
 Anyone giving a cursory look at any of the major medical journals is
-likely to find at least one ROC curve plot. We sought to assess the
-usage of ROC curve plots and to evaluate the design choices made in the
-current oncology literature by conducting a small literature review. We
+likely to find at least one ROC curve plot. I sought to assess the usage
+of ROC curve plots and to evaluate the design choices made in the
+current oncology literature by conducting a small literature review. I
 searched Pubmed for clinical trials or observational studies in humans
@@
 values of the test cutoffs, which is an integral component of the ROC
-curve. We conjecture that this is mainly due to the use of default
+curve. I conjecture that this is mainly due to the use of default
 plotting procedures in statistical software. The software, by default,
@@
 curves themselves, numerous and without clear labels or reference lines,
-merely served as decoration. We aim to solve some of these problems by
+merely served as decoration. I aim to solve some of these problems by
 providing an easy-to-use plotting interface for the ROC curve that
@@
 The panels of figure \ref{figure1} illustrate the most common styles of
-ROC curve plots, and the associated design elements. We favor the use of
+ROC curve plots, and the associated design elements. I favor the use of
 gridlines and a diagonal reference line to facilitate accurate readings
@@
 omit the additional cognitive step of matching line types or colors to
-labels. Our \pkg{plotROC} package additionally provides plotting of
+labels. My \pkg{plotROC} package additionally provides plotting of
 cutoff values, which are displayed interactively with the web-based
@@
 for points on the ROC curve are optionally calculated and displayed.
-Additionally, we use axis scales are adjusted to be denser near the
-margins 0 and 1. In medical applications, it is often necessary to have
-a very low FPR (less than 10\%, for instance), therefore the smaller
-scales are useful for accurately determining values near the margins.
-The next section details the usage of the \pkg{plotROC} \proglang{R}
-package and these features.
+Additionally, I use axis scales adjusted to be denser near the margins 0
+and 1. In medical applications, it is often necessary to have a very low
+FPR (less than 10\%, for instance), therefore the smaller scales are
+useful for accurately determining values near the margins. The next
+section details the usage of the \pkg{plotROC} \proglang{R} package and
+these features.
 
@@
 
-We created a \pkg{shiny} application \citep{shiny} in order to make the
+I created a \pkg{shiny} application \citep{shiny} in order to make the
 features more accessible to non-\proglang{R} users. A limited subset of
@@
 
-We start by creating an example data set. The marker we generate is
+I start by creating an example data set. The marker I generate is
 moderately accurate for predicting disease status.
@@
 
-Next we use the \texttt{calculate\_roc} function to compute the
-empirical ROC curve. The disease status need not be coded as 0/1, but if
-it is not, \texttt{calculate\_roc} assumes (with a warning) that the
-lowest value in sort order signifies disease-free status. This returns a
+Next I use the \texttt{calculate\_roc} function to compute the empirical
+ROC curve. The disease status need not be coded as 0/1, but if it is
+not, \texttt{calculate\_roc} assumes (with a warning) that the lowest
+value in sort order signifies disease-free status. This returns a
 dataframe with three columns: the cutoff values, the TPF and the FPF.
@@
 \begin{verbatim}
-rocdata <- calculate_roc(M.ex, D.ex)
-str(rocdata)
+roc.estimate <- calculate_roc(M.ex, D.ex)
+str(roc.estimate)
 \end{verbatim}
@@
 
-The same data.frame \texttt{rocdata} can be used to generate an
-interactive plot to view in Rstudio viewer or web browser by passing it
-to the \texttt{plot\_interactive\_roc} function. Give the function an
+The \texttt{rocdata} is passed to the \texttt{ggroc} function with an
+optional label. This creates a ggplot object of the ROC curve using the
+\pkg{ggplot2} package \citep{ggplot2}.
+
+\begin{verbatim}
+single.rocplot <- ggroc(roc.estimate, label = "Example")
+\end{verbatim}
+
+The \texttt{myrocplot} object can be used to create an interactive plot
+and display it in the Rstudio viewer or default web browser by passing
+it to the \texttt{plot\_interactive\_roc} function. Give the function an
 optional path to an html file as an argument called \texttt{file} to
@@
 \begin{verbatim}
-plot_interactive_roc(rocdata)
+plot_interactive_roc(single.rocplot)
 \end{verbatim}
@@
 
-The \texttt{rocdata} is passed to the \texttt{ggroc} function with an
-optional label. This creates a ggplot object of the ROC curve using the
-\pkg{ggplot2} package \citep{ggplot2}.
-
-\begin{verbatim}
-myrocplot <- ggroc(rocdata, label = "Example")
-\end{verbatim}
-
-We can create an interactive ROC plot using the
+An interactive ROC plot can be exported by using the
 \texttt{export\_interactive\_roc} function, which returns a character
 string containing the necessary \proglang{HTML} and
-\proglang{JavaScript}. The key interactive feature of these plots is
-that the cutoff values nearest to the mouse cursor are displayed when
-hovering over the figure. Clicking makes the cutoff label stick until
-the next click. The character string can be copy-pasted into an html
-document, or better yet, incorporated directly into a dynamic document
-using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it is
-necessary to use the \texttt{cat} function on the results and use the
+\proglang{JavaScript}. The character string can be copy-pasted into an
+html document, or better yet, incorporated directly into a dynamic
+document using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it
+is necessary to use the \texttt{cat} function on the results and use the
 chunk option \texttt{results = \textquotesingle{}asis\textquotesingle{}}
@@
 examples of interactive plots and how to incorporate them into
-\texttt{knitr} documents, see the package vignette
+\pkg{knitr} documents, see the package vignette
 (\texttt{vignette("examples", package = "plotROC")}) or the web page
-https://sachsmc.github.io/plotROC/.
+https://sachsmc.github.io/plotROC/. The next code block shows an example
+\pkg{knitr} chunk that can be used in an .Rmd document to display an
+interactive plot.
 
@@
 
-The same \texttt{ggroc} object that we called \texttt{myrocplot} can be
+The same \texttt{ggroc} object that I called \texttt{myrocplot} can be
 used to generate an ROC plot suitable for use in print. It annotates the
@@
 \begin{verbatim}
-plot_journal_roc(myrocplot, rocdata)
+plot_journal_roc(single.rocplot)
 \end{verbatim}
@@
 
+\subsubsection{Multiple ROC curves}\label{multiple-roc-curves}
+
+If you have multiple tests of different types measured on the same
+subjects, you can use the \texttt{calculate\_multi\_roc} function to
+compute the empirical ROC curve for each test. It returns a list of data
+frames with the estimates and cutoff values. Then the
+\texttt{multi\_ggroc} function creates the appropriate type of
+\texttt{ggplot} object. Confidence regions are not supported for
+multiple curves at the time of writing.
+
+\begin{verbatim}
+D.ex <- rbinom(100, 1, .5)
+
+paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
+                       M2 = rnorm(100, mean = D.ex, sd = .4), 
+                       M3 = runif(100), D = D.ex)
+
+estimate.list <- calculate_multi_roc(paired.data, c("M1", "M2", "M3"), "D")
+\end{verbatim}
+
+Labels can be added easily with the \texttt{label} option of
+\texttt{multi\_ggroc}. The length of the label element should match the
+number of plotted curves. The \texttt{multi\_ggroc} object can be passed
+to the \texttt{plot\_journal\_roc} and the
+\texttt{export\_interactive\_roc} functions, as desired. The resulting
+plot is shown in figure \ref{multi}.
+
+\begin{verbatim}
+multi.rocplot <- multi_ggroc(estimate.list, label = c("M1", "M2", "M3"))
+plot_journal_roc(multi.rocplot)
+\end{verbatim}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{figure/multi2-1.pdf}
+\caption{Illustration of \pkg{plotROC} plot with multiple curves.
+\label{multi}}
+\end{figure}
+
+Both \texttt{plot\_journal\_roc} and \texttt{export\_interactive\_roc}
+support a number of options to customize the look of the plots. By
+default, multiple curves are distinguished by different line types and
+direct labels. For multiple ROC curves that are similar to one another,
+those defaults can make it difficult to interpret the plots. Therefore,
+we also support colors and legends. The x- and y-axis can be changed by
+passing options to \texttt{ggroc} or \texttt{multi\_ggroc}. The next
+code block illustrates the available options.
+
+\begin{verbatim}
+colorplot <- multi_ggroc(estimate.list, 
+                         xlabel = "1 - Specificity", 
+                         ylabel = "Sensitivity")
+cat(
+  export_interactive_roc(colorplot, lty = rep(1, 3), 
+                         color = c("black", "purple", "orange"), 
+                         legend = TRUE)
+  )
+\end{verbatim}
+
 \subsection{Advanced options}\label{advanced-options}
@@
 
-We use the \texttt{ci = TRUE} option in \texttt{calculcate\_roc} and
+I use the \texttt{ci = TRUE} option in \texttt{calculcate\_roc} and
 \texttt{ggroc} to compute confidence regions for points on the ROC curve
-using the \citet{clopper1934use} exact method. The significance level
-can be specified using the \texttt{alpha} option.
+using the \citet{clopper1934use} exact method. Briefly, exact confidence
+intervals are calculated for the \(FPF\) and \(TPF\) separately, each at
+level \(1 - \sqrt{1 - \alpha}\). Based on result 2.4 from
+\citet{pepe2003statistical}, the cross-product of these intervals yields
+a \(100 * (1 - \alpha)\) percent rectangular confidence region for the
+pair. The significance level can be specified using the \texttt{alpha}
+option.
 
 \begin{verbatim}
-rocdata <- calculate_roc(M.ex, D.ex, ci = TRUE, alpha = 0.05)
-myrocplot <- ggroc(rocdata, label = "Example", ci = TRUE)
+roc.ci <- calculate_roc(paired.data$M1, paired.data$D, ci = TRUE, alpha = 0.05)
+ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
 \end{verbatim}
@@
 
-For use in print, we pass a small vector of cutoff locations at which to
+For use in print, I pass a small vector of cutoff locations at which to
 display the confidence regions. This is shown in figure \ref{conf}.
@@
 \begin{verbatim}
-plot_journal_roc(myrocplot, rocdata, n.cuts = 10, 
+plot_journal_roc(ci.rocplot, n.cuts = 10, 
                  ci.at = c(-.5, .5, 2.1))
@@
 regions. \label{conf}}
-\end{figure}
-
-\subsubsection{Multiple ROC curves}\label{multiple-roc-curves}
-
-If you have multiple tests of different types on the same subjects, you
-can use the \texttt{calculate\_multi\_roc} function to compute the
-empirical ROC curve for each test. Then the \texttt{multi\_ggroc}
-function creates the appropriate type of \texttt{ggplot} object.
-Confidence regions are not supported for multiple curves at the time of
-writing.
-
-\begin{verbatim}
-D.ex <- rbinom(100, 1, .5)
-
-fakedata <- data.frame(M1 = rnorm(100, mean = D.ex), 
-                       M2 = rnorm(100, mean = D.ex, sd = .4), 
-                       M3 = runif(100), D = D.ex)
-
-datalist <- calculate_multi_roc(fakedata, c("M1", "M2", "M3"), "D")
-rocplot <- multi_ggroc(datalist)
-\end{verbatim}
-
-This multi ggroc object can be passed to the \texttt{plot\_journal\_roc}
-and the \texttt{export\_interactive\_roc} functions, as desired.
-
-Labels can be added easily with the \texttt{label} option. The length of
-the label element should match the number of plotted curves. The
-resulting plot is shown in figure \ref{multi}.
-
-\begin{verbatim}
-rocplot <- multi_ggroc(datalist, label = c("M1", "M2", "M3"))
-plot_journal_roc(rocplot, datalist)
-\end{verbatim}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{figure/multi2-1.pdf}
-\caption{Illustration of \pkg{plotROC} plot with multiple curves.
-\label{multi}}
 \end{figure}
@@
            label = "AUC = 0.80") +
-  scale_y_continuous("Sensitivity") + 
-  scale_x_continuous("1 - Specificity")
+  scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
 \end{verbatim}
@@
 By default \texttt{calculate\_roc} computes the empirical ROC curve.
-There are other estimation methods out there, as we have summarized in
+There are other estimation methods out there, as I have summarized in
 the introduction. Any estimation method can be used, as long as the
@@
 
-Then we can pass this data.frame to the \texttt{ggroc} function as
+Then I can pass this data.frame to the \texttt{ggroc} function as
 before. The example is shown in figure \ref{binorm}.
@@
 \begin{verbatim}
-binorm_plot <- ggroc(binorm_rocdata, label = "Binormal")
-plot_journal_roc(binorm_plot, binorm_rocdata)
+binorm.plot <- ggroc(binorm.roc, label = "Binormal")
+plot_journal_roc(binorm.plot)
 \end{verbatim}
@@
 the \texttt{grid.export} function. This function maps each element of
-the plot to a corresponding element of the svg markup language. We keep
-track of the names of the points and labels elements so that we can add
+the plot to a corresponding element of the svg markup language. I keep
+track of the names of the points and labels elements so that I can add
 interactivity using \pkg{d3.js} and \proglang{JavaScript}. The main
-interactive feature we wanted was to be able to display the cutoff
-labels at the points on the ROC curve closest to the mouse cursor.
+interactive feature I wanted was to be able to display the cutoff labels
+at the points on the ROC curve closest to the mouse cursor.
 
-There are many ways to solve this with \pkg{d3.js}, but we decided to
-use Voronoi polygons to map the cursor location to the nearest point on
-the ROC curve. The idea is that for the set of cutoff points along the
-ROC curve, the \texttt{d3.geom.voronoi} function chain computes a set of
+There are many ways to solve this with \pkg{d3.js}, but I decided to use
+Voronoi polygons to map the cursor location to the nearest point on the
+ROC curve. The idea is that for the set of cutoff points along the ROC
+curve, the \texttt{d3.geom.voronoi} function chain computes a set of
 polygons overlaying the plotting region such that the area of each
@@
 \texttt{\textless{}animate /\textgreater{}} tags. However, the available
-features were not sufficient for our needs, which is why we used
+features were not sufficient for my needs, which is why I used
 \pkg{d3.js}. There are several other \proglang{R} packages that aim to
@@
 
-Here we have illustrated the usage and described the mechanics of a new
+Here I have illustrated the usage and described the mechanics of a new
 \proglang{R} package for creating ROC curve plots. The functions are
@@
 application, yet have sufficient flexibility to meet the needs of power
-users. Our approach to creating interactive plots differs from other
-interactive charting packages. We found that existing approaches did not
+users. My approach to creating interactive plots differs from other
+interactive charting packages. I found that existing approaches did not
 meet the highly specialized needs of plotting ROC curves. While ROC
 curve plots can technically be created with even the most basic plotting
-tools, we find that specialized functions make the results clearer and
+tools, I find that specialized functions make the results clearer and
 more informative.

### 2015-02-02 Paper/sachs2014plotROC.Rmd
typo
added qtlcharts reference

@@
 I created a \pkg{shiny} application [@shiny] in order to make the features more accessible to non-\proglang{R} users.
-A limited subset of the functions of the \pkg{plotROC} can be performed on an example dataset or on data that users upload to the website.
+A limited subset of the functions of \pkg{plotROC} can be performed on an example dataset or on data that users upload to the website.
 Resulting plots can be saved to the users' machine as a pdf or as a stand-alone html file.
@@
 \pkg{ggvis} [@ggvis], \pkg{rCharts} [@rcharts], and the more recently released \pkg{htmlwidgets} [@htmlwidgets] all leverage existing charting libraries written in \proglang{JavaScript}.
-Their general approach is to manipulate the data and create options in \proglang{R}, and then let the charting libraries handle the rendering and interactivity.
+\pkg{qtlcharts} [@qtlcharts] uses a set of custom \proglang{JavaScript} and \pkg{d3.js} functions to visualize data from genetic experiments.
+Their general approach is to manipulate the data and create options in \proglang{R}, and then let the charting libraries or functions handle the rendering and interactivity.
 \pkg{plotROC} lets \proglang{R} do the rendering, allowing the figures to be consistent across print and web-based media, and retaining the distinctive \proglang{R} style.
+This also allows users to manipulate the figures directly in \proglang{R} to suit their needs, using tools that are more accessible and familiar to most \proglang{R} users.
 
@@
 
+## Reproducibility Note
+
+This manuscript is completely reproducible using the source files available at http://github.com/sachsmc/plotROC-paper.
+The output below indicates the \proglang{R} packages and versions used.
+Compiling the pdf output also requires `pandoc` version 1.13.1 and `pdflatex`.
+
+```{r sesh}
+sessionInfo()
+```
+
 Bibliography 

### 2015-02-02 Paper/sachs2014plotROC.tex
typo
added qtlcharts reference

@@
 features more accessible to non-\proglang{R} users. A limited subset of
-the functions of the \pkg{plotROC} can be performed on an example
-dataset or on data that users upload to the website. Resulting plots can
-be saved to the users' machine as a pdf or as a stand-alone html file.
-It can be used in any modern web browser with no other dependencies at
-the website here: http://sachsmc.shinyapps.io/plotROC.
+the functions of \pkg{plotROC} can be performed on an example dataset or
+on data that users upload to the website. Resulting plots can be saved
+to the users' machine as a pdf or as a stand-alone html file. It can be
+used in any modern web browser with no other dependencies at the website
+here: http://sachsmc.shinyapps.io/plotROC.
 
@@
 \pkg{htmlwidgets} \citep{htmlwidgets} all leverage existing charting
-libraries written in \proglang{JavaScript}. Their general approach is to
-manipulate the data and create options in \proglang{R}, and then let the
-charting libraries handle the rendering and interactivity. \pkg{plotROC}
-lets \proglang{R} do the rendering, allowing the figures to be
-consistent across print and web-based media, and retaining the
-distinctive \proglang{R} style.
+libraries written in \proglang{JavaScript}. \pkg{qtlcharts}
+\citep{qtlcharts} uses a set of custom \proglang{JavaScript} and
+\pkg{d3.js} functions to visualize data from genetic experiments. Their
+general approach is to manipulate the data and create options in
+\proglang{R}, and then let the charting libraries or functions handle
+the rendering and interactivity. \pkg{plotROC} lets \proglang{R} do the
+rendering, allowing the figures to be consistent across print and
+web-based media, and retaining the distinctive \proglang{R} style. This
+also allows users to manipulate the figures directly in \proglang{R} to
+suit their needs, using tools that are more accessible and familiar to
+most \proglang{R} users.
 
@@
 
+\subsection{Reproducibility Note}\label{reproducibility-note}
+
+This manuscript is completely reproducible using the source files
+available at http://github.com/sachsmc/plotROC-paper. The output below
+indicates the \proglang{R} packages and versions used. Compiling the pdf
+output also requires \texttt{pandoc} version 1.13.1 and
+\texttt{pdflatex}.
+
+\begin{verbatim}
+sessionInfo()
+\end{verbatim}
+
+\begin{verbatim}
+R version 3.1.2 (2014-10-31)
+Platform: x86_64-apple-darwin13.4.0 (64-bit)
+
+locale:
+[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8
+
+attached base packages:
+[1] methods   stats     graphics  grDevices utils     datasets  base
+
+other attached packages:
+[1] DiagrammeR_0.3   ggplot2_1.0.0    tikzDevice_0.7.0 filehash_2.2-2  
+[5] plotROC_1.3.2    xtable_1.7-4     stringr_0.6.2    knitr_1.8
+
+loaded via a namespace (and not attached):
+ [1] colorspace_1.2-4  digest_0.6.8      evaluate_0.5.5   
+ [4] formatR_1.0.3     grid_3.1.2        gtable_0.1.2     
+ [7] htmltools_0.2.6   htmlwidgets_0.3.2 MASS_7.3-35      
+[10] munsell_0.4.2     plyr_1.8.1        proto_0.3-10     
+[13] Rcpp_0.11.3       reshape2_1.4.1    RJSONIO_1.3-0    
+[16] rstudioapi_0.2    scales_0.2.4      tools_3.1.2      
+[19] yaml_2.1.13      
+\end{verbatim}
+
 %\bibliographystyle{jss}

### 2015-02-02 Paper/sachs2015plotROC.Rmd
Submitted to JSS

--- 2015-02-02 Paper/sachs2014plotROC.Rmd
@@
 The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr].
-In a \pkg{knitr} document, it is necessary to use the `cat` function on the results and use the chunk option `results = 'asis'` so that the interactive plot is displayed correctly.
+In a \pkg{knitr} document, it is necessary to use the `cat` function on the results and use the chunk options `results = 'asis'` and `fig.keep = 'none'` so that the interactive plot is displayed correctly.
 For documents that contain multiple interactive plots, it is necessary to assign each plot a unique name using the `prefix` argument of `export_interactive_roc`.
@@
 
-```{r int-no, results = 'asis', eval = FALSE, verbatim = TRUE}
+```{r int-no, results = 'asis', fig.keep = 'none', eval = FALSE, verbatim = TRUE}
 cat(
@@
 
-This manuscript is completely reproducible using the source files available at http://github.com/sachsmc/plotROC-paper.
+This manuscript is completely reproducible using the source files.
 The output below indicates the \proglang{R} packages and versions used.

### 2015-02-02 Paper/sachs2015plotROC.tex
Submitted to JSS

--- 2015-02-02 Paper/sachs2014plotROC.tex
@@
 is necessary to use the \texttt{cat} function on the results and use the
-chunk option \texttt{results = \textquotesingle{}asis\textquotesingle{}}
-so that the interactive plot is displayed correctly. For documents that
-contain multiple interactive plots, it is necessary to assign each plot
-a unique name using the \texttt{prefix} argument of
+chunk options
+\texttt{results = \textquotesingle{}asis\textquotesingle{}} and
+\texttt{fig.keep = \textquotesingle{}none\textquotesingle{}} so that the
+interactive plot is displayed correctly. For documents that contain
+multiple interactive plots, it is necessary to assign each plot a unique
+name using the \texttt{prefix} argument of
 \texttt{export\_interactive\_roc}. This is necessary to ensure that the
@@
 \begin{verbatim}
-```{r int-no, results = 'asis'}
+```{r int-no, results = 'asis', fig.keep = 'none'}
 cat(
@@
 
-This manuscript is completely reproducible using the source files
-available at http://github.com/sachsmc/plotROC-paper. The output below
-indicates the \proglang{R} packages and versions used. Compiling the pdf
-output also requires \texttt{pandoc} version 1.13.1 and
-\texttt{pdflatex}.
+This manuscript is completely reproducible using the source files. The
+output below indicates the \proglang{R} packages and versions used.
+Compiling the pdf output also requires \texttt{pandoc} version 1.13.1
+and \texttt{pdflatex}.
 

### 2015-03-13 Paper/sachs2015plotROC.Rmd
response to first editorial review

@@
 The package has functions to create informative ROC curve plots, with sensible defaults, for use in print or as an interactive web-based plot.
-A web application was developed to reach a broader audience of scientists who do not use R.'
+A web application was developed to reach a broader audience of scientists who do not use \proglang{R}.'
 keywords: ROC curves; graphics; interactive; plots
@@
 
-```{r setup, echo = FALSE}
+```{r setup, echo = FALSE, include = FALSE}
 library(knitr)
 library(stringr)
+
 knit_hooks$set(plot = function(x, options) {
@@
 })
-opts_chunk$set(dev="tikz", comment = NA)
-knit_hooks$set(source = function(x, options){
-  if (!is.null(options$verbatim) && options$verbatim){
-    opts = gsub(",\\s*verbatim\\s*=\\s*TRUE\\s*", "", options$params.src)
-    opts = gsub(",\\s*eval\\s*=\\s*FALSE\\s*", "", opts)
-    bef = sprintf('\n\n    ```{r %s}\n', opts, "\n")
-    stringr::str_c(
-      bef, 
-      knitr:::indent_block(paste(x, collapse = '\n'), "    "), 
-      "\n    ```\n"
-    )
-  } else {
-    stringr::str_c("\n\n```", tolower(options$engine), "\n", 
-      paste(x, collapse = '\n'), "\n```\n\n"
-    )
-  }
-})
+
+options(prompt = "R> ")
+
+opts_chunk$set(dev="tikz", prompt = TRUE)
+
+render_sweave()
+
 ```
@@
 
-## About ROC Curves
+## About ROC curves
 
@@
 
-## Design Considerations
+## Design considerations
 
@@
 
-## Existing Plotting Software
+## Existing plotting software
 
@@
 Viewers can identify ROC plots through context, by observing the shape of the line, and through the addition of axis labels, titles, legends, and so on.
-In my review of the oncology literature, I observed plots with the distinctive characteristics of the plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions [@arr].
+In my review of the oncology literature, I observed plots with the distinctive characteristics of the plotting functions from Microsoft Excel [@excel], \proglang{SAS} [@sas], \proglang{SPSS} [@spss], and the base \proglang{R} plotting functions [@arr].
 
@@
 A legend for the color scale is placed along the vertical axis on the right of the plotting region.
+Due to the popularity of \pkg{ROCR}, my package supports the performance objects as returned by \pkg{ROCR}.
 The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve.
@@
 Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of a diagonal guide line, whether any cutpoints were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the AUC.
-The numerical results of the survey are summarized in table \ref{table1}.
+The numerical results of the survey are summarized in Table \ref{table1}.
 
@@
     if(tab) groupsa <- paste0("\t", groups) else groupsa <- groups
-    blah <- data.frame(lab = groupsa, `percent (count)`  = NA, check.names = FALSE, stringsAsFactors = FALSE)
+    blah <- data.frame(lab = groupsa, `Percent (count)`  = NA, check.names = FALSE, stringsAsFactors = FALSE)
     N <- sum(!is.na(var))
@@
 
-The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements.
+The panels of Figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements.
 I favor the use of gridlines and a diagonal reference line to facilitate accurate readings off of the axes.
@@
 AUCs are also given in the legend.
-Panels B and C add gridlines, diagonal reference lines, and direct labels.
+Panels C and D add gridlines, diagonal reference lines, and direct labels.
 \\label{figure1}"}
@@
 
-# Usage of the Package
+# Usage of the package
 
@@
 
-I start by creating an example data set. The marker I generate is moderately accurate for predicting disease status.
+I start by creating an example data set.
+The marker I generate is moderately accurate for predicting disease status.
+Next I use the \code{calculate_roc} function to compute the empirical ROC curve.
+The disease status need not be coded as 0/1, but if it is not, \code{calculate_roc} assumes (with a warning) that the lowest value in sort order signifies disease-free status.
+This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.
 
@@
 M.ex <- rnorm(100, mean = D.ex)
+roc.estimate <- calculate_roc(M.ex, D.ex)
+head(roc.estimate)
 ```
 
-Next I use the `calculate_roc` function to compute the empirical ROC curve.
-The disease status need not be coded as 0/1, but if it is not, `calculate_roc` assumes (with a warning) that the lowest value in sort order signifies disease-free status.
-This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.
-
-```{r calc}
-roc.estimate <- calculate_roc(M.ex, D.ex)
-str(roc.estimate)
-```
-
-The `rocdata` is passed to the `ggroc` function with an optional label. This creates a ggplot object of the ROC curve using the \pkg{ggplot2} package [@ggplot2].
+The \code{roc.estimate} object is passed to the \code{ggroc} function with an optional label. This creates a \code{ggplot} object of the ROC curve using the \pkg{ggplot2} package [@ggplot2].
 
@@
 
-The `myrocplot` object can be used to create an interactive plot and display it in the Rstudio viewer or default web browser by passing it to the `plot_interactive_roc` function.
-Give the function an optional path to an html file as an argument called `file` to save the interactive plot as a complete web page.
-A screen shot of an interactive plot is shown in figure \ref{interact}.
+The \code{single.rocplot} object can be used to create an interactive plot and display it in the Rstudio viewer or default web browser by passing it to the \code{plot_interactive_roc} function.
+Give the function an optional path to an html file as an argument called \code{file} to save the interactive plot as a complete web page.
+A screen shot of an interactive plot is shown in Figure \ref{interact}.
 Hovering over the display shows the cutoff value at the point nearest to the cursor.
@@
 
-An interactive ROC plot can be exported by using the `export_interactive_roc` function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}.
+An interactive ROC plot can be exported by using the \code{export_interactive_roc} function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}.
 The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr].
-In a \pkg{knitr} document, it is necessary to use the `cat` function on the results and use the chunk options `results = 'asis'` and `fig.keep = 'none'` so that the interactive plot is displayed correctly.
-For documents that contain multiple interactive plots, it is necessary to assign each plot a unique name using the `prefix` argument of `export_interactive_roc`.
+In a \pkg{knitr} document, it is necessary to use the \code{cat} function on the results and use the chunk options \code{results = 'asis'} and \code{fig.keep = 'none'} so that the interactive plot is displayed correctly.
+For documents that contain multiple interactive plots, it is necessary to assign each plot a unique name using the \code{prefix} argument of \code{export_interactive_roc}.
 This is necessary to ensure that the \proglang{JavaScript} code manipulates the correct svg elements.
-For examples of interactive plots and how to incorporate them into \pkg{knitr} documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the web page https://sachsmc.github.io/plotROC/.
+For examples of interactive plots and how to incorporate them into \pkg{knitr} documents, see the package vignette (\code{vignette("examples", package = "plotROC")}) or the web page https://sachsmc.github.io/plotROC/.
 The next code block shows an example \pkg{knitr} chunk that can be used in an .Rmd document to display an interactive plot.
 
-```{r int-no, results = 'asis', fig.keep = 'none', eval = FALSE, verbatim = TRUE}
+\begin{Code}
+`r paste("\x60\x60\x60{r int-no, results = 'asis', fig.keep = 'none'}")`
 cat(
@@
   )
-```
+`r paste("\x60\x60\x60")`
+\end{Code}
 
-The same `ggroc` object that I called `myrocplot` can be used to generate an ROC plot suitable for use in print.
+The same \code{ggroc} object that I called \code{single.rocplot} can be used to generate an ROC plot suitable for use in print.
 It annotates the cutoff values and is completely in black and white.
-A simple example with the default options is shown in figure \ref{first}.
+A simple example with the default options is shown in Figure \ref{first}.
 
@@
 
-If you have multiple tests of different types measured on the same subjects, you can use the `calculate_multi_roc` function to compute the empirical ROC curve for each test.
+If you have multiple tests of different types measured on the same subjects, you can use the \code{calculate_multi_roc} function to compute the empirical ROC curve for each test.
 It returns a list of data frames with the estimates and cutoff values.
-Then the `multi_ggroc` function creates the appropriate type of `ggplot` object.
+Then the \code{multi_ggroc} function creates the appropriate type of \code{ggplot} object.
 Confidence regions are not supported for multiple curves at the time of writing.
@@
 
-estimate.list <- calculate_multi_roc(paired.data, c("M1", "M2", "M3"), "D")
+estimate.list <- calculate_multi_roc(paired.data, 
+                                     c("M1", "M2", "M3"), "D")
 ```
 
-Labels can be added easily with the `label` option of `multi_ggroc`.
+Labels can be added easily with the \code{label} option of \code{multi_ggroc}.
 The length of the label element should match the number of plotted curves.
-The `multi_ggroc` object can be passed to the `plot_journal_roc` and the `export_interactive_roc` functions, as desired.
-The resulting plot is shown in figure \ref{multi}.
+The \code{multi_ggroc} object can be passed to the \code{plot_journal_roc} and the \code{export_interactive_roc} functions, as desired.
+The resulting plot is shown in Figure \ref{multi}.
 
@@
 
-Both `plot_journal_roc` and `export_interactive_roc` support a number of options to customize the look of the plots.
+Both \code{plot_journal_roc} and \code{export_interactive_roc} support a number of options to customize the look of the plots.
 By default, multiple curves are distinguished by different line types and direct labels.
@@
 Therefore, we also support colors and legends.
-The x- and y-axis can be changed by passing options to `ggroc` or `multi_ggroc`.
+The x- and y-axis can be changed by passing options to \code{ggroc} or \code{multi_ggroc}.
 The next code block illustrates the available options.
@@
 
-I use the `ci = TRUE` option in `calculcate_roc` and `ggroc` to compute confidence regions for points on the ROC curve using the @clopper1934use exact method.
+I use the \code{ci = TRUE} option in \code{calculcate_roc} and \code{ggroc} to compute confidence regions for points on the ROC curve using the @clopper1934use exact method.
 Briefly, exact confidence intervals are calculated for the $FPF$ and $TPF$ separately, each at level $1 - \sqrt{1 - \alpha}$.
 Based on result 2.4 from @pepe2003statistical, the cross-product of these intervals yields a $100 * (1 - \alpha)$ percent rectangular confidence region for the pair.
-The significance level can be specified using the `alpha` option.
+The significance level can be specified using the \code{alpha} option.
 
 ```{r test-a-ci}
-roc.ci <- calculate_roc(paired.data$M1, paired.data$D, ci = TRUE, alpha = 0.05)
+roc.ci <- calculate_roc(paired.data$M1, paired.data$D, 
+                        ci = TRUE, alpha = 0.05)
 ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
@@
 
-For use in print, I pass a small vector of cutoff locations at which to display the confidence regions. This is shown in figure \ref{conf}.
+For use in print, I pass a small vector of cutoff locations at which to display the confidence regions. This is shown in Figure \ref{conf}.
 
@@
 Therefore, themes and annotations can be added in the usual \pkg{ggplot2} way.
-A `plot_journal_roc` figure with a new theme, title, axis label, and AUC annotation is shown in figure \ref{annotate}.
+A \code{plot_journal_roc} figure with a new theme, title, axis label, and AUC annotation is shown in Figure \ref{annotate}.
 
@@
 
-By default `calculate_roc` computes the empirical ROC curve.
+By default \code{calculate_roc} computes the empirical ROC curve.
 There are other estimation methods out there, as I have summarized in the introduction.
 Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned.
-Then you can simply pass those values in a data frame to the `ggroc` function.
+Then you can simply pass those values in a data frame to the \code{ggroc} function.
 For example, let us use the binormal method to create a smooth curve.
@@
 
-Then I can pass this data.frame to the `ggroc` function as before. The example is shown in figure \ref{binorm}.
+Then I can pass this \code{data.frame} to the \code{ggroc} function as before. The example is shown in Figure \ref{binorm}.
 
@@
 
+\pkg{plotROC} also supports the use of the \pkg{ROCR} estimation method. Objects as returned by the function \code{rocr::performance} can be passed directly to \code{ggroc}:
+
+```{r rocr, message = FALSE, warning = FALSE}
+library(ROCR)
+rocr.est <- performance(prediction(M.ex, D.ex), "tpr", "fpr")
+rocr.plot <- ggroc(rocr.est, label = "ROCR object")
+```
+
 Another potential use of this approach is for plotting time-dependent ROC curves for time-to-event outcomes estimated as desribed in [@heagerty2000time].
+Here is an example using the \pkg{survivalROC} package [@survroc] for estimation:
 
-# How it Works
+```{r survival}
+library(survivalROC)
+survT <- rexp(350, 1/5)
+cens <- rbinom(350, 1, .1)
+M <- -8 * sqrt(survT) + rnorm(350, sd = survT)
+sroc <- lapply(c(2, 5, 10), function(t){ 
+  stroc <- survivalROC(Stime = survT, status = cens, marker = M, 
+                       predict.time = t, method = "NNE", 
+                       span = .25 * 350^(-.2))
+  data.frame(TPF = stroc[["TP"]], FPF = stroc[["FP"]], 
+             c = stroc[["cut.values"]], 
+             time = rep(stroc[["predict.time"]], 
+                        length(stroc[["FP"]])))
+  })
+
+surv.plot <- multi_ggroc(sroc, label = paste("t =", c(2, 5, 10)))
+```
+
+# How it works
 
 \pkg{plotROC} makes use of \pkg{ggplot2} [@ggplot2], \pkg{gridSVG} [@gridsvg], and \pkg{d3.js} [@bostock2011d3] to create interactive plots.
-The first step in the process is to create `ggplot` objects using the `ggroc` or `multi_ggroc` functions.
-These functions return standard `ggplot` objects that include basic styling, hidden cutoff labels, and hidden confidence regions.
+The first step in the process is to create \code{ggplot} objects using the \code{ggroc} or \code{multi_ggroc} functions.
+These functions return standard \code{ggplot} objects that include basic styling, hidden cutoff labels, and hidden confidence regions.
 They can be plotted and inspected in the \proglang{R} console.
 These form the basis for both the print versions and the interactive versions of the plots.
-Creating a print version by using the `plot_journal_roc` function simply makes visible a subset of the hidden cutoff labels and confidence regions, if available.
+Creating a print version by using the \code{plot_journal_roc} function simply makes visible a subset of the hidden cutoff labels and confidence regions, if available.
 
-\pkg{plotROC} makes interactive plots by first converting the `ggplot` object into a scalable vector graphic (svg) object with the `grid.export` function.
+\pkg{plotROC} makes interactive plots by first converting the \code{ggplot} object into a scalable vector graphic (svg) object with the \code{gridSVG::grid.export} function.
 This function maps each element of the plot to a corresponding element of the svg markup language.
@@
 There are many ways to solve this with \pkg{d3.js}, but I decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve.
-The idea is that for the set of cutoff points along the ROC curve, the `d3.geom.voronoi` function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point.
+The idea is that for the set of cutoff points along the ROC curve, the \code{d3.geom.voronoi} function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point.
 Hover events are bound to the polygons so that when the mouse cursor moves around the plotting region, the closest point on the ROC curve is made visible.
 Similarly, click events are bound to the polygons so that the appropriate confidence region is made visible upon clicking.
-The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by `export_interactive_roc`.
+The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by \code{export_interactive_roc}.
 Figure \ref{flow} outlines the \pkg{plotROC} process.
@@
 
-This approach is similar to what is done in the \pkg{gridSVG} `grid.animate` function, which uses the svg `<animate />` tags.
+This approach is similar to what is done in the \pkg{gridSVG} \code{grid.animate} function, which uses the svg \code{<animate />} tags.
 However, the available features were not sufficient for my needs, which is why I used \pkg{d3.js}.
@@
 
-## Reproducibility Note
+## Reproducibility note
 
@@
 The output below indicates the \proglang{R} packages and versions used.
-Compiling the pdf output also requires `pandoc` version 1.13.1 and `pdflatex`.
+Compiling the pdf output also requires \pkg{pandoc} version 1.13.1 and \pkg{pdflatex}.
 

### 2015-03-13 Paper/sachs2015plotROC.tex
response to first editorial review

--- 2015-01-30 Paper/sachs2014plotROC.tex
@@
-\documentclass[article]{jss}
+\documentclass[codesnippet]{jss}
 
@@
   print or as an interactive web-based plot. A web application was
-  developed to reach a broader audience of scientists who do not use R.
+  developed to reach a broader audience of scientists who do not use
+  \proglang{R}.
 }
@@
 
-\subsection{About ROC Curves}\label{about-roc-curves}
+\subsection{About ROC curves}\label{about-roc-curves}
 
@@
 
-\subsection{Design Considerations}\label{design-considerations}
+\subsection{Design considerations}\label{design-considerations}
 
@@
 
-\subsection{Existing Plotting
-Software}\label{existing-plotting-software}
+\subsection{Existing plotting
+software}\label{existing-plotting-software}
 
@@
 review of the oncology literature, I observed plots with the distinctive
-characteristics of the plotting functions from Microsoft Office,
-\proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting
-functions \citep{arr}.
+characteristics of the plotting functions from Microsoft Excel
+\citep{excel}, \proglang{SAS} \citep{sas}, \proglang{SPSS} \citep{spss},
+and the base \proglang{R} plotting functions \citep{arr}.
 
@@
 color scale along the ROC curve itself. A legend for the color scale is
-placed along the vertical axis on the right of the plotting region. The
-\pkg{pROC} package \citep{pROC} is mainly focused on estimating
-confidence intervals and regions for restricted ranges of the ROC curve.
-The plotting methods therein use the base \proglang{R} plotting
-functions to create nice displays of the curves along with shaded
-confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2}
-\citep{ggplot2} plotting library to create clear, informative ROC plots,
-with interactive features for use on the web, and sensible defaults for
-use in print.
+placed along the vertical axis on the right of the plotting region. Due
+to the popularity of \pkg{ROCR}, my package supports the performance
+objects as returned by \pkg{ROCR}. The \pkg{pROC} package \citep{pROC}
+is mainly focused on estimating confidence intervals and regions for
+restricted ranges of the ROC curve. The plotting methods therein use the
+base \proglang{R} plotting functions to create nice displays of the
+curves along with shaded confidence regions. My \pkg{plotROC} package
+uses the \pkg{ggplot2} \citep{ggplot2} plotting library to create clear,
+informative ROC plots, with interactive features for use on the web, and
+sensible defaults for use in print.
 
@@
 annotations such as the AUC. The numerical results of the survey are
-summarized in table \ref{table1}.
+summarized in Table \ref{table1}.
 
@@
   \hline
- & percent (count) \\ 
+ & Percent (count) \\ 
   \hline
@@
 
-The panels of figure \ref{figure1} illustrate the most common styles of
+The panels of Figure \ref{figure1} illustrate the most common styles of
 ROC curve plots, and the associated design elements. I favor the use of
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{figure/figure1-1.pdf}
-\caption{Illustration of design choices in plotting ROC curves. Panel A
-shows a sparse ROC curve, with no design additions inside the plotting
-region. The plot results in more white space than anything else. It is
-difficult to accurately determine values without reference lines. Panel
-B shows a plot comparing 2 curves, with different line types and a
-legend. AUCs are also given in the legend. Panels B and C add gridlines,
-diagonal reference lines, and direct labels. \label{figure1}}
+\begin{Schunk}
+\begin{figure}
+\includegraphics{figure/figure1-1} \caption[Illustration of design choices in plotting ROC curves]{Illustration of design choices in plotting ROC curves.
+Panel A shows a sparse ROC curve, with no design additions inside the plotting region.
+The plot results in more white space than anything else.
+It is difficult to accurately determine values without reference lines.
+Panel B shows a plot comparing 2 curves, with different line types and a legend.
+AUCs are also given in the legend.
+Panels C and D add gridlines, diagonal reference lines, and direct labels.
+\label{figure1}}\label{fig:figure1}
 \end{figure}
+\end{Schunk}
 
-\section{Usage of the Package}\label{usage-of-the-package}
+\section{Usage of the package}\label{usage-of-the-package}
 
@@
 
-\begin{verbatim}
-install.packages("plotROC")
-library(plotROC)
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> install.packages("plotROC")
+R> library(plotROC)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{verbatim}
-shiny_plotROC()
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> shiny_plotROC()
+\end{Sinput}
+\end{Schunk}
 
@@
 I start by creating an example data set. The marker I generate is
-moderately accurate for predicting disease status.
+moderately accurate for predicting disease status. Next I use the
+\code{calculate_roc} function to compute the empirical ROC curve. The
+disease status need not be coded as 0/1, but if it is not,
+\code{calculate_roc} assumes (with a warning) that the lowest value in
+sort order signifies disease-free status. This returns a dataframe with
+three columns: the cutoff values, the TPF and the FPF.
 
-\begin{verbatim}
-D.ex <- rbinom(100, size = 1, prob = .5)
-M.ex <- rnorm(100, mean = D.ex)
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> D.ex <- rbinom(100, size = 1, prob = .5)
+R> M.ex <- rnorm(100, mean = D.ex)
+R> roc.estimate <- calculate_roc(M.ex, D.ex)
+R> head(roc.estimate)
+\end{Sinput}
+\begin{Soutput}
+          c TPF       FPF
+1 -2.182331   1 0.9811321
+2 -1.674012   1 0.9622642
+3 -1.568181   1 0.9433962
+4 -1.550585   1 0.9245283
+5 -1.539929   1 0.9056604
+6 -1.190555   1 0.8867925
+\end{Soutput}
+\end{Schunk}
 
-Next I use the \texttt{calculate\_roc} function to compute the empirical
-ROC curve. The disease status need not be coded as 0/1, but if it is
-not, \texttt{calculate\_roc} assumes (with a warning) that the lowest
-value in sort order signifies disease-free status. This returns a
-dataframe with three columns: the cutoff values, the TPF and the FPF.
+The \code{roc.estimate} object is passed to the \code{ggroc} function
+with an optional label. This creates a \code{ggplot} object of the ROC
+curve using the \pkg{ggplot2} package \citep{ggplot2}.
 
-\begin{verbatim}
-roc.estimate <- calculate_roc(M.ex, D.ex)
-str(roc.estimate)
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> single.rocplot <- ggroc(roc.estimate, label = "Example")
+\end{Sinput}
+\end{Schunk}
 
-\begin{verbatim}
-'data.frame':   100 obs. of  3 variables:
- $ c  : num  -2.18 -1.67 -1.57 -1.55 -1.54 ...
- $ TPF: num  1 1 1 1 1 1 1 1 1 1 ...
- $ FPF: num  0.981 0.962 0.943 0.925 0.906 ...
-\end{verbatim}
+The \code{single.rocplot} object can be used to create an interactive
+plot and display it in the Rstudio viewer or default web browser by
+passing it to the \code{plot_interactive_roc} function. Give the
+function an optional path to an html file as an argument called
+\code{file} to save the interactive plot as a complete web page. A
+screen shot of an interactive plot is shown in Figure \ref{interact}.
+Hovering over the display shows the cutoff value at the point nearest to
+the cursor. Clicking makes the cutoff label stick until the next click,
+and if confidence regions are available, clicks will also display those
+as grey rectangles.
 
-The \texttt{rocdata} is passed to the \texttt{ggroc} function with an
-optional label. This creates a ggplot object of the ROC curve using the
-\pkg{ggplot2} package \citep{ggplot2}.
-
-\begin{verbatim}
-single.rocplot <- ggroc(roc.estimate, label = "Example")
-\end{verbatim}
-
-The \texttt{myrocplot} object can be used to create an interactive plot
-and display it in the Rstudio viewer or default web browser by passing
-it to the \texttt{plot\_interactive\_roc} function. Give the function an
-optional path to an html file as an argument called \texttt{file} to
-save the interactive plot as a complete web page. A screen shot of an
-interactive plot is shown in figure \ref{interact}. Hovering over the
-display shows the cutoff value at the point nearest to the cursor.
-Clicking makes the cutoff label stick until the next click, and if
-confidence regions are available, clicks will also display those as grey
-rectangles.
-
-\begin{verbatim}
-plot_interactive_roc(single.rocplot)
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> plot_interactive_roc(single.rocplot)
+\end{Sinput}
+\end{Schunk}
 
@@
 An interactive ROC plot can be exported by using the
-\texttt{export\_interactive\_roc} function, which returns a character
-string containing the necessary \proglang{HTML} and
-\proglang{JavaScript}. The character string can be copy-pasted into an
-html document, or better yet, incorporated directly into a dynamic
-document using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it
-is necessary to use the \texttt{cat} function on the results and use the
-chunk option \texttt{results = \textquotesingle{}asis\textquotesingle{}}
-so that the interactive plot is displayed correctly. For documents that
-contain multiple interactive plots, it is necessary to assign each plot
-a unique name using the \texttt{prefix} argument of
-\texttt{export\_interactive\_roc}. This is necessary to ensure that the
-\proglang{JavaScript} code manipulates the correct svg elements. For
-examples of interactive plots and how to incorporate them into
-\pkg{knitr} documents, see the package vignette
-(\texttt{vignette("examples", package = "plotROC")}) or the web page
-https://sachsmc.github.io/plotROC/. The next code block shows an example
-\pkg{knitr} chunk that can be used in an .Rmd document to display an
-interactive plot.
+\code{export_interactive_roc} function, which returns a character string
+containing the necessary \proglang{HTML} and \proglang{JavaScript}. The
+character string can be copy-pasted into an html document, or better
+yet, incorporated directly into a dynamic document using \pkg{knitr}
+\citep{knitr}. In a \pkg{knitr} document, it is necessary to use the
+\code{cat} function on the results and use the chunk options
+\code{results = 'asis'} and \code{fig.keep = 'none'} so that the
+interactive plot is displayed correctly. For documents that contain
+multiple interactive plots, it is necessary to assign each plot a unique
+name using the \code{prefix} argument of \code{export_interactive_roc}.
+This is necessary to ensure that the \proglang{JavaScript} code
+manipulates the correct svg elements. For examples of interactive plots
+and how to incorporate them into \pkg{knitr} documents, see the package
+vignette (\code{vignette("examples", package = "plotROC")}) or the web
+page https://sachsmc.github.io/plotROC/. The next code block shows an
+example \pkg{knitr} chunk that can be used in an .Rmd document to
+display an interactive plot.
 
-\begin{verbatim}
-```{r int-no, results = 'asis'}
+\begin{Code}
+```{r int-no, results = 'asis', fig.keep = 'none'}
 cat(
@@
 ```
-\end{verbatim}
+\end{Code}
 
-The same \texttt{ggroc} object that I called \texttt{myrocplot} can be
+The same \code{ggroc} object that I called \code{single.rocplot} can be
 used to generate an ROC plot suitable for use in print. It annotates the
 cutoff values and is completely in black and white. A simple example
-with the default options is shown in figure \ref{first}.
+with the default options is shown in Figure \ref{first}.
 
-\begin{verbatim}
-plot_journal_roc(single.rocplot)
-\end{verbatim}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{figure/print-1.pdf}
-\caption{Illustration of ROC curve plot generated by \pkg{plotROC} for
-use in print. \label{first}}
+\begin{Schunk}
+\begin{Sinput}
+R> plot_journal_roc(single.rocplot)
+\end{Sinput}
+\begin{figure}
+\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} for use in print. \label{first}}\label{fig:print}
 \end{figure}
+\end{Schunk}
 
@@
 If you have multiple tests of different types measured on the same
-subjects, you can use the \texttt{calculate\_multi\_roc} function to
-compute the empirical ROC curve for each test. It returns a list of data
-frames with the estimates and cutoff values. Then the
-\texttt{multi\_ggroc} function creates the appropriate type of
-\texttt{ggplot} object. Confidence regions are not supported for
-multiple curves at the time of writing.
+subjects, you can use the \code{calculate_multi_roc} function to compute
+the empirical ROC curve for each test. It returns a list of data frames
+with the estimates and cutoff values. Then the \code{multi_ggroc}
+function creates the appropriate type of \code{ggplot} object.
+Confidence regions are not supported for multiple curves at the time of
+writing.
 
-\begin{verbatim}
-D.ex <- rbinom(100, 1, .5)
+\begin{Schunk}
+\begin{Sinput}
+R> D.ex <- rbinom(100, 1, .5)
+R> 
+R> paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
++                        M2 = rnorm(100, mean = D.ex, sd = .4), 
++                        M3 = runif(100), D = D.ex)
+R> 
+R> estimate.list <- calculate_multi_roc(paired.data, 
++                                      c("M1", "M2", "M3"), "D")
+\end{Sinput}
+\end{Schunk}
 
-paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
-                       M2 = rnorm(100, mean = D.ex, sd = .4), 
-                       M3 = runif(100), D = D.ex)
+Labels can be added easily with the \code{label} option of
+\code{multi_ggroc}. The length of the label element should match the
+number of plotted curves. The \code{multi_ggroc} object can be passed to
+the \code{plot_journal_roc} and the \code{export_interactive_roc}
+functions, as desired. The resulting plot is shown in Figure
+\ref{multi}.
 
-estimate.list <- calculate_multi_roc(paired.data, c("M1", "M2", "M3"), "D")
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> multi.rocplot <- multi_ggroc(estimate.list, label = c("M1", "M2", "M3"))
+R> plot_journal_roc(multi.rocplot)
+\end{Sinput}
+\begin{figure}
+\includegraphics{figure/multi2-1} \caption{Illustration of \pkg{plotROC} plot with multiple curves. \label{multi} }\label{fig:multi2}
+\end{figure}
+\end{Schunk}
 
-Labels can be added easily with the \texttt{label} option of
-\texttt{multi\_ggroc}. The length of the label element should match the
-number of plotted curves. The \texttt{multi\_ggroc} object can be passed
-to the \texttt{plot\_journal\_roc} and the
-\texttt{export\_interactive\_roc} functions, as desired. The resulting
-plot is shown in figure \ref{multi}.
+Both \code{plot_journal_roc} and \code{export_interactive_roc} support a
+number of options to customize the look of the plots. By default,
+multiple curves are distinguished by different line types and direct
+labels. For multiple ROC curves that are similar to one another, those
+defaults can make it difficult to interpret the plots. Therefore, we
+also support colors and legends. The x- and y-axis can be changed by
+passing options to \code{ggroc} or \code{multi_ggroc}. The next code
+block illustrates the available options.
 
-\begin{verbatim}
-multi.rocplot <- multi_ggroc(estimate.list, label = c("M1", "M2", "M3"))
-plot_journal_roc(multi.rocplot)
-\end{verbatim}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{figure/multi2-1.pdf}
-\caption{Illustration of \pkg{plotROC} plot with multiple curves.
-\label{multi}}
-\end{figure}
-
-Both \texttt{plot\_journal\_roc} and \texttt{export\_interactive\_roc}
-support a number of options to customize the look of the plots. By
-default, multiple curves are distinguished by different line types and
-direct labels. For multiple ROC curves that are similar to one another,
-those defaults can make it difficult to interpret the plots. Therefore,
-we also support colors and legends. The x- and y-axis can be changed by
-passing options to \texttt{ggroc} or \texttt{multi\_ggroc}. The next
-code block illustrates the available options.
-
-\begin{verbatim}
-colorplot <- multi_ggroc(estimate.list, 
-                         xlabel = "1 - Specificity", 
-                         ylabel = "Sensitivity")
-cat(
-  export_interactive_roc(colorplot, lty = rep(1, 3), 
-                         color = c("black", "purple", "orange"), 
-                         legend = TRUE)
-  )
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> colorplot <- multi_ggroc(estimate.list, 
++                          xlabel = "1 - Specificity", 
++                          ylabel = "Sensitivity")
+R> cat(
++   export_interactive_roc(colorplot, lty = rep(1, 3), 
++                          color = c("black", "purple", "orange"), 
++                          legend = TRUE)
++   )
+\end{Sinput}
+\end{Schunk}
 
@@
 
-I use the \texttt{ci = TRUE} option in \texttt{calculcate\_roc} and
-\texttt{ggroc} to compute confidence regions for points on the ROC curve
+I use the \code{ci = TRUE} option in \code{calculcate_roc} and
+\code{ggroc} to compute confidence regions for points on the ROC curve
 using the \citet{clopper1934use} exact method. Briefly, exact confidence
@@
 a \(100 * (1 - \alpha)\) percent rectangular confidence region for the
-pair. The significance level can be specified using the \texttt{alpha}
+pair. The significance level can be specified using the \code{alpha}
 option.
 
-\begin{verbatim}
-roc.ci <- calculate_roc(paired.data$M1, paired.data$D, ci = TRUE, alpha = 0.05)
-ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> roc.ci <- calculate_roc(paired.data$M1, paired.data$D, 
++                         ci = TRUE, alpha = 0.05)
+R> ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{verbatim}
-cat(
-  export_interactive_roc(ci.rocplot, 
-                         prefix = "aci")
-  )
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> cat(
++   export_interactive_roc(ci.rocplot, 
++                          prefix = "aci")
++   )
+\end{Sinput}
+\end{Schunk}
 
 For use in print, I pass a small vector of cutoff locations at which to
-display the confidence regions. This is shown in figure \ref{conf}.
+display the confidence regions. This is shown in Figure \ref{conf}.
 
-\begin{verbatim}
-plot_journal_roc(ci.rocplot, n.cuts = 10, 
-                 ci.at = c(-.5, .5, 2.1))
-\end{verbatim}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{figure/printci-1.pdf}
-\caption{Illustration of \pkg{plotROC} plot with exact confidence
-regions. \label{conf}}
+\begin{Schunk}
+\begin{Sinput}
+R> plot_journal_roc(ci.rocplot, n.cuts = 10, 
++                  ci.at = c(-.5, .5, 2.1))
+\end{Sinput}
+\begin{figure}
+\includegraphics{figure/printci-1} \caption{Illustration of \pkg{plotROC} plot with exact confidence regions. \label{conf}}\label{fig:printci}
 \end{figure}
+\end{Schunk}
 
@@
 plotted. Therefore, themes and annotations can be added in the usual
-\pkg{ggplot2} way. A \texttt{plot\_journal\_roc} figure with a new
-theme, title, axis label, and AUC annotation is shown in figure
-\ref{annotate}.
+\pkg{ggplot2} way. A \code{plot_journal_roc} figure with a new theme,
+title, axis label, and AUC annotation is shown in Figure \ref{annotate}.
 
-\begin{verbatim}
-library(ggplot2)
-plot_journal_roc(ci.rocplot, n.cuts = 10, 
-                 ci.at = c(-.5, .5, 2.1)) + 
-  theme_grey() + 
-  geom_abline(intercept = 0, slope = 1, color = "white") + 
-  ggtitle("Themes and annotations") + 
-  annotate("text", x = .75, y = .25, 
-           label = "AUC = 0.80") +
-  scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
-\end{verbatim}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{figure/print2-1.pdf}
-\caption{Using \pkg{ggplot2} themes and annotations with \pkg{plotROC}
-objects. \label{annotate}}
+\begin{Schunk}
+\begin{Sinput}
+R> library(ggplot2)
+R> plot_journal_roc(ci.rocplot, n.cuts = 10, 
++                  ci.at = c(-.5, .5, 2.1)) + 
++   theme_grey() + 
++   geom_abline(intercept = 0, slope = 1, color = "white") + 
++   ggtitle("Themes and annotations") + 
++   annotate("text", x = .75, y = .25, 
++            label = "AUC = 0.80") +
++   scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
+\end{Sinput}
+\begin{figure}
+\includegraphics{figure/print2-1} \caption{Using \pkg{ggplot2} themes and annotations with \pkg{plotROC} objects. \label{annotate}}\label{fig:print2}
 \end{figure}
+\end{Schunk}
 
@@
 
-By default \texttt{calculate\_roc} computes the empirical ROC curve.
-There are other estimation methods out there, as I have summarized in
-the introduction. Any estimation method can be used, as long as the
-cutoff, the TPF and the FPF are returned. Then you can simply pass those
-values in a data frame to the \texttt{ggroc} function. For example, let
-us use the binormal method to create a smooth curve. This approach
-assumes that the test distribution is normal conditional on disease
-status.
+By default \code{calculate_roc} computes the empirical ROC curve. There
+are other estimation methods out there, as I have summarized in the
+introduction. Any estimation method can be used, as long as the cutoff,
+the TPF and the FPF are returned. Then you can simply pass those values
+in a data frame to the \code{ggroc} function. For example, let us use
+the binormal method to create a smooth curve. This approach assumes that
+the test distribution is normal conditional on disease status.
 
-\begin{verbatim}
-D.ex <- rbinom(100, 1, .5)
-M.ex <- rnorm(100, mean = D.ex, sd = .5)
+\begin{Schunk}
+\begin{Sinput}
+R> D.ex <- rbinom(100, 1, .5)
+R> M.ex <- rnorm(100, mean = D.ex, sd = .5)
+R> 
+R> mu1 <- mean(M.ex[D.ex == 1])
+R> mu0 <- mean(M.ex[D.ex == 0])
+R> s1 <- sd(M.ex[D.ex == 1])
+R> s0 <- sd(M.ex[D.ex == 0])
+R> c.ex <- seq(min(M.ex), max(M.ex), length.out = 300)
+R> 
+R> binorm.roc <- data.frame(c = c.ex, 
++                              FPF = pnorm((mu0 - c.ex)/s0), 
++                              TPF = pnorm((mu1 - c.ex)/s1)
++                              )
+\end{Sinput}
+\end{Schunk}
 
-mu1 <- mean(M.ex[D.ex == 1])
-mu0 <- mean(M.ex[D.ex == 0])
-s1 <- sd(M.ex[D.ex == 1])
-s0 <- sd(M.ex[D.ex == 0])
-c.ex <- seq(min(M.ex), max(M.ex), length.out = 300)
+Then I can pass this \code{data.frame} to the \code{ggroc} function as
+before. The example is shown in Figure \ref{binorm}.
 
-binorm.roc <- data.frame(c = c.ex, 
-                             FPF = pnorm((mu0 - c.ex)/s0), 
-                             TPF = pnorm((mu1 - c.ex)/s1)
-                             )
-\end{verbatim}
+\begin{Schunk}
+\begin{Sinput}
+R> binorm.plot <- ggroc(binorm.roc, label = "Binormal")
+R> plot_journal_roc(binorm.plot)
+\end{Sinput}
+\begin{figure}
+\includegraphics{figure/binormal-1} \caption[Illustration of smooth binormal ROC curve]{Illustration of smooth binormal ROC curve. \label{binorm}}\label{fig:binormal}
+\end{figure}
+\end{Schunk}
 
-Then I can pass this data.frame to the \texttt{ggroc} function as
-before. The example is shown in figure \ref{binorm}.
+\pkg{plotROC} also supports the use of the \pkg{ROCR} estimation method.
+Objects as returned by the function \code{rocr::performance} can be
+passed directly to \code{ggroc}:
 
-\begin{verbatim}
-binorm.plot <- ggroc(binorm.roc, label = "Binormal")
-plot_journal_roc(binorm.plot)
-\end{verbatim}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{figure/binormal-1.pdf}
-\caption{Illustration of smooth binormal ROC curve. \label{binorm}}
-\end{figure}
+\begin{Schunk}
+\begin{Sinput}
+R> library(ROCR)
+R> rocr.est <- performance(prediction(M.ex, D.ex), "tpr", "fpr")
+R> rocr.plot <- ggroc(rocr.est, label = "ROCR object")
+\end{Sinput}
+\end{Schunk}
 
@@
 ROC curves for time-to-event outcomes estimated as desribed in
-\citep{heagerty2000time}.
+\citep{heagerty2000time}. Here is an example using the \pkg{survivalROC}
+package \citep{survroc} for estimation:
 
-\section{How it Works}\label{how-it-works}
+\begin{Schunk}
+\begin{Sinput}
+R> library(survivalROC)
+R> survT <- rexp(350, 1/5)
+R> cens <- rbinom(350, 1, .1)
+R> M <- -8 * sqrt(survT) + rnorm(350, sd = survT)
+R> sroc <- lapply(c(2, 5, 10), function(t){ 
++   stroc <- survivalROC(Stime = survT, status = cens, marker = M, 
++                        predict.time = t, method = "NNE", 
++                        span = .25 * 350^(-.2))
++   data.frame(TPF = stroc[["TP"]], FPF = stroc[["FP"]], 
++              c = stroc[["cut.values"]], 
++              time = rep(stroc[["predict.time"]], 
++                         length(stroc[["FP"]])))
++   })
+R> 
+R> surv.plot <- multi_ggroc(sroc, label = paste("t =", c(2, 5, 10)))
+\end{Sinput}
+\end{Schunk}
+
+\section{How it works}\label{how-it-works}
 
@@
 interactive plots. The first step in the process is to create
-\texttt{ggplot} objects using the \texttt{ggroc} or
-\texttt{multi\_ggroc} functions. These functions return standard
-\texttt{ggplot} objects that include basic styling, hidden cutoff
-labels, and hidden confidence regions. They can be plotted and inspected
-in the \proglang{R} console. These form the basis for both the print
-versions and the interactive versions of the plots. Creating a print
-version by using the \texttt{plot\_journal\_roc} function simply makes
-visible a subset of the hidden cutoff labels and confidence regions, if
-available.
+\code{ggplot} objects using the \code{ggroc} or \code{multi_ggroc}
+functions. These functions return standard \code{ggplot} objects that
+include basic styling, hidden cutoff labels, and hidden confidence
+regions. They can be plotted and inspected in the \proglang{R} console.
+These form the basis for both the print versions and the interactive
+versions of the plots. Creating a print version by using the
+\code{plot_journal_roc} function simply makes visible a subset of the
+hidden cutoff labels and confidence regions, if available.
 
 \pkg{plotROC} makes interactive plots by first converting the
-\texttt{ggplot} object into a scalable vector graphic (svg) object with
-the \texttt{grid.export} function. This function maps each element of
-the plot to a corresponding element of the svg markup language. I keep
-track of the names of the points and labels elements so that I can add
-interactivity using \pkg{d3.js} and \proglang{JavaScript}. The main
-interactive feature I wanted was to be able to display the cutoff labels
-at the points on the ROC curve closest to the mouse cursor.
+\code{ggplot} object into a scalable vector graphic (svg) object with
+the \code{gridSVG::grid.export} function. This function maps each
+element of the plot to a corresponding element of the svg markup
+language. I keep track of the names of the points and labels elements so
+that I can add interactivity using \pkg{d3.js} and
+\proglang{JavaScript}. The main interactive feature I wanted was to be
+able to display the cutoff labels at the points on the ROC curve closest
+to the mouse cursor.
 
@@
 ROC curve. The idea is that for the set of cutoff points along the ROC
-curve, the \texttt{d3.geom.voronoi} function chain computes a set of
+curve, the \code{d3.geom.voronoi} function chain computes a set of
 polygons overlaying the plotting region such that the area of each
@@
 returned in the character string provided by
-\texttt{export\_interactive\_roc}. Figure \ref{flow} outlines the
+\code{export_interactive_roc}. Figure \ref{flow} outlines the
 \pkg{plotROC} process.
@@
 This approach is similar to what is done in the \pkg{gridSVG}
-\texttt{grid.animate} function, which uses the svg
-\texttt{\textless{}animate /\textgreater{}} tags. However, the available
-features were not sufficient for my needs, which is why I used
-\pkg{d3.js}. There are several other \proglang{R} packages that aim to
-create interactive figures. The authors of \pkg{animint} \citep{animint}
-created an extensive \proglang{JavaScript} library that creates plots in
-a similar way as \pkg{ggplot2}. A set of interactive features can be
-added to plots using \pkg{d3.js}. \pkg{ggvis} \citep{ggvis},
-\pkg{rCharts} \citep{rcharts}, and the more recently released
-\pkg{htmlwidgets} \citep{htmlwidgets} all leverage existing charting
-libraries written in \proglang{JavaScript}. Their general approach is to
-manipulate the data and create options in \proglang{R}, and then let the
-charting libraries handle the rendering and interactivity. \pkg{plotROC}
-lets \proglang{R} do the rendering, allowing the figures to be
-consistent across print and web-based media, and retaining the
-distinctive \proglang{R} style.
+\code{grid.animate} function, which uses the svg \code{<animate />}
+tags. However, the available features were not sufficient for my needs,
+which is why I used \pkg{d3.js}. There are several other \proglang{R}
+packages that aim to create interactive figures. The authors of
+\pkg{animint} \citep{animint} created an extensive \proglang{JavaScript}
+library that creates plots in a similar way as \pkg{ggplot2}. A set of
+interactive features can be added to plots using \pkg{d3.js}.
+\pkg{ggvis} \citep{ggvis}, \pkg{rCharts} \citep{rcharts}, and the more
+recently released \pkg{htmlwidgets} \citep{htmlwidgets} all leverage
+existing charting libraries written in \proglang{JavaScript}.
+\pkg{qtlcharts} \citep{qtlcharts} uses a set of custom
+\proglang{JavaScript} and \pkg{d3.js} functions to visualize data from
+genetic experiments. Their general approach is to manipulate the data
+and create options in \proglang{R}, and then let the charting libraries
+or functions handle the rendering and interactivity. \pkg{plotROC} lets
+\proglang{R} do the rendering, allowing the figures to be consistent
+across print and web-based media, and retaining the distinctive
+\proglang{R} style. This also allows users to manipulate the figures
+directly in \proglang{R} to suit their needs, using tools that are more
+accessible and familiar to most \proglang{R} users.
 
@@
 
+\subsection{Reproducibility note}\label{reproducibility-note}
+
+This manuscript is completely reproducible using the source files. The
+output below indicates the \proglang{R} packages and versions used.
+Compiling the pdf output also requires \pkg{pandoc} version 1.13.1 and
+\pkg{pdflatex}.
+
+\begin{Schunk}
+\begin{Sinput}
+R> sessionInfo()
+\end{Sinput}
+\begin{Soutput}
+R version 3.1.2 (2014-10-31)
+Platform: x86_64-apple-darwin13.4.0 (64-bit)
+
+locale:
+[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8
+
+attached base packages:
+[1] methods   stats     graphics  grDevices utils     datasets  base
+
+other attached packages:
+ [1] DiagrammeR_0.4    survivalROC_1.0.3 ROCR_1.0-5       
+ [4] gplots_2.16.0     ggplot2_1.0.0     tikzDevice_0.8.1 
+ [7] plotROC_1.3.3     xtable_1.7-4      stringr_0.6.2    
+[10] knitr_1.9
+
+loaded via a namespace (and not attached):
+ [1] bitops_1.0-6       caTools_1.17.1     colorspace_1.2-4  
+ [4] digest_0.6.8       evaluate_0.5.5     filehash_2.2-2    
+ [7] formatR_1.0.3      gdata_2.13.3       grid_3.1.2        
+[10] gtable_0.1.2       gtools_3.4.1       highr_0.4         
+[13] htmltools_0.2.6    htmlwidgets_0.3.2  KernSmooth_2.23-14
+[16] MASS_7.3-39        munsell_0.4.2      plyr_1.8.1        
+[19] proto_0.3-10       Rcpp_0.11.4        reshape2_1.4.1    
+[22] RJSONIO_1.3-0      rstudioapi_0.2     scales_0.2.4      
+[25] tools_3.1.2        yaml_2.1.13       
+\end{Soutput}
+\end{Schunk}
+
 %\bibliographystyle{jss}

### 2015-03-13 Paper/sachs2015plotROC.R

@@ -1,0 +1,245 @@
+## ----setup, echo = FALSE, include = FALSE--------------------------------
+library(knitr)
+library(stringr)
+
+knit_hooks$set(plot = function(x, options) {
+  if ('tikz' %in% options$dev && !options$external) {
+    hook_plot_tex(x, options)
+  } else hook_plot_md(x, options)
+})
+
+options(prompt = "R> ")
+
+opts_chunk$set(dev="tikz", prompt = TRUE)
+
+render_sweave()
+
+
+## ----review, echo = FALSE, results = 'asis'------------------------------
+library(xtable)
+
+rev <- read.csv("litreview/rocusage.csv", stringsAsFactors = FALSE)
+rev <- subset(rev, flag != "comedy")
+rev$axis.names <- gsub("fpr/tpr", "FPF/TPF", rev$axis.names, fixed = TRUE)
+
+mytable <- function(var, groups = sort(unique(var)), tab = FALSE, name = NULL) {
+
+    if(tab) groupsa <- paste0("\t", groups) else groupsa <- groups
+    blah <- data.frame(lab = groupsa, `Percent (count)`  = NA, check.names = FALSE, stringsAsFactors = FALSE)
+    N <- sum(!is.na(var))
+    for(k in 1:length(groups)){
+
+      nk <- sum(var == groups[k])
+      perc <- nk/N
+      cnt <- nk
+
+      blah[k, 2] <- paste0(format(100 * perc, digits = 1, nsmall = 1), " (", cnt, ")")
+
+    }
+
+    if(all(groups == c("no", "yes"))) {
+
+      c(name, paste0(format(100 * perc, digits = 1, nsmall = 1), " (", cnt, ")"))
+
+      } else blah
+
+  }
+
+rev$numcurves <- with(rev, ifelse(number.of.curves >= 4, "4+", as.character(number.of.curves)))
+rev$cutyesno <- with(rev, ifelse(cutoffs != "none" & cutoffs != "points", "yes", "no"))
+
+a1 <- rbind(c("Number of curves", ""), mytable(rev$numcurves, tab = TRUE))
+a2 <- c("\tAverage (SD)", paste0(round(mean(rev$number.of.curves), 1), " (", round(sd(rev$number.of.curves), 1), ")"))
+
+b1 <- rbind(c("Axis labels", ""), mytable(rev$axis.names, tab = TRUE))
[…]

### 2015-10-14 Paper/sachs2015plotROC.Rmd
definitions

@@
 I describe my new \proglang{R} package that was created to address the shortcomings of existing tools.
-The package has functions to create informative ROC curve plots, with sensible defaults, for use in print or as an interactive web-based plot.
+The package has functions to create informative ROC curve plots, with sensible defaults and a simple interface, for use in print or as an interactive web-based plot.
 A web application was developed to reach a broader audience of scientists who do not use \proglang{R}.'
@@
 In medicine, ROC curves have a long history of use for evaluating diagnostic tests in radiology and general diagnostics.
-ROC curves have also been used for decades in signal detection theory.
+ROC curves originated in the field of signal detection theory .
 
-The accuracy of a diagnostic test can be evaluated by considering two possible types of errors: false positives, and false negatives.
-For a continuous measurement that I denote as $M$, convention dictates that a test positive is defined as $M$ exceeding some fixed threshold $c$: $M > c$.
-In reference to the binary outcome that I denote as $D$, a good outcome of the test is when the test is positive for an individual who truly has a disease: $D = 1$.
-A bad outcome is when the test is positive for an individual who does not have the disease $D = 0$.
+For a continuous measurement that I denote as $M$, convention dictates that a test positive is defined as $M$ equalling or exceeding some fixed threshold $c$: $M \geq c$.
+The goal of ROC analysis is to evaluate the classification accuracy of $M$ in reference to the binary outcome $D$, which takes possible values $0$ (negative) or $1$ (positive).
+It is implicitly assumed that the subpopulation having $D = 1$ tends to have larger values of $M$ compared to the subpopulation with $D = 0$.
+The classification accuracy of $M$ can be evaluated by considering the confusion matrix (Table \ref{confus}).
+The confusion matrix cross-classifies the predicted outcome $M \geq c$ versus the true outcome $D$.
+The four cells of the matrix correspond to the possible classification outcomes: a true positive, a false positive, a true negative, and a false negative.
+ROC analysis assesses the trade-offs between the test's fraction of true positives versus the false positives.
+
+\begin{table}
+\begin{tabular}{l|l|c|c|c}
+\multicolumn{2}{c}{}&\multicolumn{2}{c}{$D$}&\\
+\cline{3-4}
+\multicolumn{2}{c|}{}&1&0&\multicolumn{1}{c}{Total}\\
+\cline{2-4}
+\multirow{2}{*}{$M$}& $M \geq c$ & true positive & false positive & test positive\\
+\cline{2-4}
+& $M < c$ & false negative & true negative & test negative\\
+\cline{2-4}
+\multicolumn{1}{c}{} & \multicolumn{1}{c}{Total} & \multicolumn{1}{c}{disease positive} & \multicolumn{1}{c}{disease negative} & \multicolumn{1}{c}{total}\\
+\end{tabular}
+\caption{\label{confus} Confusion matrix illustrating the possible classification outcomes of a continuous test $M$ with threshold $c$ versus the true outcome $D$. }
+\end{table}
 
@@
 
-$$ TPF(c) = P\{ M > c | D = 1 \} $$
+$$ TPF(c) = P\{ M \geq c | D = 1 \} $$
 
@@
 
-$$ FPF(c) = P\{ M > c | D = 0 \} $$
+$$ FPF(c) = P\{ M \geq c | D = 0 \} $$
 
-Since the cutoff $c$ is not fixed in advance, I can plot the TPF against the FPF for all possible values of $c$.
+Since the cutoff $c$ is not fixed in advance, one can plot the TPF against the FPF for all possible values of $c$.
 This is exactly what the ROC curve is, a plot of $FPF(c)$ on the $x$ axis and $TPF(c)$ along the $y$ axis as $c$ varies.
@@
 The ROC plot of a useless test is thus the diagonal line.
-A perfect test that is completely informative about disease status has $TPF(c) = 1$ and $FPF(c) = 0$ for all $c$.
+A perfect test that is completely informative about disease status has $TPF(c) = 1$ and $FPF(c) = 0$ for at least one value $c$.
+If the assumption that support of $M | D = 1$ is greater than that of $M | D = 0$ holds, then the ROC curve will lie in the upper left quadrant above the diagonal line, however this may not be the case in a particular sample.
 
@@
 
-$$ \widehat{TPF(c)} = \frac{\sum_{i = 1}^n 1\{M_i > c\} \cdot 1\{D_i = 1\}}{\sum_{i=1}^n 1\{D_i = 1\}}, $$
+$$ \widehat{TPF(c)} = \frac{\sum_{i = 1}^n 1\{M_i \geq c\} \cdot 1\{D_i = 1\}}{\sum_{i=1}^n 1\{D_i = 1\}}, $$
 

### 2015-10-14 Paper/sachs2015plotROC.Rmd
motivation

@@
 A legend for the color scale is placed along the vertical axis on the right of the plotting region.
-Due to the popularity of \pkg{ROCR}, my package supports the performance objects as returned by \pkg{ROCR}.
-The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve.
+Due to the popularity of \pkg{ROCR}, my package provides a convenience function to plot ROC curves from the performance objects as returned by \pkg{ROCR}.
+The \pkg{pROC} package [@pROC] provides an option for plotting cutoff labels (`print.thres = TREU`) and is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve.
 The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions.
@@
 
-## Motivation
+## Motivation and Literature Review
 
@@
 # C
-plot(TPF ~ FPF, data = R1, type = 's', lwd  = 2, xlab = "False positive fraction", ylab = "True positive fraction", main = "C")
+plot(TPF ~ FPF, data = R1, type = 's', lwd  = 2, xlab = "False positive fraction", ylab = "True positive fraction", main = "C", xlim = c(-.05, 1))
 abline(v = seq(0, 1, by = .2), lwd = 1, col = "grey80")
@@
 lines(TPF ~ FPF, data = R1, type = 's', lwd = 2)
-text(.2, .8, "Test A")
+text(.35, .6, "Test A")
+dexc <- seq(1, length(R1$FPF), length.out = 7)[-c(1, 7)]
+text(R1$FPF[dexc] - .075, R1$TPF[dexc] + .05, round(R1$c[dexc], 1))
+lines(TPF ~ FPF, data = R1[dexc, ], type = 'p', pch = 20)
 
@@
 ```
+
+The vast majority of the figures that were reviewed looked more like those in panels A and B than those in C and D.
+While plots like this do technically display the trade-offs between false- and true-positives, they minimize the amount of useful information that can be displayed in the space needed to plot an ROC curve.
+The plots created by \pkg{plotROC} attempt to increase the amount of information displayed in ROC charts that would otherwise be mostly white space.
+This is useful not only for print media, where space is limited, but also during data analysis.
+The analyst can quickly and easily view information that would otherwise be obscured by standard plotting software.
+The interactive features take this one step further, enhancing the plots with high density and easily accessible supplementary information.
 

### 2015-10-14 Paper/sachs2015plotROC.Rmd
Examples

@@
 A legend for the color scale is placed along the vertical axis on the right of the plotting region.
-Due to the popularity of \pkg{ROCR}, my package provides a convenience function to plot ROC curves from the performance objects as returned by \pkg{ROCR}.
 The \pkg{pROC} package [@pROC] provides an option for plotting cutoff labels (`print.thres = TREU`) and is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve.
@@
 
-I start by creating an example data set.
-The marker I generate is moderately accurate for predicting disease status.
-Next I use the \code{calculate_roc} function to compute the empirical ROC curve.
-The disease status need not be coded as 0/1, but if it is not, \code{calculate_roc} assumes (with a warning) that the lowest value in sort order signifies disease-free status.
-This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.
+I start by creating an example data set. There are 2 markers, one that is moderately predictive and one that is not as predictive.
 
@@
 library(plotROC)
-D.ex <- rbinom(100, size = 1, prob = .5)
-M.ex <- rnorm(100, mean = D.ex)
-roc.estimate <- calculate_roc(M.ex, D.ex)
-head(roc.estimate)
+set.seed(2529)
+D.ex <- rbinom(200, size = 1, prob = .5)
+M1 <- rnorm(200, mean = D.ex, sd = .65)
+M2 <- rnorm(200, mean = D.ex, sd = 1.5)
+
+test <- data.frame(D = D.ex, D.str = c("Healthy", "Ill")[D.ex + 1], 
+                   M1 = M1, M2 = M2, stringsAsFactors = FALSE)
 ```
 
-The \code{roc.estimate} object is passed to the \code{ggroc} function with an optional label. This creates a \code{ggplot} object of the ROC curve using the \pkg{ggplot2} package [@ggplot2].
+### The Roc Geom
 
-```{r test-a}
-single.rocplot <- ggroc(roc.estimate, label = "Example")
+As of version 1.0.1, the \pkg{ggplot2} package [@ggplot2] exports the previously internal functions \code{Geom} and \code{Stat}.
+This enables developers to create their own statistical tranformations and geometric layers, while enjoying all of the other features of \pkg{ggplot2}.
+We have implemented the empirical ROC curve estimate and the calculation of exact confidence regions as statistical transformations: \code{stat_roc} and \code{stat_rocci}, respectively.
+We have also defined geometric layers for the ROC curve and confidence regions for the ROC curve: \code{geom_roc} and \code{geom_rocci}, respectively.
+For further discussion and details of the grammar of graphics as implemented in \pkg{ggplot2}, we refer readers to @wickham2010layered and the \pkg{ggplot2} vignettes.
+
+To use the ROC geometric layer, I use the \code{ggplot} function to define the aesthetic mappings, and the \code{geom_roc} function to add an ROC curve layer.
+The \code{geom_roc} function requires the named aesthetics \code{d} for disease status, and \code{m} for marker.
+By default, the ROC geom and stat are linked, so that when \code{geom_roc} is called, \code{stat_roc} does the computation, and when \code{stat_roc} is called, \code{geom_roc} is used to plot the layer.
+The disease status need not be coded as 0/1, but if it is not, \code{stat_roc} assumes (with a warning) that the lowest value in sort order signifies disease-free status.
+
+```{r calc}
+basicplot <- ggplot(test, aes(d = D, m = M1)) + geom_roc()
 ```
 
-The \code{single.rocplot} object can be used to create an interactive plot and display it in the Rstudio viewer or default web browser by passing it to the \code{plot_interactive_roc} function.
+The \code{geom_roc} layer includes the ROC curve line combined with points and labels to display the values of the biomarker at the different cutpoints.
+It accepts the argument \code{n.cuts} to define the number of cutpoints to display along the curve.
+Labels can be supressed by using \code{n.cuts = 0} or \code{labels = FALSE}, however points will be displayed in the latter case.
+The size of the labels and the number of significant digits can be adjusted with \code{labelsize} and \code{labelround}, respectively.
+
+```{r opts, eval = FALSE}
+ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0)
+ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 5, labelsize = 5, labelround = 2)
+ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 50, labels = FALSE)
+```
+
+### Confidence regions and the Rocci Geom
+
+It is common to compute confidence regions for points on the ROC curve using the @clopper1934use exact method.
+Briefly, exact confidence intervals are calculated for the $FPF$ and $TPF$ separately, each at level $1 - \sqrt{1 - \alpha}$.
+Based on result 2.4 from @pepe2003statistical, the cross-product of these intervals yields a $100 * (1 - \alpha)$ percent rectangular confidence region for the pair.
+
+This is implemented in the \code{stat_rocci} and displayed as a \code{geom_rocci} layer.
+These both require the same aesthetics as the ROC geom, \code{d} for disease status and \code{m} for marker.
+By default, a set of 3 evenly spaced points along the curve are chosed to display confidence regions.
+Points corresponding to the confidence regions are distiguished from the others with a different symbol.
+You can select points by passing a vector of values in the range of \code{m} to the \code{ci.at} argument.
+By default, the significance level $\alpha$ is set to 0.05, this can be changed using the \code{sig.level} option.
+An example is shown in Figure \ref{ciex}.
+
+```{r test-a-ci, eval = 1, fig.cap = "Illustration of \\pkg{plotROC} with exact confidence regions \\label{ciex}."}
+basicplot + geom_rocci()
+
+basicplot + geom_rocci(sig.level = .01)
+ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0) +
+  geom_rocci(ci.at = quantile(M1, c(.1, .4, .5, .6, .9))) 
+```
+
+### Styles and Labels
+
+The same objects like \code{basicplot} with Roc and/or Rocci layers can be treated like any other ggplot objects.
+It can be printed to display the figure, and other layers can be added to the plot.
+We provide the function `style_roc()` which is a layer containing a theme, modified gridlines, and axes.
+Adding the `style_roc()` layer to the ggplot object creates a plot with sensible defaults for use in print.
+This function has options for the number and location of major and minor breaks, addition of the diagnoal guideline, axis labels, and any theme object created for use with ggplot2 can be supplied.
+
+The \code{direct_label} function takes a ggplot object as an argument and annotates the figure with a direct label with a automatically chosen location.
+It attempts to intellegently select an appropriate location for the label, but the location can be adjusted with \code{nudge_x, nudge_y} and \code{label.angle}.
+If the \code{labels} argument is NULL, it will take the name from the mapped aesthetic.
+A simple example with the default options is shown in Figure \ref{first}.
+
+```{r print, eval = 1, fig.width = 6, fig.height = 6, fig.cap = "Illustration of ROC curve plot generated by \\pkg{plotROC} for use in print. \\label{first}"}
+direct_label(basicplot) + style_roc()
+
+basicplot + style_roc(theme = theme_grey, xlab = "1 - Specificity")
+direct_label(basicplot, labels = "Biomarker", nudge_y = -.1) + style_roc()
+```
+
+### Interactive ROC plots
+
+The \code{basicplot} object, which is of class ggplot, can be used to create an interactive plot and display it in the Rstudio viewer or default web browser by passing it to the \code{plot_interactive_roc} function.
 Give the function an optional path to an html file as an argument called \code{file} to save the interactive plot as a complete web page.
@@
 Clicking makes the cutoff label stick until the next click, and if confidence regions are available, clicks will also display those as grey rectangles.
+By default, \code{plot_interactive_roc} removes any existing Rocci geom and adds a high-density layer of confidence regions.
+This can be suppressed by using the \code{add.cis = FALSE} option.
+The points and labels layer of the Roc geom can be hidden by using the \code{hide.points = TRUE} option.
+Then, points and labels will be displayed only when the mouse is hovering over the plotting region.
+Also by default, the \code{style_roc} function is applied, the settings can be modified by passing a call to that function.
 
 ```{r inter, eval = FALSE}
-plot_interactive_roc(single.rocplot)
+plot_interactive_roc(basicplot)
+plot_interactive_roc(basicplot, hide.points = TRUE)
+plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw()))
 ```
@@
 
+Hovering over the display shows the cutoff value at the point nearest to the cursor.
+Clicking makes the cutoff label stick until the next click, and if confidence regions are available, clicks will also display those as grey rectangles.
+The confidence regions are automatically detected.
+When the user clicks on the ROC curve, the confidence region for the TPF and FPF is overlaid using a grey rectangle.
+The label and region stick until the next click.
+
 An interactive ROC plot can be exported by using the \code{export_interactive_roc} function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}.
+The \proglang{JavaScript} source can be omitted by using the \code{omit.js = TRUE} option.
+You may wish to do this when there are multiple interactive figures in a single document; the source only needs to be included once.
 The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr].
@@
 cat(
-  export_interactive_roc(single.rocplot, 
+  export_interactive_roc(basicplot, 
                         prefix = "a")
@@
 
-The same \code{ggroc} object that I called \code{single.rocplot} can be used to generate an ROC plot suitable for use in print.
-It annotates the cutoff values and is completely in black and white.
-A simple example with the default options is shown in Figure \ref{first}.
+### Multiple ROC curves
 
-```{r print, fig.width = 6, fig.height = 6, fig.cap = "Illustration of ROC curve plot generated by \\pkg{plotROC} for use in print. \\label{first}"}
-plot_journal_roc(single.rocplot)
+If you have grouping factors in your dataset, or you have multiple markers measured on the same subjects, you may wish to plot multiple ROC curves on the same plot.
+\pkg{plotROC} fully supports faceting and grouping done by \pkg{ggplot2}.
+In out example dataset, we have 2 markers measured in a paired manner:
+
+```{r head}
+head(test)
 ```
 
-### Multiple ROC curves
+These data are in wide format, with the 2 markers going across 2 columns.
+\pkg{ggplot2} requires long format, with the marker result in a single column, and a third variable identifying the marker.
+We provide the convenience function \code{melt_roc} to perform this transformation.
+The arguments are the data frame, a name or index identifying the disease status column, and a vector of names or indices identifying the the markers.
+Optionally, the names argument gives a vector of names to assign to the marker, replacing their column names.
+The result is a data frame in long format.
 
-If you have multiple tests of different types measured on the same subjects, you can use the \code{calculate_multi_roc} function to compute the empirical ROC curve for each test.
-It returns a list of data frames with the estimates and cutoff values.
-Then the \code{multi_ggroc} function creates the appropriate type of \code{ggplot} object.
-Confidence regions are not supported for multiple curves at the time of writing.
-
-```{r multistart}
-D.ex <- rbinom(100, 1, .5)
-
-paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
-                       M2 = rnorm(100, mean = D.ex, sd = .4), 
-                       M3 = runif(100), D = D.ex)
-
-estimate.list <- calculate_multi_roc(paired.data, 
-                                     c("M1", "M2", "M3"), "D")
+```{r}
+longtest <- melt_roc(test, "D", c("M1", "M2"))
+head(longtest)
 ```
 
-Labels can be added easily with the \code{label} option of \code{multi_ggroc}.
-The length of the label element should match the number of plotted curves.
-The \code{multi_ggroc} object can be passed to the \code{plot_journal_roc} and the \code{export_interactive_roc} functions, as desired.
-The resulting plot is shown in Figure \ref{multi}.
+Then, the dataset can be passed to the \code{ggplot} function, with the marker name given as a grouping or faceting variable.
 
-```{r multi2, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} plot with multiple curves. \\label{multi} "}
-multi.rocplot <- multi_ggroc(estimate.list, label = c("M1", "M2", "M3"))
-plot_journal_roc(multi.rocplot)
+```{r group, eval = FALSE}
+ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
+ggplot(longtest, aes(d = D, m = M)) + geom_roc() + facet_wrap(~ name) + style_roc()
+ggplot(longtest, aes(d = D, m = M, linetype = name)) + geom_roc() + geom_rocci()
+ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
+pairplot <- ggplot(longtest, aes(d = D, m = M, color = name)) + 
+  geom_roc(show.legend = FALSE) + style_roc()
+direct_label(pairplot)
+
+pairplot + geom_rocci()
+pairplot + geom_rocci(linetype = 1)
 ```
 
-Both \code{plot_journal_roc} and \code{export_interactive_roc} support a number of options to customize the look of the plots.
-By default, multiple curves are distinguished by different line types and direct labels.
-For multiple ROC curves that are similar to one another, those defaults can make it difficult to interpret the plots.
-Therefore, we also support colors and legends.
-The x- and y-axis can be changed by passing options to \code{ggroc} or \code{multi_ggroc}.
-The next code block illustrates the available options.
+Showing multiple curves is also useful when there is a factor that affects the classification accuracy of the test. Let's create another example dataset.
 
-```{r multi3, eval = FALSE}
-colorplot <- multi_ggroc(estimate.list, 
-                         xlabel = "1 - Specificity", 
-                         ylabel = "Sensitivity")
-cat(
-  export_interactive_roc(colorplot, lty = rep(1, 3), 
-                         color = c("black", "purple", "orange"), 
-                         legend = TRUE)
-  )
+```{r ex2, echo = -1}
+set.seed(400)
+D.cov <- rbinom(400, 1, .5)
+gender <- c("Male", "Female")[rbinom(400, 1, .49) + 1]
+M.diff <- rnorm(400, mean = D.cov, sd = ifelse(gender == "Male", .5, 1.5))
+
+test.cov <- data.frame(D = D.cov, gender = gender, M = M.diff)
 ```
 
-## Advanced options
-
-### Click to view confidence region
-
-I use the \code{ci = TRUE} option in \code{calculcate_roc} and \code{ggroc} to compute confidence regions for points on the ROC curve using the @clopper1934use exact method.
-Briefly, exact confidence intervals are calculated for the $FPF$ and $TPF$ separately, each at level $1 - \sqrt{1 - \alpha}$.
-Based on result 2.4 from @pepe2003statistical, the cross-product of these intervals yields a $100 * (1 - \alpha)$ percent rectangular confidence region for the pair.
-The significance level can be specified using the \code{alpha} option.
-
-```{r test-a-ci}
-roc.ci <- calculate_roc(paired.data$M1, paired.data$D, 
-                        ci = TRUE, alpha = 0.05)
-ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
+```{r covplot, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} with multiple curves. \\label{multi}"}
+bygend <- ggplot(test.cov, aes(d = D, m = M, color = gender)) + 
+  geom_roc(show.legend = FALSE)
+direct_label(bygend) + style_roc()
 ```
 
-For interactive plots, the confidence regions are automatically detected.
-When the user clicks on the ROC curve, the confidence region for the TPF and FPF is overlaid using a grey rectangle.
-The label and region stick until the next click.
+Interactive versions of the plots with grouping and faceting are fully supported.
 
-```{r int-nob, eval = FALSE}
-cat(
-  export_interactive_roc(ci.rocplot, 
-                         prefix = "aci")
-  )
-```
-
-For use in print, I pass a small vector of cutoff locations at which to display the confidence regions. This is shown in Figure \ref{conf}.
-
-```{r printci, fig.width = 6, fig.height = 6, fig.cap = "Illustration of \\pkg{plotROC} plot with exact confidence regions. \\label{conf}"}
-plot_journal_roc(ci.rocplot, n.cuts = 10, 
-                 ci.at = c(-.5, .5, 2.1))
-```
+## Advanced options
 
@@
 A \code{plot_journal_roc} figure with a new theme, title, axis label, and AUC annotation is shown in Figure \ref{annotate}.
+\pkg{plotROC} provides the convenience function \code{calc_auc} that takes a ggplot object that has an Roc layer, extracts the data, and calculates the AUC.
 
 ```{r print2, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
-library(ggplot2)
-plot_journal_roc(ci.rocplot, n.cuts = 10, 
-                 ci.at = c(-.5, .5, 2.1)) + 
-  theme_grey() + 
-  geom_abline(intercept = 0, slope = 1, color = "white") + 
+basicplot + 
+  style_roc(theme = theme_grey) +
+  theme(axis.text = element_text(colour = "blue")) +
   ggtitle("Themes and annotations") + 
   annotate("text", x = .75, y = .25, 
-           label = "AUC = 0.80") +
+           label = paste("AUC =", round(calc_auc(basicplot)$AUC, 2))) +
   scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
@@
 Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned.
-Then you can simply pass those values in a data frame to the \code{ggroc} function.
+Then you can simply pass those values in a data frame to the \code{ggplot} function and override the default statistical tranformation to \code{stat_identity}.
 For example, let us use the binormal method to create a smooth curve.
@@
 ```{r binormalsetup}
-D.ex <- rbinom(100, 1, .5)
-M.ex <- rnorm(100, mean = D.ex, sd = .5)
-
+D.ex <- test$D
+M.ex <- test$M1
 mu1 <- mean(M.ex[D.ex == 1])
@@
                              )
+
+binorm.plot <- ggplot(binorm.roc, aes(x = FPF, y = TPF, label = c)) + 
+  geom_roc(stat = "identity") + style_roc(theme = theme_grey)
 ```
 
-Then I can pass this \code{data.frame} to the \code{ggroc} function as before. The example is shown in Figure \ref{binorm}.
+The example is shown in Figure \ref{binorm}. Interactive plots with \code{stat = 'identity'} are not currently supported.
 
 ```{r binormal, echo = TRUE, fig.width=6, fig.height=6, fig.cap = "Illustration of smooth binormal ROC curve. \\label{binorm}"}
-binorm.plot <- ggroc(binorm.roc, label = "Binormal")
-plot_journal_roc(binorm.plot)
-```
-
-\pkg{plotROC} also supports the use of the \pkg{ROCR} estimation method. Objects as returned by the function \code{rocr::performance} can be passed directly to \code{ggroc}:
-
-```{r rocr, message = FALSE, warning = FALSE}
-library(ROCR)
-rocr.est <- performance(prediction(M.ex, D.ex), "tpr", "fpr")
-rocr.plot <- ggroc(rocr.est, label = "ROCR object")
+binorm.plot
 ```
@@
 
-surv.plot <- multi_ggroc(sroc, label = paste("t =", c(2, 5, 10)))
+sroclong <- do.call(rbind, sroc)
+
+survplot <- ggplot(sroclong, aes(x = FPF, y = TPF, label = c, color = time)) + 
+  geom_roc(labels = FALSE, stat = "identity")
 ```
@@
 \pkg{plotROC} makes use of \pkg{ggplot2} [@ggplot2], \pkg{gridSVG} [@gridsvg], and \pkg{d3.js} [@bostock2011d3] to create interactive plots.
-The first step in the process is to create \code{ggplot} objects using the \code{ggroc} or \code{multi_ggroc} functions.
-These functions return standard \code{ggplot} objects that include basic styling, hidden cutoff labels, and hidden confidence regions.
+The first step in the process is to create \code{ggplot} objects with Roc and/or Rocci layers.
 They can be plotted and inspected in the \proglang{R} console.
 These form the basis for both the print versions and the interactive versions of the plots.
-Creating a print version by using the \code{plot_journal_roc} function simply makes visible a subset of the hidden cutoff labels and confidence regions, if available.
+Basic styling and labeling can be added with the \code{style_roc} and \code{direct_label} functions.
 
@@
 While ROC curve plots can technically be created with even the most basic plotting tools, I find that specialized functions make the results clearer and more informative.
+The functions are integrated into the existing and popular \pkg{ggplot2} package, so that all the benefits and features therein can be used effectively.
 

### 2015-10-14 Paper/response-to-review.md
compiled

# Reviewer 1

> This is a really cool package that uses state of the art tech with ggplot and shiny, as well as interactive plots, to display ROC curves.
> Unfortunately, the package is rather difficult to use and quite inconsistent. It also produces some bogus results (see below). As it is now, it looks more like a pretty cool tech demo than a scientific software I would use in my own research. Here are a few comments to try to improve it, as well as the manuscript itself.

## Bug report:
> The confidence intervals are apparently bogus. The following code executed in Rstudio or from the command line, with R 3.1.3 on Ubuntu 14.04 (from CRAN's Debian repository) and R 3.1.2 on Mac OS Yosemite (from Fink):

set.seed(42)
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
roc.estimate <- calculate_roc(round(M.ex, 2), D.ex, ci = TRUE)
plot_interactive_roc(ggroc(roc.estimate, ci = TRUE))

After clicking on a point around 1.6, I got the screenshot shown in Figure 1. This seems related to the rounding (here to the 2nd decimal).
Figure 1: the CI region doesn't span the cutoff point.

Thank you for the bug report. I believe this was due to the way the previous function handled ties in the results. The computation of the ROC curve in R and the js functions for interactivity have been completely revamped. I cannot reproduce the error using the new update in any browser that I've tested, so I believe it has been fixed.

## General:
> • The manuscript is written partially in the first person. It is surprising and not totally consistent throughout the manuscript, some of which is written in passive tense (for instance “Labels can be added easily [...]”).

The manuscript has been edited to be consistent with the use of active voice.

> • The abstract highlights the literature review as a significant finding. Surprisingly it is is only briefly mentioned in the introduction. This seems a bit inconsistent.

> • Strictly speaking one could argue the plotROC package "obscures the cutoffs", as they are not displayed by default when calling the ggroc function. This is not so different from existing packages (as reviewed in subsection 1.3) that allow accessing the thresholds and adding them to the plot in different ways. The user has to request an interactive or journal plot to get them. The abstract and introduction should be rephrased in a more neutral way on this topic, as this is not the key point addressed by this package.
## Introduction:
> • The introduction states that "ROC curves have been used for decades in signal detection theory". In fact, they originate from this field (s/n in radio receivers, hence the name)
> • The "good outcome", "bad outcome" is confusing to design correct / incorrect classifications. Especially in the medical field. I don't think it is necessary.
> • "A perfect test [...] has TPF(c) = 1 and FPF(c) = 0 for all c". This is not correct. It is sufficient that this equality holds true for a single c.
> • The formula hat(TPF(c)) is slightly confusing. It may be useful to first introduce the confusion table with true | false positive | negatives. It seems correct nonetheless.
> • "The greater the AUC, the more informative the test". This statement is incomplete at best. What matters is | AUC - 0.5 |. An AUC = 0 is also the result of a perfect test (for which M < c if I am not mistaken).
> • Existing software: pROC allows to plot cutoffs quite easily (with print.thres=TRUE, just checked now). Of course it still uses the base R plots.
> • I am surprised a Pubmed search returned only 54 papers. This part would probably deserve to be expanded a little bit, and maybe turned into a whole section.
## Usage of the package:
> • Plots such as fig 4 will be hard to read in small figures, as is often seen in journals.
> • Fig 5 is unclear. Especially, what was the cutoff used to compute the region, as several ones are displayed in the region?
> • In general, the use of shaded region instead of bars is quite surprising, although perhaps more accurate than bars.
> • Why are the regions rectangle-shaped and not oval? That would seem even more correct, although probably quite difficult to set up.
> • The gridlines in figures 3-5 & 7 are so thin and light-coloured that they did not appear when I printed this manuscript on paper. The printer was not (yet) running out of ink.
> • When reporting an AUC and confidence intervals in a publication, it seems to me at least as important to report values in addition to a plot. One can annotate a curve with the AUC (p. 13), with, for instance, label = “AUC=0.80”. But this value comes out of nowhere, and does not even look realistic (to me it looks more like AUC = 0.7). In addition, it isn't an “Advanced option”, as the subsection suggests, but rather as something quite basic. An easy way out would be to add a few additional lines computing this AUC to the manuscript. The case of CI seem a bit more difficult to deal with.

## Package:

> • The usage requires a lot of typing, and does not appear to be straightforward. For instance, the function calculate_roc must be used to create the FPF and TPF matrix. This will create a lot of headaches for infrequent users: is it calculate_, or create_, or generate_, or ...?
> • This wouldn't be an issue if users could browse the functions of the package as one can usually do by typing ?plotROC. Unfortunately, this leads to a stub that has never been authored. Please update man/plotROC-package.R with some useful pointers. Of course one can always find this paper, but it is much easier to get a proper help.
> • It is of course a design choice, but the use of underscores over dots (or camelCase) in function names seems quite surprising and may disturb some users. See Bååth, R. (2012). The State of Naming Conventions in R. R Journal, 4, 74-75. R Foundation for Statistical Computing for a review of current usage of function name styles.
> • Wouldn't it be more idiomatic to introduce the roc as a ggplot geometry? That would allow some much more straightforward syntax like:
data.for.roc <- data.frame(M.ex, D.ex)
qplot(M.ex, D.ex, data = data.for.roc) + geom_roc()
or suchlike that would be much easier to remember.

> • I don't understand why one has to pass a ggroc argument to plot_interactive_roc. That is a lot of functions to use in row and in the right order to get an interactive plot, and 3 times (_)?roc. I suspect a few well designed objects would make it much more user-friendly.
> • What do the modCss and modJs functions do? The documentation is really limited and an example would help. Same with verify_d. As it stands now these functions seem useless.
> • The help files lack examples altogether. Not only does it help a lot to have copy- pasteable code snippets as a start, but it can also help checking the package for potential regressions during updates.
> • Performance: as most R packages doing ROC curves, this one scales pretty badly with larger sample sizes. I recommend using the cumsum function to compute the FPF/TPF. This will be much faster than the current implementation with sapply (which is really just a hidden slow loop). See the ROCR package that does it correctly. The calculate_roc function with the sample data of subsection 2.4 scaled to 1E5 data points literally collapsed and took > 10 minutes to compute. Such sample sizes may be uncommon in oncology, but are not rare in genomics or computer science. ROCR returns nearly instantly with such a low (sic!) sample size and scales much higher.
> • How do I add the cut-offs to the plot? If I say ggplot(..., ci.at = 0.5) I get an unused argument error and nothing is added. Is this only available in the interactive plot?

## Shiny app:

> • The shiny application has an hard dependency on ggthemes and is unusable without that package. Unfortunately it is only “Suggest”ed, and and error message is printed only after a query to the shiny web application that doesn't work. PlotROC should either “Depend” on that package, as for shiny, or catch its absence and print the error message immediately when the shiny app is launched.
> • On my Ubuntu system, the Download button opens a 404 error page with URL as http://127.0.0.1:3048/session/61edd0aac13532dc34788b220e9d27c8/downlo ad/printDownload?w= instead of downloading the image. I can right click the image and “View image” normally. The download works properly on my Mac where a tab shortly opens before the download starts.
> • The browse button appears very low in the page and is not visible enough IMO.

# Reviewer 2

> The paper about plotROC describes a new R package for plotting receiver operating characteristic curves (ROC) curves for the evaluation of binary classifiers.

> The paper is well written and gives background on ROC curves, motivation for improving ROC curves based on a literature survey, as well as code examples on how to use plotROC.

> The greatest strengths of the new package are: 
	- Clean and nice graphics (based on ggplot2)
	- The attempt to provide sensible defaults (e.g. for fontsize, axis tickmarks, etc.)
	- Options for interactivity, i.e. providing a web-interface using shiny. This is intended to make high quality ROC analyses accessible to non-technical users

> However, the package falls short of the goal proclaimed in the title of providing a "better tool for plotting ROC curves". While the strengths above deserve credit, other established tools (e.g. the referenced packages pROC and ROCR) provide several other features that are not covered in the current package (e.g. other performance measures, AUC, partial AUC, averaging of curves, aggregating curves across bootstraps or cross-validation folds, and confidence limits and regions on many of those measures).

> Also the paper and package suffer from a number of ambiguities and inaccuracies (either technical or in the description):
	- The extreme point TPF/TPR=1/1 is not included in the output of the calculate_roc function. This could lead to problems when plotting curves that don't span the full 0/0 to 1/1 space and also when computing AUCs.
	- plotROC seems to implictly define that samples ">" the threshold are positive, samples "<=" the threshold are negative. In this it differs from e.g. Fawcett "An introduction to ROC analysis" or other software (e.g. the ROCR package) that define samples ">=" the threshold as a positive prediction.
	- It is not obvious how plotROC handles ties (i.e. samples with identical numerical measurements)
	- The paper defines "good outcome" and "bad outcome" in Section 1.1. These definitions seem to ignore half of the confusion matrix (i.e. ignoring true and false negatives).
	- In "Help and Documentation" of the webpage, the confusion matrix and the definitions seem to be copied from Wikipedia. Credits and reference are needed.
	- The website states "confidence level = 0.05". Should this be "significance level" instead?

> Overall, the paper and package provide some new noteworthy ideas to ROC analysis with the attempt to make high quality analyses more accessible to non-technical experts. The current tool does not have the maturity and comprehensiveness of other tools - therefore the users will be left with the choice between alternative tools, each with their own strengths and weaknesses.

### 2015-10-14 Paper/sachs2015plotROC.Rmd
compiled

@@
 
-opts_chunk$set(dev="tikz", prompt = TRUE)
+opts_chunk$set(dev="tikz", prompt = TRUE, sanitize = TRUE)
 
@@
 \begin{table}
-\begin{tabular}{l|l|c|c|c}
-\multicolumn{2}{c}{}&\multicolumn{2}{c}{$D$}&\\
-\cline{3-4}
-\multicolumn{2}{c|}{}&1&0&\multicolumn{1}{c}{Total}\\
-\cline{2-4}
-\multirow{2}{*}{$M$}& $M \geq c$ & true positive & false positive & test positive\\
-\cline{2-4}
-& $M < c$ & false negative & true negative & test negative\\
-\cline{2-4}
-\multicolumn{1}{c}{} & \multicolumn{1}{c}{Total} & \multicolumn{1}{c}{disease positive} & \multicolumn{1}{c}{disease negative} & \multicolumn{1}{c}{total}\\
+\begin{tabular}{l|c|c|c}
+\multicolumn{1}{c}{}&\multicolumn{2}{c}{$D$}&\\
+\cline{2-3}
+\multicolumn{1}{c|}{}&1&0&\multicolumn{1}{c}{Total}\\
+\cline{2-3}
+ $M \geq c$ & true positive & false positive & test positive\\
+\cline{2-3}
+ $M < c$ & false negative & true negative & test negative\\
+\cline{2-3}
+\multicolumn{1}{c}{Total} & \multicolumn{1}{c}{disease positive} & \multicolumn{1}{c}{disease negative} & \multicolumn{1}{c}{}\\
 \end{tabular}
@@
 A legend for the color scale is placed along the vertical axis on the right of the plotting region.
-The \pkg{pROC} package [@pROC] provides an option for plotting cutoff labels (`print.thres = TREU`) and is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve.
+The \pkg{pROC} package [@pROC] provides an option for plotting cutoff labels (\code{print.thres = TRUE}) and is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve.
 The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions.
@@
 
-```{r figure1, fig.width = 6, fig.height = 6, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves.
+```{r figure1, fig.width = 6, fig.height = 6, message = FALSE, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves.
 Panel A shows a sparse ROC curve, with no design additions inside the plotting region.
@@
 
-```{r test-a-ci, eval = 1, fig.cap = "Illustration of \\pkg{plotROC} with exact confidence regions \\label{ciex}."}
+```{r test-a-ci, eval = 1, fig.cap = "Illustration of \\pkg{plotROC} with exact confidence regions. \\label{ciex}"}
 basicplot + geom_rocci()
@@
 
-```{r print2, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
-basicplot + 
-  style_roc(theme = theme_grey) +
+```{r print2s, warning = FALSE, message = FALSE}
+annotate <- basicplot + style_roc(theme = theme_grey) +
   theme(axis.text = element_text(colour = "blue")) +
@@
   annotate("text", x = .75, y = .25, 
-           label = paste("AUC =", round(calc_auc(basicplot)$AUC, 2))) +
+           label = paste("AUC =", round(calc_auc(basicplot)["AUC"], 2))) +
   scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
 ```
+
+```{r print2, echo = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
+annotate
+```
+
+The results are compatible with all other \pkg{ggplot2} layers and functions, and interactive versions are supported as long as there is an Roc layer present.
 
@@
 The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by \code{export_interactive_roc}.
-Figure \ref{flow} outlines the \pkg{plotROC} process.
-
-```{r diagram, include = FALSE}
-library(DiagrammeR)
-
-DiagrammeR("
-  graph LR;
-    A[ROC Estimate]-->|ggroc|B(ggplot object);
-    B-->|plot_journal_roc|C(Static Plot);
-    B-->|grid.export|D(svg object);
-    D-->E(Interactive Plot);
-    F(d3.js functions)-->E;
-    G(css styles)-->E;
-
-  style A fill:#DCEBE3
-  style C fill:#A2EB86
-  style E fill:#A2EB86
-")
-
-## export as pdf
-
-```
-
-\begin{figure}[ht]
-\centering
-\includegraphics{figure/diagram.pdf}
-\caption{Flowchart illustrating the approach that \pkg{plotROC} takes to generate either static plots for print or interactive plots for web-use. \label{flow}}
-\end{figure}
 

### 2015-10-14 Paper/sachs2015plotROC.tex
compiled

@@
   address the shortcomings of existing tools. The package has functions to
-  create informative ROC curve plots, with sensible defaults, for use in
-  print or as an interactive web-based plot. A web application was
-  developed to reach a broader audience of scientists who do not use
-  \proglang{R}.
+  create informative ROC curve plots, with sensible defaults and a simple
+  interface, for use in print or as an interactive web-based plot. A web
+  application was developed to reach a broader audience of scientists who
+  do not use \proglang{R}.
 }
@@
 medicine, ROC curves have a long history of use for evaluating
-diagnostic tests in radiology and general diagnostics. ROC curves have
-also been used for decades in signal detection theory.
+diagnostic tests in radiology and general diagnostics. ROC curves
+originated in the field of signal detection theory .
 
-The accuracy of a diagnostic test can be evaluated by considering two
-possible types of errors: false positives, and false negatives. For a
-continuous measurement that I denote as \(M\), convention dictates that
-a test positive is defined as \(M\) exceeding some fixed threshold
-\(c\): \(M > c\). In reference to the binary outcome that I denote as
-\(D\), a good outcome of the test is when the test is positive for an
-individual who truly has a disease: \(D = 1\). A bad outcome is when the
-test is positive for an individual who does not have the disease
-\(D = 0\).
+For a continuous measurement that I denote as \(M\), convention dictates
+that a test positive is defined as \(M\) equalling or exceeding some
+fixed threshold \(c\): \(M \geq c\). The goal of ROC analysis is to
+evaluate the classification accuracy of \(M\) in reference to the binary
+outcome \(D\), which takes possible values \(0\) (negative) or \(1\)
+(positive). It is implicitly assumed that the subpopulation having
+\(D = 1\) tends to have larger values of \(M\) compared to the
+subpopulation with \(D = 0\). The classification accuracy of \(M\) can
+be evaluated by considering the confusion matrix (Table \ref{confus}).
+The confusion matrix cross-classifies the predicted outcome \(M \geq c\)
+versus the true outcome \(D\). The four cells of the matrix correspond
+to the possible classification outcomes: a true positive, a false
+positive, a true negative, and a false negative. ROC analysis assesses
+the trade-offs between the test's fraction of true positives versus the
+false positives.
+
+\begin{table}
+\begin{tabular}{l|c|c|c}
+\multicolumn{1}{c}{}&\multicolumn{2}{c}{$D$}&\\
+\cline{2-3}
+\multicolumn{1}{c|}{}&1&0&\multicolumn{1}{c}{Total}\\
+\cline{2-3}
+ $M \geq c$ & true positive & false positive & test positive\\
+\cline{2-3}
+ $M < c$ & false negative & true negative & test negative\\
+\cline{2-3}
+\multicolumn{1}{c}{Total} & \multicolumn{1}{c}{disease positive} & \multicolumn{1}{c}{disease negative} & \multicolumn{1}{c}{}\\
+\end{tabular}
+\caption{\label{confus} Confusion matrix illustrating the possible classification outcomes of a continuous test $M$ with threshold $c$ versus the true outcome $D$. }
+\end{table}
 
@@
 
-\[ TPF(c) = P\{ M > c | D = 1 \} \]
+\[ TPF(c) = P\{ M \geq c | D = 1 \} \]
 
@@
 
-\[ FPF(c) = P\{ M > c | D = 0 \} \]
+\[ FPF(c) = P\{ M \geq c | D = 0 \} \]
 
-Since the cutoff \(c\) is not fixed in advance, I can plot the TPF
+Since the cutoff \(c\) is not fixed in advance, one can plot the TPF
 against the FPF for all possible values of \(c\). This is exactly what
@@
 thus the diagonal line. A perfect test that is completely informative
-about disease status has \(TPF(c) = 1\) and \(FPF(c) = 0\) for all
-\(c\).
+about disease status has \(TPF(c) = 1\) and \(FPF(c) = 0\) for at least
+one value \(c\). If the assumption that support of \(M | D = 1\) is
+greater than that of \(M | D = 0\) holds, then the ROC curve will lie in
+the upper left quadrant above the diagonal line, however this may not be
+the case in a particular sample.
 
@@
 
-\[ \widehat{TPF(c)} = \frac{\sum_{i = 1}^n 1\{M_i > c\} \cdot 1\{D_i = 1\}}{\sum_{i=1}^n 1\{D_i = 1\}}, \]
+\[ \widehat{TPF(c)} = \frac{\sum_{i = 1}^n 1\{M_i \geq c\} \cdot 1\{D_i = 1\}}{\sum_{i=1}^n 1\{D_i = 1\}}, \]
 
@@
 color scale along the ROC curve itself. A legend for the color scale is
-placed along the vertical axis on the right of the plotting region. Due
-to the popularity of \pkg{ROCR}, my package supports the performance
-objects as returned by \pkg{ROCR}. The \pkg{pROC} package \citep{pROC}
-is mainly focused on estimating confidence intervals and regions for
-restricted ranges of the ROC curve. The plotting methods therein use the
-base \proglang{R} plotting functions to create nice displays of the
-curves along with shaded confidence regions. My \pkg{plotROC} package
-uses the \pkg{ggplot2} \citep{ggplot2} plotting library to create clear,
-informative ROC plots, with interactive features for use on the web, and
-sensible defaults for use in print.
+placed along the vertical axis on the right of the plotting region. The
+\pkg{pROC} package \citep{pROC} provides an option for plotting cutoff
+labels (\code{print.thres = TRUE}) and is mainly focused on estimating
+confidence intervals and regions for restricted ranges of the ROC curve.
+The plotting methods therein use the base \proglang{R} plotting
+functions to create nice displays of the curves along with shaded
+confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2}
+\citep{ggplot2} plotting library to create clear, informative ROC plots,
+with interactive features for use on the web, and sensible defaults for
+use in print.
 
-\subsection{Motivation}\label{motivation}
+\subsection{Motivation and Literature
+Review}\label{motivation-and-literature-review}
 
@@
 
+The vast majority of the figures that were reviewed looked more like
+those in panels A and B than those in C and D. While plots like this do
+technically display the trade-offs between false- and true-positives,
+they minimize the amount of useful information that can be displayed in
+the space needed to plot an ROC curve. The plots created by
+\pkg{plotROC} attempt to increase the amount of information displayed in
+ROC charts that would otherwise be mostly white space. This is useful
+not only for print media, where space is limited, but also during data
+analysis. The analyst can quickly and easily view information that would
+otherwise be obscured by standard plotting software. The interactive
+features take this one step further, enhancing the plots with high
+density and easily accessible supplementary information.
+
 \section{Usage of the package}\label{usage-of-the-package}
@@
 
-I start by creating an example data set. The marker I generate is
-moderately accurate for predicting disease status. Next I use the
-\code{calculate_roc} function to compute the empirical ROC curve. The
-disease status need not be coded as 0/1, but if it is not,
-\code{calculate_roc} assumes (with a warning) that the lowest value in
-sort order signifies disease-free status. This returns a dataframe with
-three columns: the cutoff values, the TPF and the FPF.
+I start by creating an example data set. There are 2 markers, one that
+is moderately predictive and one that is not as predictive.
 
@@
 \begin{Sinput}
-R> D.ex <- rbinom(100, size = 1, prob = .5)
-R> M.ex <- rnorm(100, mean = D.ex)
-R> roc.estimate <- calculate_roc(M.ex, D.ex)
-R> head(roc.estimate)
+R> set.seed(2529)
+R> D.ex <- rbinom(200, size = 1, prob = .5)
+R> M1 <- rnorm(200, mean = D.ex, sd = .65)
+R> M2 <- rnorm(200, mean = D.ex, sd = 1.5)
+R> 
+R> test <- data.frame(D = D.ex, D.str = c("Healthy", "Ill")[D.ex + 1], 
++                    M1 = M1, M2 = M2, stringsAsFactors = FALSE)
 \end{Sinput}
-\begin{Soutput}
-          c TPF       FPF
-1 -2.182331   1 0.9811321
-2 -1.674012   1 0.9622642
-3 -1.568181   1 0.9433962
-4 -1.550585   1 0.9245283
-5 -1.539929   1 0.9056604
-6 -1.190555   1 0.8867925
-\end{Soutput}
 \end{Schunk}
 
-The \code{roc.estimate} object is passed to the \code{ggroc} function
-with an optional label. This creates a \code{ggplot} object of the ROC
-curve using the \pkg{ggplot2} package \citep{ggplot2}.
+\subsubsection{The Roc Geom}\label{the-roc-geom}
+
+As of version 1.0.1, the \pkg{ggplot2} package \citep{ggplot2} exports
+the previously internal functions \code{Geom} and \code{Stat}. This
+enables developers to create their own statistical tranformations and
+geometric layers, while enjoying all of the other features of
+\pkg{ggplot2}. We have implemented the empirical ROC curve estimate and
+the calculation of exact confidence regions as statistical
+transformations: \code{stat_roc} and \code{stat_rocci}, respectively. We
+have also defined geometric layers for the ROC curve and confidence
+regions for the ROC curve: \code{geom_roc} and \code{geom_rocci},
+respectively. For further discussion and details of the grammar of
+graphics as implemented in \pkg{ggplot2}, we refer readers to
+\citet{wickham2010layered} and the \pkg{ggplot2} vignettes.
+
+To use the ROC geometric layer, I use the \code{ggplot} function to
+define the aesthetic mappings, and the \code{geom_roc} function to add
+an ROC curve layer. The \code{geom_roc} function requires the named
+aesthetics \code{d} for disease status, and \code{m} for marker. By
+default, the ROC geom and stat are linked, so that when \code{geom_roc}
+is called, \code{stat_roc} does the computation, and when
+\code{stat_roc} is called, \code{geom_roc} is used to plot the layer.
+The disease status need not be coded as 0/1, but if it is not,
+\code{stat_roc} assumes (with a warning) that the lowest value in sort
+order signifies disease-free status.
 
@@
 \begin{Sinput}
-R> single.rocplot <- ggroc(roc.estimate, label = "Example")
+R> basicplot <- ggplot(test, aes(d = D, m = M1)) + geom_roc()
 \end{Sinput}
@@
 
-The \code{single.rocplot} object can be used to create an interactive
-plot and display it in the Rstudio viewer or default web browser by
-passing it to the \code{plot_interactive_roc} function. Give the
-function an optional path to an html file as an argument called
-\code{file} to save the interactive plot as a complete web page. A
-screen shot of an interactive plot is shown in Figure \ref{interact}.
-Hovering over the display shows the cutoff value at the point nearest to
-the cursor. Clicking makes the cutoff label stick until the next click,
-and if confidence regions are available, clicks will also display those
-as grey rectangles.
+The \code{geom_roc} layer includes the ROC curve line combined with
+points and labels to display the values of the biomarker at the
+different cutpoints. It accepts the argument \code{n.cuts} to define the
+number of cutpoints to display along the curve. Labels can be supressed
+by using \code{n.cuts = 0} or \code{labels = FALSE}, however points will
+be displayed in the latter case. The size of the labels and the number
+of significant digits can be adjusted with \code{labelsize} and
+\code{labelround}, respectively.
 
@@
 \begin{Sinput}
-R> plot_interactive_roc(single.rocplot)
+R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0)
+R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 5, labelsize = 5, labelround = 2)
+R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 50, labels = FALSE)
+\end{Sinput}
+\end{Schunk}
+
+\subsubsection{Confidence regions and the Rocci
+Geom}\label{confidence-regions-and-the-rocci-geom}
+
+It is common to compute confidence regions for points on the ROC curve
+using the \citet{clopper1934use} exact method. Briefly, exact confidence
+intervals are calculated for the \(FPF\) and \(TPF\) separately, each at
+level \(1 - \sqrt{1 - \alpha}\). Based on result 2.4 from
+\citet{pepe2003statistical}, the cross-product of these intervals yields
+a \(100 * (1 - \alpha)\) percent rectangular confidence region for the
+pair.
+
+This is implemented in the \code{stat_rocci} and displayed as a
+\code{geom_rocci} layer. These both require the same aesthetics as the
+ROC geom, \code{d} for disease status and \code{m} for marker. By
+default, a set of 3 evenly spaced points along the curve are chosed to
+display confidence regions. Points corresponding to the confidence
+regions are distiguished from the others with a different symbol. You
+can select points by passing a vector of values in the range of \code{m}
+to the \code{ci.at} argument. By default, the significance level
+\(\alpha\) is set to 0.05, this can be changed using the
+\code{sig.level} option. An example is shown in Figure \ref{ciex}.
+
+\begin{Schunk}
+\begin{Sinput}
+R> basicplot + geom_rocci()
+\end{Sinput}
+\begin{figure}
+\includegraphics{figure/test-a-ci-1} \caption{Illustration of \pkg{plotROC} with exact confidence regions. \label{ciex}}\label{fig:test-a-ci}
+\end{figure}
+\begin{Sinput}
+R> ## 
+R> ## basicplot + geom_rocci(sig.level = .01)
+R> ## ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0) +
+R> ##   geom_rocci(ci.at = quantile(M1, c(.1, .4, .5, .6, .9)))
+\end{Sinput}
+\end{Schunk}
+
+\subsubsection{Styles and Labels}\label{styles-and-labels}
+
+The same objects like \code{basicplot} with Roc and/or Rocci layers can
+be treated like any other ggplot objects. It can be printed to display
+the figure, and other layers can be added to the plot. We provide the
+function \texttt{style\_roc()} which is a layer containing a theme,
+modified gridlines, and axes. Adding the \texttt{style\_roc()} layer to
+the ggplot object creates a plot with sensible defaults for use in
+print. This function has options for the number and location of major
+and minor breaks, addition of the diagnoal guideline, axis labels, and
+any theme object created for use with ggplot2 can be supplied.
+
+The \code{direct_label} function takes a ggplot object as an argument
+and annotates the figure with a direct label with a automatically chosen
+location. It attempts to intellegently select an appropriate location
+for the label, but the location can be adjusted with
+\code{nudge_x, nudge_y} and \code{label.angle}. If the \code{labels}
+argument is NULL, it will take the name from the mapped aesthetic. A
+simple example with the default options is shown in Figure \ref{first}.
+
+\begin{Schunk}
+\begin{Sinput}
+R> direct_label(basicplot) + style_roc()
+\end{Sinput}
+\begin{figure}
+\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} for use in print. \label{first}}\label{fig:print}
+\end{figure}
+\begin{Sinput}
+R> ## 
+R> ## basicplot + style_roc(theme = theme_grey, xlab = "1 - Specificity")
+R> ## direct_label(basicplot, labels = "Biomarker", nudge_y = -.1) + style_roc()
+\end{Sinput}
+\end{Schunk}
+
+\subsubsection{Interactive ROC plots}\label{interactive-roc-plots}
+
+The \code{basicplot} object, which is of class ggplot, can be used to
+create an interactive plot and display it in the Rstudio viewer or
+default web browser by passing it to the \code{plot_interactive_roc}
+function. Give the function an optional path to an html file as an
+argument called \code{file} to save the interactive plot as a complete
+web page. A screen shot of an interactive plot is shown in Figure
+\ref{interact}. Hovering over the display shows the cutoff value at the
+point nearest to the cursor. Clicking makes the cutoff label stick until
+the next click, and if confidence regions are available, clicks will
+also display those as grey rectangles. By default,
+\code{plot_interactive_roc} removes any existing Rocci geom and adds a
+high-density layer of confidence regions. This can be suppressed by
+using the \code{add.cis = FALSE} option. The points and labels layer of
+the Roc geom can be hidden by using the \code{hide.points = TRUE}
+option. Then, points and labels will be displayed only when the mouse is
+hovering over the plotting region. Also by default, the \code{style_roc}
+function is applied, the settings can be modified by passing a call to
+that function.
+
+\begin{Schunk}
+\begin{Sinput}
+R> plot_interactive_roc(basicplot)
+R> plot_interactive_roc(basicplot, hide.points = TRUE)
+R> plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw()))
 \end{Sinput}
@@
 
+Hovering over the display shows the cutoff value at the point nearest to
+the cursor. Clicking makes the cutoff label stick until the next click,
+and if confidence regions are available, clicks will also display those
+as grey rectangles. The confidence regions are automatically detected.
+When the user clicks on the ROC curve, the confidence region for the TPF
+and FPF is overlaid using a grey rectangle. The label and region stick
+until the next click.
+
 An interactive ROC plot can be exported by using the
@@
 containing the necessary \proglang{HTML} and \proglang{JavaScript}. The
-character string can be copy-pasted into an html document, or better
-yet, incorporated directly into a dynamic document using \pkg{knitr}
-\citep{knitr}. In a \pkg{knitr} document, it is necessary to use the
-\code{cat} function on the results and use the chunk options
-\code{results = 'asis'} and \code{fig.keep = 'none'} so that the
-interactive plot is displayed correctly. For documents that contain
-multiple interactive plots, it is necessary to assign each plot a unique
-name using the \code{prefix} argument of \code{export_interactive_roc}.
-This is necessary to ensure that the \proglang{JavaScript} code
-manipulates the correct svg elements. For examples of interactive plots
-and how to incorporate them into \pkg{knitr} documents, see the package
-vignette (\code{vignette("examples", package = "plotROC")}) or the web
-page https://sachsmc.github.io/plotROC/. The next code block shows an
-example \pkg{knitr} chunk that can be used in an .Rmd document to
-display an interactive plot.
+\proglang{JavaScript} source can be omitted by using the
+\code{omit.js = TRUE} option. You may wish to do this when there are
+multiple interactive figures in a single document; the source only needs
+to be included once. The character string can be copy-pasted into an
+html document, or better yet, incorporated directly into a dynamic
+document using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it
+is necessary to use the \code{cat} function on the results and use the
+chunk options \code{results = 'asis'} and \code{fig.keep = 'none'} so
+that the interactive plot is displayed correctly. For documents that
+contain multiple interactive plots, it is necessary to assign each plot
+a unique name using the \code{prefix} argument of
+\code{export_interactive_roc}. This is necessary to ensure that the
+\proglang{JavaScript} code manipulates the correct svg elements. For
+examples of interactive plots and how to incorporate them into
+\pkg{knitr} documents, see the package vignette
+(\code{vignette("examples", package = "plotROC")}) or the web page
+https://sachsmc.github.io/plotROC/. The next code block shows an example
+\pkg{knitr} chunk that can be used in an .Rmd document to display an
+interactive plot.
 
@@
 
-The same \code{ggroc} object that I called \code{single.rocplot} can be
-used to generate an ROC plot suitable for use in print. It annotates the
-cutoff values and is completely in black and white. A simple example
-with the default options is shown in Figure \ref{first}.
+\subsubsection{Multiple ROC curves}\label{multiple-roc-curves}
+
+If you have grouping factors in your dataset, or you have multiple
+markers measured on the same subjects, you may wish to plot multiple ROC
+curves on the same plot. \pkg{plotROC} fully supports faceting and
+grouping done by \pkg{ggplot2}. In out example dataset, we have 2
+markers measured in a paired manner:
 
@@
 \begin{Sinput}
-R> plot_journal_roc(single.rocplot)
+R> head(test)
+\end{Sinput}
+\begin{Soutput}
+  D   D.str         M1          M2
+1 1     Ill 1.48117155 -2.50636605
+2 1     Ill 0.61994478  1.46861033
+3 0 Healthy 0.57613345  0.07532573
+4 1     Ill 0.85433197  2.41997703
+5 0 Healthy 0.05258342  0.01863718
+6 1     Ill 0.66703989  0.24732453
+\end{Soutput}
+\end{Schunk}
+
+These data are in wide format, with the 2 markers going across 2
+columns. \pkg{ggplot2} requires long format, with the marker result in a
+single column, and a third variable identifying the marker. We provide
+the convenience function \code{melt_roc} to perform this transformation.
+The arguments are the data frame, a name or index identifying the
+disease status column, and a vector of names or indices identifying the
+the markers. Optionally, the names argument gives a vector of names to
+assign to the marker, replacing their column names. The result is a data
+frame in long format.
+
+\begin{Schunk}
+\begin{Sinput}
+R> longtest <- melt_roc(test, "D", c("M1", "M2"))
+R> head(longtest)
+\end{Sinput}
+\begin{Soutput}
+    D          M name
+M11 1 1.48117155   M1
+M12 1 0.61994478   M1
+M13 0 0.57613345   M1
+M14 1 0.85433197   M1
+M15 0 0.05258342   M1
+M16 1 0.66703989   M1
+\end{Soutput}
+\end{Schunk}
+
+Then, the dataset can be passed to the \code{ggplot} function, with the
+marker name given as a grouping or faceting variable.
+
+\begin{Schunk}
+\begin{Sinput}
+R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
+R> ggplot(longtest, aes(d = D, m = M)) + geom_roc() + facet_wrap(~ name) + style_roc()
+R> ggplot(longtest, aes(d = D, m = M, linetype = name)) + geom_roc() + geom_rocci()
+R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
+R> pairplot <- ggplot(longtest, aes(d = D, m = M, color = name)) + 
++   geom_roc(show.legend = FALSE) + style_roc()
+R> direct_label(pairplot)
+R> 
+R> pairplot + geom_rocci()
+R> pairplot + geom_rocci(linetype = 1)
+\end{Sinput}
+\end{Schunk}
+
+Showing multiple curves is also useful when there is a factor that
+affects the classification accuracy of the test. Let's create another
+example dataset.
+
+\begin{Schunk}
+\begin{Sinput}
+R> D.cov <- rbinom(400, 1, .5)
+R> gender <- c("Male", "Female")[rbinom(400, 1, .49) + 1]
+R> M.diff <- rnorm(400, mean = D.cov, sd = ifelse(gender == "Male", .5, 1.5))
+R> 
+R> test.cov <- data.frame(D = D.cov, gender = gender, M = M.diff)
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{Sinput}
+R> bygend <- ggplot(test.cov, aes(d = D, m = M, color = gender)) + 
++   geom_roc(show.legend = FALSE)
+R> direct_label(bygend) + style_roc()
 \end{Sinput}
 \begin{figure}
-\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} for use in print. \label{first}}\label{fig:print}
+\includegraphics{figure/covplot-1} \caption{Illustration of \pkg{plotROC} with multiple curves. \label{multi}}\label{fig:covplot}
 \end{figure}
@@
 
-\subsubsection{Multiple ROC curves}\label{multiple-roc-curves}
-
-If you have multiple tests of different types measured on the same
-subjects, you can use the \code{calculate_multi_roc} function to compute
-the empirical ROC curve for each test. It returns a list of data frames
-with the estimates and cutoff values. Then the \code{multi_ggroc}
-function creates the appropriate type of \code{ggplot} object.
-Confidence regions are not supported for multiple curves at the time of
-writing.
-
-\begin{Schunk}
-\begin{Sinput}
-R> D.ex <- rbinom(100, 1, .5)
-R> 
-R> paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
-+                        M2 = rnorm(100, mean = D.ex, sd = .4), 
-+                        M3 = runif(100), D = D.ex)
-R> 
-R> estimate.list <- calculate_multi_roc(paired.data, 
-+                                      c("M1", "M2", "M3"), "D")
-\end{Sinput}
-\end{Schunk}
-
-Labels can be added easily with the \code{label} option of
-\code{multi_ggroc}. The length of the label element should match the
-number of plotted curves. The \code{multi_ggroc} object can be passed to
-the \code{plot_journal_roc} and the \code{export_interactive_roc}
-functions, as desired. The resulting plot is shown in Figure
-\ref{multi}.
-
-\begin{Schunk}
-\begin{Sinput}
-R> multi.rocplot <- multi_ggroc(estimate.list, label = c("M1", "M2", "M3"))
-R> plot_journal_roc(multi.rocplot)
-\end{Sinput}
-\begin{figure}
-\includegraphics{figure/multi2-1} \caption{Illustration of \pkg{plotROC} plot with multiple curves. \label{multi} }\label{fig:multi2}
-\end{figure}
-\end{Schunk}
-
-Both \code{plot_journal_roc} and \code{export_interactive_roc} support a
-number of options to customize the look of the plots. By default,
-multiple curves are distinguished by different line types and direct
-labels. For multiple ROC curves that are similar to one another, those
-defaults can make it difficult to interpret the plots. Therefore, we
-also support colors and legends. The x- and y-axis can be changed by
-passing options to \code{ggroc} or \code{multi_ggroc}. The next code
-block illustrates the available options.
-
-\begin{Schunk}
-\begin{Sinput}
-R> colorplot <- multi_ggroc(estimate.list, 
-+                          xlabel = "1 - Specificity", 
-+                          ylabel = "Sensitivity")
-R> cat(
-+   export_interactive_roc(colorplot, lty = rep(1, 3), 
-+                          color = c("black", "purple", "orange"), 
-+                          legend = TRUE)
-+   )
-\end{Sinput}
-\end{Schunk}
+Interactive versions of the plots with grouping and faceting are fully
+supported.
 
 \subsection{Advanced options}\label{advanced-options}
-
-\subsubsection{Click to view confidence
-region}\label{click-to-view-confidence-region}
-
-I use the \code{ci = TRUE} option in \code{calculcate_roc} and
-\code{ggroc} to compute confidence regions for points on the ROC curve
-using the \citet{clopper1934use} exact method. Briefly, exact confidence
-intervals are calculated for the \(FPF\) and \(TPF\) separately, each at
-level \(1 - \sqrt{1 - \alpha}\). Based on result 2.4 from
-\citet{pepe2003statistical}, the cross-product of these intervals yields
-a \(100 * (1 - \alpha)\) percent rectangular confidence region for the
-pair. The significance level can be specified using the \code{alpha}
-option.
-
-\begin{Schunk}
-\begin{Sinput}
-R> roc.ci <- calculate_roc(paired.data$M1, paired.data$D, 
-+                         ci = TRUE, alpha = 0.05)
-R> ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
-\end{Sinput}
-\end{Schunk}
-
-For interactive plots, the confidence regions are automatically
-detected. When the user clicks on the ROC curve, the confidence region
-for the TPF and FPF is overlaid using a grey rectangle. The label and
-region stick until the next click.
-
-\begin{Schunk}
-\begin{Sinput}
-R> cat(
-+   export_interactive_roc(ci.rocplot, 
-+                          prefix = "aci")
-+   )
-\end{Sinput}
-\end{Schunk}
-
-For use in print, I pass a small vector of cutoff locations at which to
-display the confidence regions. This is shown in Figure \ref{conf}.
-
-\begin{Schunk}
-\begin{Sinput}
-R> plot_journal_roc(ci.rocplot, n.cuts = 10, 
-+                  ci.at = c(-.5, .5, 2.1))
-\end{Sinput}
-\begin{figure}
-\includegraphics{figure/printci-1} \caption{Illustration of \pkg{plotROC} plot with exact confidence regions. \label{conf}}\label{fig:printci}
-\end{figure}
-\end{Schunk}
 
@@
 title, axis label, and AUC annotation is shown in Figure \ref{annotate}.
+\pkg{plotROC} provides the convenience function \code{calc_auc} that
+takes a ggplot object that has an Roc layer, extracts the data, and
+calculates the AUC.
 
@@
 \begin{Sinput}
-R> library(ggplot2)
-R> plot_journal_roc(ci.rocplot, n.cuts = 10, 
-+                  ci.at = c(-.5, .5, 2.1)) + 
-+   theme_grey() + 
-+   geom_abline(intercept = 0, slope = 1, color = "white") + 
+R> annotate <- basicplot + style_roc(theme = theme_grey) +
++   theme(axis.text = element_text(colour = "blue")) +
 +   ggtitle("Themes and annotations") + 
 +   annotate("text", x = .75, y = .25, 
-+            label = "AUC = 0.80") +
++            label = paste("AUC =", round(calc_auc(basicplot)["AUC"], 2))) +
 +   scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
 \end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
 \begin{figure}
@@
 \end{Schunk}
+
+The results are compatible with all other \pkg{ggplot2} layers and
+functions, and interactive versions are supported as long as there is an
+Roc layer present.
 
@@
 the TPF and the FPF are returned. Then you can simply pass those values
-in a data frame to the \code{ggroc} function. For example, let us use
-the binormal method to create a smooth curve. This approach assumes that
-the test distribution is normal conditional on disease status.
+in a data frame to the \code{ggplot} function and override the default
+statistical tranformation to \code{stat_identity}. For example, let us
+use the binormal method to create a smooth curve. This approach assumes
+that the test distribution is normal conditional on disease status.
 
@@
 \begin{Sinput}
-R> D.ex <- rbinom(100, 1, .5)
-R> M.ex <- rnorm(100, mean = D.ex, sd = .5)
-R> 
+R> D.ex <- test$D
+R> M.ex <- test$M1
 R> mu1 <- mean(M.ex[D.ex == 1])
@@
 +                              )
+R> 
+R> binorm.plot <- ggplot(binorm.roc, aes(x = FPF, y = TPF, label = c)) + 
++   geom_roc(stat = "identity") + style_roc(theme = theme_grey)
 \end{Sinput}
@@
 
-Then I can pass this \code{data.frame} to the \code{ggroc} function as
-before. The example is shown in Figure \ref{binorm}.
+The example is shown in Figure \ref{binorm}. Interactive plots with
+\code{stat = 'identity'} are not currently supported.
 
@@
 \begin{Sinput}
-R> binorm.plot <- ggroc(binorm.roc, label = "Binormal")
-R> plot_journal_roc(binorm.plot)
+R> binorm.plot
 \end{Sinput}
@@
 \end{figure}
-\end{Schunk}
-
-\pkg{plotROC} also supports the use of the \pkg{ROCR} estimation method.
-Objects as returned by the function \code{rocr::performance} can be
-passed directly to \code{ggroc}:
-
-\begin{Schunk}
-\begin{Sinput}
-R> library(ROCR)
-R> rocr.est <- performance(prediction(M.ex, D.ex), "tpr", "fpr")
-R> rocr.plot <- ggroc(rocr.est, label = "ROCR object")
-\end{Sinput}
 \end{Schunk}
@@
 R> 
-R> surv.plot <- multi_ggroc(sroc, label = paste("t =", c(2, 5, 10)))
+R> sroclong <- do.call(rbind, sroc)
+R> 
+R> survplot <- ggplot(sroclong, aes(x = FPF, y = TPF, label = c, color = time)) + 
++   geom_roc(labels = FALSE, stat = "identity")
 \end{Sinput}
@@
 interactive plots. The first step in the process is to create
-\code{ggplot} objects using the \code{ggroc} or \code{multi_ggroc}
-functions. These functions return standard \code{ggplot} objects that
-include basic styling, hidden cutoff labels, and hidden confidence
-regions. They can be plotted and inspected in the \proglang{R} console.
-These form the basis for both the print versions and the interactive
-versions of the plots. Creating a print version by using the
-\code{plot_journal_roc} function simply makes visible a subset of the
-hidden cutoff labels and confidence regions, if available.
+\code{ggplot} objects with Roc and/or Rocci layers. They can be plotted
+and inspected in the \proglang{R} console. These form the basis for both
+the print versions and the interactive versions of the plots. Basic
+styling and labeling can be added with the \code{style_roc} and
+\code{direct_label} functions.
 
@@
 returned in the character string provided by
-\code{export_interactive_roc}. Figure \ref{flow} outlines the
-\pkg{plotROC} process.
-
-\begin{figure}[ht]
-\centering
-\includegraphics{figure/diagram.pdf}
-\caption{Flowchart illustrating the approach that \pkg{plotROC} takes to generate either static plots for print or interactive plots for web-use. \label{flow}}
-\end{figure}
+\code{export_interactive_roc}.
 
@@
 tools, I find that specialized functions make the results clearer and
-more informative.
+more informative. The functions are integrated into the existing and
+popular \pkg{ggplot2} package, so that all the benefits and features
+therein can be used effectively.
 
@@
 \begin{Soutput}
-R version 3.1.2 (2014-10-31)
+R version 3.2.2 (2015-08-14)
 Platform: x86_64-apple-darwin13.4.0 (64-bit)
+Running under: OS X 10.10.5 (Yosemite)
 
@@
 other attached packages:
- [1] DiagrammeR_0.4    survivalROC_1.0.3 ROCR_1.0-5       
- [4] gplots_2.16.0     ggplot2_1.0.0     tikzDevice_0.8.1 
- [7] plotROC_1.3.3     xtable_1.7-4      stringr_0.6.2    
-[10] knitr_1.9
+[1] survivalROC_1.0.3  tikzDevice_0.8.1   plotROC_2.0.0     
+[4] ggplot2_1.0.1.9003 xtable_1.7-4       stringr_1.0.0     
+[7] knitr_1.11
 
 loaded via a namespace (and not attached):
- [1] bitops_1.0-6       caTools_1.17.1     colorspace_1.2-4  
- [4] digest_0.6.8       evaluate_0.5.5     filehash_2.2-2    
- [7] formatR_1.0.3      gdata_2.13.3       grid_3.1.2        
-[10] gtable_0.1.2       gtools_3.4.1       highr_0.4         
-[13] htmltools_0.2.6    htmlwidgets_0.3.2  KernSmooth_2.23-14
-[16] MASS_7.3-39        munsell_0.4.2      plyr_1.8.1        
-[19] proto_0.3-10       Rcpp_0.11.4        reshape2_1.4.1    
-[22] RJSONIO_1.3-0      rstudioapi_0.2     scales_0.2.4      
-[25] tools_3.1.2        yaml_2.1.13       
+ [1] Rcpp_0.12.1      digest_0.6.8     grid_3.2.2       plyr_1.8.3      
+ [5] gtable_0.1.2     formatR_1.2.1    magrittr_1.5     evaluate_0.8    
+ [9] scales_0.3.0     highr_0.5.1      stringi_0.5-5    labeling_0.3    
+[13] tools_3.2.2      munsell_0.4.2    colorspace_1.2-6 filehash_2.3    
 \end{Soutput}

### 2015-10-15 Paper/response-to-review.md
final responses and edits

@@
+# Summary of changes
+
+Thanks to the reviewers' comments, and some recent changes to ggplot2, the plotROC package has been greatly improved.
+The interface is now much more user-friendly, requires less typing, and is completely compatible with ggplot2.
+The computation is more efficient, and the plotting tools are much more flexible.
+I believe the package is much closer to my goal of providing a better tool for plotting ROC curves.
+Here are the main changes:
+
+- The calculate_roc, ggroc, and plot_journal_roc functions are deprecated. 
+- I have implemented the geom_roc, geom_rocci geometric layers, and stat_roc, stat_rocci statistical transformations.
+A recent update to ggplot2 now exports the Geom and Stat functions so that developers can create their own and be compatible with the other features of ggplot2.
+This is what we have done instead of the previous approach of "hacking" the existing ggplot2 geoms.
+The usage is now simplified to:
+
+    p1 <- ggplot(data, aes(m = M, d = D)) + geom_roc() + geom_rocci()
+
+ - The stat functions perform the calculations and the geoms add the visual elements. The user only has to specify the marker value m and binary outcome d, optionally with grouping or faceting. 
+ - The resulting ggplot object can be passed to the plot_interactive_roc or export_interactive_roc functions to create interactive figures. Grouping and faceting are fully supported. 
+ - The geom_roc displays cutoff labels by default.
+We have included an additional function style_roc() that adds the layer of styling to match the previous plot_journal_roc().
+Its usage is as simple as adding it to a ggplot object: p1 + style_roc().
+ - Direct labels can be added now with the direct_label() function. This takes the ggplot object as an argument so that it can extract the names: direct_label(p1). 
+ - The AUC can be calculated by using the calc_auc() function on a ggplot object: calc_auc(p1). It calculates the AUC for each group or facet present in the plot. 
+ - We provide the convenience function melt_roc(), which takes a dataset with multiple markers in different columns, and converts it to long format for use with the ggplot function.
+An example of its usage is in the paper.
+
+The new version is not currently on CRAN, but I expect it will be in the next few months after ggplot2 is updated. To install, follow these two steps:
+
+devtools::install_github("hadley/ggplot2")
+devtools::install_github("sachsmc/plotROC")
+
+I thank the reviewers and editors for their helpful comments, and I hope you find the updated paper and package to be vastly improved. Below are point-by-point responses to the reviewers' comments.
+
 # Reviewer 1
@@
 
-After clicking on a point around 1.6, I got the screenshot shown in Figure 1. This seems related to the rounding (here to the 2nd decimal).
-Figure 1: the CI region doesn't span the cutoff point.
+> After clicking on a point around 1.6, I got the screenshot shown in Figure 1. This seems related to the rounding (here to the 2nd decimal).
 
@@
 I believe this was due to the way the previous function handled ties in the results.
-The computation of the ROC curve in R and the js functions for interactivity have been completely revamped.
+The computation of the ROC curve in R and the javascript functions for interactivity have been completely revamped.
 I cannot reproduce the error using the new update in any browser that I've tested, so I believe it has been fixed.
@@
 
+I added another paragraph of discussion at the end of the section about the literature review and how it motivated the package development.
+The discussion of the review is now its own subsection in the introduction.
+It is difficult to discuss at length because it is mostly a subjective assessment of the figures.
+
 > • Strictly speaking one could argue the plotROC package "obscures the cutoffs", as they are not displayed by default when calling the ggroc function.
@@
 The abstract and introduction should be rephrased in a more neutral way on this topic, as this is not the key point addressed by this package.
+
+In the update, calling the geom_roc function includes cutoff labels by default. The labels and points are inherently part of the roc geometric layer, and the user can optionally suppress them.
+
 ## Introduction:
 > • The introduction states that "ROC curves have been used for decades in signal detection theory". In fact, they originate from this field (s/n in radio receivers, hence the name)
+
+This has been noted in the first paragraph of the introduction.
+
 > • The "good outcome", "bad outcome" is confusing to design correct / incorrect classifications. Especially in the medical field. I don't think it is necessary.
+
+This section has been rewritten for clarity, and the confusion matrix has been added as a table to clarify the terminology.
+
 > • "A perfect test [...] has TPF(c) = 1 and FPF(c) = 0 for all c". This is not correct. It is sufficient that this equality holds true for a single c.
+
+Thank you, this error has been corrected.
+
 > • The formula hat(TPF(c)) is slightly confusing. It may be useful to first introduce the confusion table with true | false positive | negatives. It seems correct nonetheless.
+
+I have added a phrase clarifying what the indicator function is, and a sentence simply stating that the formula is simply counting the proportion of test positives out of the group of disease positives.
+
 > • "The greater the AUC, the more informative the test".
@@
 An AUC = 0 is also the result of a perfect test (for which M < c if I am not mistaken).
+
+This is true when you consider the case where the marker and disease status are negatively correlated.
+To avoid this issue, as is typical in the ROC curve literature, we make the implicit assumption that larger values of M are more indicative of D = 1.
+We note this in the introduction, and point out that it is possible for the estimate to lie below the diagonal in a given sample.
+
 > • Existing software: pROC allows to plot cutoffs quite easily (with print.thres=TRUE, just checked now). Of course it still uses the base R plots.
+
+Thank you for pointing this out, I had overlooked it. This is noted in the summary of existing software.
+
 > • I am surprised a Pubmed search returned only 54 papers. This part would probably deserve to be expanded a little bit, and maybe turned into a whole section.
+
+I was also surprised by this, and perhaps my search term was too restrictive.
+I just checked again, and the search returned 59 papers (October 15).
+The search term is included in the paper source in the litreview/pubmed-search-term.txt file.
+As mentioned above the subsection called "Motivation" was renamed to Literature Review, and slightly expanded.
+
 ## Usage of the package:
 > • Plots such as fig 4 will be hard to read in small figures, as is often seen in journals.
+
+If the image is scaled properly before printing, the fonts should appear in a readable size.
+This is of course adjustable by using the labelsize option of geom_roc.
+Figure 1 is scaled appropriately, and I find it to be readable.
+
 > • Fig 5 is unclear. Especially, what was the cutoff used to compute the region, as several ones are displayed in the region?
+
+I have updated the Rocci geom so that the point where the region is calculated is marked as an "X" by default, to distinguish it from the other points. I hope this makes it more clear.
+
 > • In general, the use of shaded region instead of bars is quite surprising, although perhaps more accurate than bars.
+
+I did consider using boxes instead of shaded regions, and it is possible to do this using the linetype = 1 and alpha.box = 0 options.
+I am opposed to bars because it is difficult to determine where the boundary is in the corners of the region.
+
 > • Why are the regions rectangle-shaped and not oval? That would seem even more correct, although probably quite difficult to set up.
+
+They are rectangles because they are a cross product of two 1-dimensional confidence intervals.
+It is possible to get oval regions if you assume bivariate normality, but I think the exact confidence regions are a better match for the empirical estimate.
+If you think there's interest, I would consider adding a stat_binormal to compute that estimate as well.
+
 > • The gridlines in figures 3-5 & 7 are so thin and light-coloured that they did not appear when I printed this manuscript on paper. The printer was not (yet) running out of ink.
+
+This is a result of the default theme in ggplot2, which unfortunately I cannot change.
+Of course the user can change the appearance of the grid lines using the + theme(panel.grid.major = ..., panel.grid.minor = ...) construction.
+
 > • When reporting an AUC and confidence intervals in a publication, it seems to me at least as important to report values in addition to a plot.
@@
 
+Agreed, and you are correct in your assessment that the AUC=0.80 label was fake.
+I have added a function called calc_auc that takes a ggplot object that contains an Roc layer, extracts the data, and calculates the auc.
+It computes the AUC for each group and returns a data frame containing the group specification and a column containing the AUC estimate.
+An example of its usage is given on page 13 and in the help file.
+
 ## Package:
@@
 This will create a lot of headaches for infrequent users: is it calculate_, or create_, or generate_, or ...?
+
+I think you will find the update much more user friendly, and consistent with the general usage of ggplot2.
+
 > • This wouldn't be an issue if users could browse the functions of the package as one can usually do by typing ?plotROC.
@@
 Of course one can always find this paper, but it is much easier to get a proper help.
+
+This stub has been updated with descriptions, links to key functions, and a series of examples.
+
 > • It is of course a design choice, but the use of underscores over dots (or camelCase) in function names seems quite surprising and may disturb some users.
@@
 R Foundation for Statistical Computing for a review of current usage of function name styles.
+
+This was a personal choice and unfortunately it won't please everyone. I opted to be consistent with the ggplot2 package, which uses underscores for functions and dots for objects and argument names.
+
 > • Wouldn't it be more idiomatic to introduce the roc as a ggplot geometry? That would allow some much more straightforward syntax like:
@@
 
+Yes I totally agree, and now is is possible since the Geom and Stat functions are exported from ggplot2.
+However, the qplot function in ggplot2 is now being deprecated, thus it is necessary to use the ggplot(data, aes(m = M, d = D)) + geom_roc() construction.
+I expect the updates to be on CRAN in the next few months.
+
 > • I don't understand why one has to pass a ggroc argument to plot_interactive_roc.
@@
 I suspect a few well designed objects would make it much more user-friendly.
+
+I believe the new design is much more user friendly and flexible, although it is still necessary to pass the ggplot object to plot_interactive_roc.
+
 > • What do the modCss and modJs functions do? The documentation is really limited and an example would help. Same with verify_d. As it stands now these functions seem useless.
+
+These functions have been removed, but to answer your question, they were internal functions that inserted the CSS and javascript source into the html documents generated by export_interactive_roc.
+In the update, the javascript source files are self-contained in the inst directory, and are pasted into the documents using the getD3() function, which is also internal.
+verify_d is also an internal function that checks the class labels and converts them to 0/1.
+These functions are not intended to be used by end-users.
+
 > • The help files lack examples altogether.
 Not only does it help a lot to have copy- pasteable code snippets as a start, but it can also help checking the package for potential regressions during updates.
+
+I've added a number of examples to the main help files, including the package help file ?plotROC.
+
 > • Performance: as most R packages doing ROC curves, this one scales pretty badly with larger sample sizes.
@@
 ROCR returns nearly instantly with such a low (sic!) sample size and scales much higher.
+
+The computation has been rewritten to use cumsum instead of a loop.
+It now scales up much better.
+There are still bottlenecks in the rendering functions done by ggplot2 and gridSVG, although once the output is written, the interactive figures with large sample sizes are quite responsive.
+
 > • How do I add the cut-offs to the plot? If I say ggplot(..., ci.at = 0.5) I get an unused argument error and nothing is added. Is this only available in the interactive plot?
+
+Cutoffs are included by default, and the user can adjust the number of cutoffs using the geom_roc(n.cuts = x) option.
+For now, the placement of cutoffs is automatically determined and set to be evenly spaced.
+Confidence regions can be placed at specified points using the geom_rocc(ci.at = c(a, b, c)) construction.
 
@@
 PlotROC should either “Depend” on that package, as for shiny, or catch its absence and print the error message immediately when the shiny app is launched.
+
+This has been changed and now the only hard dependency is ggplot2.
+
 > • On my Ubuntu system, the Download button opens a 404 error page with URL as http://127.0.0.1:3048/session/61edd0aac13532dc34788b220e9d27c8/downlo ad/printDownload?w= instead of downloading the image.
@@
 The download works properly on my Mac where a tab shortly opens before the download starts.
+I cannot replicate the error, perhaps it was a shiny bug that got fixed?
+Is this using the locally running version of the shiny app, or the shinyapps.io hosted version?
+Which browsers did this occur in?
+It seems to work on Ubuntu 14.04 with Firefox.
+
 > • The browse button appears very low in the page and is not visible enough IMO.
+
+I swapped the locations of the data selector and the upload button. Also there is some bolded help text to draw attention to the upload button.
 
@@
 
+It was my intention to develop a useful tool for plotting, which seemed to be lacking from the other packages that perform many useful statistical features as you have mentioned.
+I have included an additional function for calculating the AUC, but I think the other features are best left to the other packages.
+With the recent update to ggplot2, it is now possible to add additional stats and geoms.
+A potential future line of work would be to implement some of the statistical transformations and alternative estimates for ROC curves.
+
 > Also the paper and package suffer from a number of ambiguities and inaccuracies (either technical or in the description):
@@
 This could lead to problems when plotting curves that don't span the full 0/0 to 1/1 space and also when computing AUCs.
+
+This has been corrected in the updated version. The (0,0) and (1,1) points are now always included in the computation output.
+
 	- plotROC seems to implictly define that samples ">" the threshold are positive, samples "<=" the threshold are negative.
@@
 the ROCR package) that define samples ">=" the threshold as a positive prediction.
+
+This was an oversight and I thank you for pointing out the error. This has been corrected in the computation and in the text of the introduction.
+
 	- It is not obvious how plotROC handles ties (i.e. samples with identical numerical measurements)
+
+The computation handles ties correctly by including them in the calculation of the proportions, and then removing the duplicated elements from the results.
+This is the same as what is implemented in the ROCR package.
+
 	- The paper defines "good outcome" and "bad outcome" in Section 1.1. These definitions seem to ignore half of the confusion matrix (i.e. ignoring true and false negatives).
+
+This section has been revamped and edited for clarity. The good and bad outcome terminology has been removed and the confusion matrix illustrating the concepts has been added as table 1.
+
 	- In "Help and Documentation" of the webpage, the confusion matrix and the definitions seem to be copied from Wikipedia. Credits and reference are needed.
+
+This is correct, it has been adapted from Wikipedia. I added a reference and link.
+
 	- The website states "confidence level = 0.05". Should this be "significance level" instead?
+
+Yes this has been changed to significance level.
 
@@
 The current tool does not have the maturity and comprehensiveness of other tools - therefore the users will be left with the choice between alternative tools, each with their own strengths and weaknesses.
+
+Thank you for the praise and useful criticism, I believe the updated package is much more user friendly and flexible as a plotting tool.

### 2015-10-15 Paper/sachs2015plotROC.Rmd
final responses and edits

@@
 
-opts_chunk$set(dev="tikz", prompt = TRUE, sanitize = TRUE)
+opts_chunk$set(dev="tikz", prompt = TRUE, sanitize = TRUE, comment = NA)
 
@@
 In medicine, ROC curves have a long history of use for evaluating diagnostic tests in radiology and general diagnostics.
-ROC curves originated in the field of signal detection theory .
+ROC curves originated in the field of signal detection theory.
 
-For a continuous measurement that I denote as $M$, convention dictates that a test positive is defined as $M$ equalling or exceeding some fixed threshold $c$: $M \geq c$.
+For a continuous measurement that I denote as $M$, convention dictates that a test positive is defined as $M$ equalling or exceeding some threshold $c$: $M \geq c$.
 The goal of ROC analysis is to evaluate the classification accuracy of $M$ in reference to the binary outcome $D$, which takes possible values $0$ (negative) or $1$ (positive).
@@
 The four cells of the matrix correspond to the possible classification outcomes: a true positive, a false positive, a true negative, and a false negative.
-ROC analysis assesses the trade-offs between the test's fraction of true positives versus the false positives.
+ROC analysis assesses the trade-offs between the test's fraction of true positives versus the false positives as $c$ varies over the range of $M$.
 
@@
 
-$$ FPF(c) = P\{ M \geq c | D = 0 \} $$
+$$ FPF(c) = P\{ M \geq c | D = 0 \}. $$
 
@@
 Given a sample of test and disease status pairs, $(M_1, D_1), \ldots, (M_n, D_n)$, I can estimate the ROC curve by computing proportions in the diseased and healthy subgroups separately.
-Specifically, given a fixed cutoff $c$, an estimate of the $TPF(c)$ is
+Specifically, given a fixed cutoff $c$, an estimate of the $TPF(c)$ is the proportion of test positives among the diseased subgroup:
 
@@
 
-where $1\{\cdot\}$ is the indicator function.
-An estimate for $FPF(c)$ is given by a similar expression with $D_i = 1$ replaced with $D_i = 0$.
+where $1\{\cdot\}$ is the indicator function that equals 1 when the condition inside the parentheses is true and 0 otherwise.
+An estimate for $FPF(c)$ is the proportion of test positives among the healthy subgroup and is given by a similar expression with $D_i = 1$ replaced with $D_i = 0$.
 Calculating these proportions for $c$ equal to each unique value of the observed $M_i$ yields what is known as the empirical ROC curve estimate.
@@
 
-## Motivation and Literature Review
+## Literature Review
 
@@
 I searched Pubmed for clinical trials or observational studies in humans reported in major oncology journals for the past 10 years for the terms "ROC Curve" OR "ROC Analysis" OR "Receiver operating characteristic curve".
-The search was conducted on October 8, 2014 and returned 54 papers.
-From those papers, 47 images were extracted and reviewed.
+I conducted the search on October 8, 2014 and it returned 54 papers.
+From those papers, I extracted and reviewed 47 images.
 The exact specifications for the Pubmed query are available in the manuscript source files.
@@
 Each image consisted of a single ROC curve plot or a panel of multiple plots.
-Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of a diagonal guide line, whether any cutpoints were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the AUC.
+I inspected each plot manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of a diagonal guide line, whether any cutpoints were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the AUC.
 The numerical results of the survey are summarized in Table \ref{table1}.
@@
 
-The vast majority of the figures that were reviewed looked more like those in panels A and B than those in C and D.
+The vast majority of the figures that I reviewed looked more like those in panels A and B than those in C and D.
 While plots like this do technically display the trade-offs between false- and true-positives, they minimize the amount of useful information that can be displayed in the space needed to plot an ROC curve.
@@
 As of version 1.0.1, the \pkg{ggplot2} package [@ggplot2] exports the previously internal functions \code{Geom} and \code{Stat}.
-This enables developers to create their own statistical tranformations and geometric layers, while enjoying all of the other features of \pkg{ggplot2}.
+This enables developers to create their own statistical tranformations (stats) and geometric layers (geoms), while enjoying all of the other features of \pkg{ggplot2}.
 We have implemented the empirical ROC curve estimate and the calculation of exact confidence regions as statistical transformations: \code{stat_roc} and \code{stat_rocci}, respectively.
@@
 To use the ROC geometric layer, I use the \code{ggplot} function to define the aesthetic mappings, and the \code{geom_roc} function to add an ROC curve layer.
-The \code{geom_roc} function requires the named aesthetics \code{d} for disease status, and \code{m} for marker.
+The \code{geom_roc} function requires that the named aesthetics \code{d} for disease status, and \code{m} for marker be present in the \code{aes} function call inside \code{ggplot}.
 By default, the ROC geom and stat are linked, so that when \code{geom_roc} is called, \code{stat_roc} does the computation, and when \code{stat_roc} is called, \code{geom_roc} is used to plot the layer.
@@
 ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0)
-ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 5, labelsize = 5, labelround = 2)
+ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 5, 
+                                            labelsize = 5, labelround = 2)
 ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 50, labels = FALSE)
@@
 These both require the same aesthetics as the ROC geom, \code{d} for disease status and \code{m} for marker.
-By default, a set of 3 evenly spaced points along the curve are chosed to display confidence regions.
+By default, a set of 3 evenly spaced points along the curve are chosen to display confidence regions.
 Points corresponding to the confidence regions are distiguished from the others with a different symbol.
@@
 basicplot + geom_rocci()
-
 basicplot + geom_rocci(sig.level = .01)
@@
 Adding the `style_roc()` layer to the ggplot object creates a plot with sensible defaults for use in print.
-This function has options for the number and location of major and minor breaks, addition of the diagnoal guideline, axis labels, and any theme object created for use with ggplot2 can be supplied.
+This function has options for the number and location of major and minor breaks, addition of the diagonal guideline, axis labels, and any theme object created for use with ggplot2 can be supplied.
 
@@
 
-```{r print, eval = 1, fig.width = 6, fig.height = 6, fig.cap = "Illustration of ROC curve plot generated by \\pkg{plotROC} for use in print. \\label{first}"}
+```{r print, eval = 1, fig.width = 6, fig.height = 6, fig.cap = "Illustration of ROC curve plot generated by \\pkg{plotROC} with the styles and direct labels applied for use in print. \\label{first}"}
 direct_label(basicplot) + style_roc()
-
 basicplot + style_roc(theme = theme_grey, xlab = "1 - Specificity")
-direct_label(basicplot, labels = "Biomarker", nudge_y = -.1) + style_roc()
+direct_label(basicplot, labels = "Biomarker", nudge_y = -.1)
 ```
@@
 Then, points and labels will be displayed only when the mouse is hovering over the plotting region.
-Also by default, the \code{style_roc} function is applied, the settings can be modified by passing a call to that function.
+Also by default, the \code{style_roc} function is applied, the settings can be modified by passing a call to that function or setting it to NULL.
 
@@
 
-Then, the dataset can be passed to the \code{ggplot} function, with the marker name given as a grouping or faceting variable.
+Then, the dataset can be passed to the \code{ggplot} function, with the marker name given as a grouping or faceting variable. Some examples of its usage are given below.
 
 ```{r group, eval = FALSE}
-ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
-ggplot(longtest, aes(d = D, m = M)) + geom_roc() + facet_wrap(~ name) + style_roc()
+ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc()
+ggplot(longtest, aes(d = D, m = M)) + geom_roc() + facet_wrap(~ name)
 ggplot(longtest, aes(d = D, m = M, linetype = name)) + geom_roc() + geom_rocci()
-ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
+ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() 
 pairplot <- ggplot(longtest, aes(d = D, m = M, color = name)) + 
-  geom_roc(show.legend = FALSE) + style_roc()
+  geom_roc(show.legend = FALSE)
 direct_label(pairplot)
@@
 gender <- c("Male", "Female")[rbinom(400, 1, .49) + 1]
-M.diff <- rnorm(400, mean = D.cov, sd = ifelse(gender == "Male", .5, 1.5))
+M.diff <- rnorm(400, mean = D.cov, 
+                sd = ifelse(gender == "Male", .5, 1.5))
 
@@
 Therefore, themes and annotations can be added in the usual \pkg{ggplot2} way.
-A \code{plot_journal_roc} figure with a new theme, title, axis label, and AUC annotation is shown in Figure \ref{annotate}.
+A figure with a new theme, title, axis label, and AUC annotation is shown in Figure \ref{annotate}.
 \pkg{plotROC} provides the convenience function \code{calc_auc} that takes a ggplot object that has an Roc layer, extracts the data, and calculates the AUC.
@@
 binorm.plot <- ggplot(binorm.roc, aes(x = FPF, y = TPF, label = c)) + 
-  geom_roc(stat = "identity") + style_roc(theme = theme_grey)
+  geom_roc(stat = "identity") + style_roc()
 ```
@@
 
-```{r binormal, echo = TRUE, fig.width=6, fig.height=6, fig.cap = "Illustration of smooth binormal ROC curve. \\label{binorm}"}
+```{r binormal, echo = FALSE, fig.width=6, fig.height=6, fig.cap = "Illustration of smooth binormal ROC curve. \\label{binorm}"}
 binorm.plot
@@
 This also allows users to manipulate the figures directly in \proglang{R} to suit their needs, using tools that are more accessible and familiar to most \proglang{R} users.
+Then, the \proglang{JavaScript} adds a layer of interactivity to the rendered figures.
 

### 2015-10-15 Paper/sachs2015plotROC.tex
final responses and edits

@@
 diagnostic tests in radiology and general diagnostics. ROC curves
-originated in the field of signal detection theory .
+originated in the field of signal detection theory.
 
@@
 that a test positive is defined as \(M\) equalling or exceeding some
-fixed threshold \(c\): \(M \geq c\). The goal of ROC analysis is to
-evaluate the classification accuracy of \(M\) in reference to the binary
-outcome \(D\), which takes possible values \(0\) (negative) or \(1\)
-(positive). It is implicitly assumed that the subpopulation having
-\(D = 1\) tends to have larger values of \(M\) compared to the
-subpopulation with \(D = 0\). The classification accuracy of \(M\) can
-be evaluated by considering the confusion matrix (Table \ref{confus}).
-The confusion matrix cross-classifies the predicted outcome \(M \geq c\)
-versus the true outcome \(D\). The four cells of the matrix correspond
-to the possible classification outcomes: a true positive, a false
-positive, a true negative, and a false negative. ROC analysis assesses
-the trade-offs between the test's fraction of true positives versus the
-false positives.
+threshold \(c\): \(M \geq c\). The goal of ROC analysis is to evaluate
+the classification accuracy of \(M\) in reference to the binary outcome
+\(D\), which takes possible values \(0\) (negative) or \(1\) (positive).
+It is implicitly assumed that the subpopulation having \(D = 1\) tends
+to have larger values of \(M\) compared to the subpopulation with
+\(D = 0\). The classification accuracy of \(M\) can be evaluated by
+considering the confusion matrix (Table \ref{confus}). The confusion
+matrix cross-classifies the predicted outcome \(M \geq c\) versus the
+true outcome \(D\). The four cells of the matrix correspond to the
+possible classification outcomes: a true positive, a false positive, a
+true negative, and a false negative. ROC analysis assesses the
+trade-offs between the test's fraction of true positives versus the
+false positives as \(c\) varies over the range of \(M\).
 
@@
 
-\[ FPF(c) = P\{ M \geq c | D = 0 \} \]
+\[ FPF(c) = P\{ M \geq c | D = 0 \}. \]
 
@@
 Specifically, given a fixed cutoff \(c\), an estimate of the \(TPF(c)\)
-is
+is the proportion of test positives among the diseased subgroup:
 
@@
 
-where \(1\{\cdot\}\) is the indicator function. An estimate for
-\(FPF(c)\) is given by a similar expression with \(D_i = 1\) replaced
+where \(1\{\cdot\}\) is the indicator function that equals 1 when the
+condition inside the parentheses is true and 0 otherwise. An estimate
+for \(FPF(c)\) is the proportion of test positives among the healthy
+subgroup and is given by a similar expression with \(D_i = 1\) replaced
 with \(D_i = 0\). Calculating these proportions for \(c\) equal to each
@@
 
-\subsection{Motivation and Literature
-Review}\label{motivation-and-literature-review}
+\subsection{Literature Review}\label{literature-review}
 
@@
 ``ROC Curve'' OR ``ROC Analysis'' OR ``Receiver operating characteristic
-curve''. The search was conducted on October 8, 2014 and returned 54
-papers. From those papers, 47 images were extracted and reviewed. The
-exact specifications for the Pubmed query are available in the
-manuscript source files.
+curve''. I conducted the search on October 8, 2014 and it returned 54
+papers. From those papers, I extracted and reviewed 47 images. The exact
+specifications for the Pubmed query are available in the manuscript
+source files.
 
 Each image consisted of a single ROC curve plot or a panel of multiple
-plots. Each plot was inspected manually for the following design
-features: the number of curves displayed, the type of axis labels
-(sensitivity/ 1 - specificity or true/false positive fractions),
-presence or absence of grid lines, presence or absence of a diagonal
-guide line, whether any cutpoints were indicated, the type of curve
-label (legend or direct label), and presence of other textual
-annotations such as the AUC. The numerical results of the survey are
-summarized in Table \ref{table1}.
+plots. I inspected each plot manually for the following design features:
+the number of curves displayed, the type of axis labels (sensitivity/ 1
+- specificity or true/false positive fractions), presence or absence of
+grid lines, presence or absence of a diagonal guide line, whether any
+cutpoints were indicated, the type of curve label (legend or direct
+label), and presence of other textual annotations such as the AUC. The
+numerical results of the survey are summarized in Table \ref{table1}.
 
@@
 
-The vast majority of the figures that were reviewed looked more like
-those in panels A and B than those in C and D. While plots like this do
+The vast majority of the figures that I reviewed looked more like those
+in panels A and B than those in C and D. While plots like this do
 technically display the trade-offs between false- and true-positives,
@@
 the previously internal functions \code{Geom} and \code{Stat}. This
-enables developers to create their own statistical tranformations and
-geometric layers, while enjoying all of the other features of
-\pkg{ggplot2}. We have implemented the empirical ROC curve estimate and
-the calculation of exact confidence regions as statistical
+enables developers to create their own statistical tranformations
+(stats) and geometric layers (geoms), while enjoying all of the other
+features of \pkg{ggplot2}. We have implemented the empirical ROC curve
+estimate and the calculation of exact confidence regions as statistical
 transformations: \code{stat_roc} and \code{stat_rocci}, respectively. We
@@
 define the aesthetic mappings, and the \code{geom_roc} function to add
-an ROC curve layer. The \code{geom_roc} function requires the named
-aesthetics \code{d} for disease status, and \code{m} for marker. By
+an ROC curve layer. The \code{geom_roc} function requires that the named
+aesthetics \code{d} for disease status, and \code{m} for marker be
+present in the \code{aes} function call inside \code{ggplot}. By
 default, the ROC geom and stat are linked, so that when \code{geom_roc}
@@
 R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0)
-R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 5, labelsize = 5, labelround = 2)
+R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 5, 
++                                             labelsize = 5, labelround = 2)
 R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 50, labels = FALSE)
@@
 ROC geom, \code{d} for disease status and \code{m} for marker. By
-default, a set of 3 evenly spaced points along the curve are chosed to
+default, a set of 3 evenly spaced points along the curve are chosen to
 display confidence regions. Points corresponding to the confidence
@@
 \begin{Sinput}
-R> ## 
 R> ## basicplot + geom_rocci(sig.level = .01)
@@
 print. This function has options for the number and location of major
-and minor breaks, addition of the diagnoal guideline, axis labels, and
+and minor breaks, addition of the diagonal guideline, axis labels, and
 any theme object created for use with ggplot2 can be supplied.
@@
 \begin{figure}
-\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} for use in print. \label{first}}\label{fig:print}
+\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} with the styles and direct labels applied for use in print. \label{first}}\label{fig:print}
 \end{figure}
 \begin{Sinput}
-R> ## 
 R> ## basicplot + style_roc(theme = theme_grey, xlab = "1 - Specificity")
-R> ## direct_label(basicplot, labels = "Biomarker", nudge_y = -.1) + style_roc()
+R> ## direct_label(basicplot, labels = "Biomarker", nudge_y = -.1)
 \end{Sinput}
@@
 function is applied, the settings can be modified by passing a call to
-that function.
+that function or setting it to NULL.
 
@@
 Then, the dataset can be passed to the \code{ggplot} function, with the
-marker name given as a grouping or faceting variable.
+marker name given as a grouping or faceting variable. Some examples of
+its usage are given below.
 
@@
 \begin{Sinput}
-R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
-R> ggplot(longtest, aes(d = D, m = M)) + geom_roc() + facet_wrap(~ name) + style_roc()
+R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc()
+R> ggplot(longtest, aes(d = D, m = M)) + geom_roc() + facet_wrap(~ name)
 R> ggplot(longtest, aes(d = D, m = M, linetype = name)) + geom_roc() + geom_rocci()
-R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
+R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() 
 R> pairplot <- ggplot(longtest, aes(d = D, m = M, color = name)) + 
-+   geom_roc(show.legend = FALSE) + style_roc()
++   geom_roc(show.legend = FALSE)
 R> direct_label(pairplot)
@@
 R> gender <- c("Male", "Female")[rbinom(400, 1, .49) + 1]
-R> M.diff <- rnorm(400, mean = D.cov, sd = ifelse(gender == "Male", .5, 1.5))
+R> M.diff <- rnorm(400, mean = D.cov, 
++                 sd = ifelse(gender == "Male", .5, 1.5))
 R> 
@@
 plotted. Therefore, themes and annotations can be added in the usual
-\pkg{ggplot2} way. A \code{plot_journal_roc} figure with a new theme,
-title, axis label, and AUC annotation is shown in Figure \ref{annotate}.
-\pkg{plotROC} provides the convenience function \code{calc_auc} that
-takes a ggplot object that has an Roc layer, extracts the data, and
-calculates the AUC.
+\pkg{ggplot2} way. A figure with a new theme, title, axis label, and AUC
+annotation is shown in Figure \ref{annotate}. \pkg{plotROC} provides the
+convenience function \code{calc_auc} that takes a ggplot object that has
+an Roc layer, extracts the data, and calculates the AUC.
 
@@
 R> binorm.plot <- ggplot(binorm.roc, aes(x = FPF, y = TPF, label = c)) + 
-+   geom_roc(stat = "identity") + style_roc(theme = theme_grey)
++   geom_roc(stat = "identity") + style_roc()
 \end{Sinput}
@@
 \begin{Schunk}
-\begin{Sinput}
-R> binorm.plot
-\end{Sinput}
 \begin{figure}
@@
 directly in \proglang{R} to suit their needs, using tools that are more
-accessible and familiar to most \proglang{R} users.
+accessible and familiar to most \proglang{R} users. Then, the
+\proglang{JavaScript} adds a layer of interactivity to the rendered
+figures.
 

### 2015-10-19 Paper/response-to-review.md
Revised and resubmitted

@@
 
+```r
 p1 <- ggplot(data, aes(m = M, d = D)) + geom_roc() + geom_rocci() 
+```
 
@@
 
+```r
 devtools::install_github("hadley/ggplot2")
 devtools::install_github("sachsmc/plotROC")
+```
 
@@
 
+```r
 set.seed(42)
@@
 plot_interactive_roc(ggroc(roc.estimate, ci = TRUE))
+```
 
@@
 
-The manuscript has been edited to be consistent with the use of active voice.
+I have edited the manuscript to be more consistent with the use of active voice.
 
@@
 > • Wouldn't it be more idiomatic to introduce the roc as a ggplot geometry? That would allow some much more straightforward syntax like:
-data.for.roc <- data.frame(M.ex, D.ex)
-qplot(M.ex, D.ex, data = data.for.roc) + geom_roc()
-or suchlike that would be much easier to remember.
+> ```r
+> data.for.roc <- data.frame(M.ex, D.ex)
+> qplot(M.ex, D.ex, data = data.for.roc) + geom_roc()
+> ```
+>or suchlike that would be much easier to remember.
 
@@
 The download works properly on my Mac where a tab shortly opens before the download starts.
+
 I cannot replicate the error, perhaps it was a shiny bug that got fixed?
@@
 
-	- plotROC seems to implictly define that samples ">" the threshold are positive, samples "<=" the threshold are negative.
+>	- plotROC seems to implictly define that samples ">" the threshold are positive, samples "<=" the threshold are negative.
 In this it differs from e.g.
@@
 
-	- It is not obvious how plotROC handles ties (i.e. samples with identical numerical measurements)
+>	- It is not obvious how plotROC handles ties (i.e. samples with identical numerical measurements)
 
@@
 
-	- The paper defines "good outcome" and "bad outcome" in Section 1.1. These definitions seem to ignore half of the confusion matrix (i.e. ignoring true and false negatives).
+>	- The paper defines "good outcome" and "bad outcome" in Section 1.1. These definitions seem to ignore half of the confusion matrix (i.e. ignoring true and false negatives).
 
@@
 
-	- In "Help and Documentation" of the webpage, the confusion matrix and the definitions seem to be copied from Wikipedia. Credits and reference are needed.
+>	- In "Help and Documentation" of the webpage, the confusion matrix and the definitions seem to be copied from Wikipedia. Credits and reference are needed.
 
@@
 
-	- The website states "confidence level = 0.05". Should this be "significance level" instead?
+>	- The website states "confidence level = 0.05". Should this be "significance level" instead?
 

### 2015-10-19 Paper/sachs2015plotROC.Rmd
Revised and resubmitted

@@
 \begin{table}
+\begin{center}
 \begin{tabular}{l|c|c|c}
@@
 \caption{\label{confus} Confusion matrix illustrating the possible classification outcomes of a continuous test $M$ with threshold $c$ versus the true outcome $D$. }
+\end{center}
 \end{table}
@@
 
-\pkg{plotROC} can be installed from the Comprehensive \proglang{R} Archive Network, or installed from source.
+Users can install the latest version of \pkg{plotROC} from github, as shown below.
+The package will appear on the Comprehensive \proglang{R} Archive Network in the near future.
+It can also be downloaded from https://github.com/sachsmc/plotROC and installed from source.
+As of October 19, 2015, \pkg{plotROC} also requires the development version of \pkg{ggplot2} also available from github.
 
 ```{r load, eval = FALSE}
-install.packages("plotROC")
+devtools::install_github("hadley/ggplot2")
+devtools::install_github("sachsmc/plotROC")
 library(plotROC)
@@
 plot_interactive_roc(basicplot, hide.points = TRUE)
-plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw()))
+plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw))
 ```
@@
 
-An interactive ROC plot can be exported by using the \code{export_interactive_roc} function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}.
+Users can export an interactive ROC plot by using the \code{export_interactive_roc} function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}.
 The \proglang{JavaScript} source can be omitted by using the \code{omit.js = TRUE} option.
-You may wish to do this when there are multiple interactive figures in a single document; the source only needs to be included once.
+Users may wish to do this when there are multiple interactive figures in a single document; the source only needs to be included once.
 The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr].
@@
 If you have grouping factors in your dataset, or you have multiple markers measured on the same subjects, you may wish to plot multiple ROC curves on the same plot.
-\pkg{plotROC} fully supports faceting and grouping done by \pkg{ggplot2}.
-In out example dataset, we have 2 markers measured in a paired manner:
+\pkg{plotROC} fully supports faceting and grouping as done by \pkg{ggplot2}.
+In my example dataset, there are 2 markers measured in a paired manner:
 
@@
 \pkg{ggplot2} requires long format, with the marker result in a single column, and a third variable identifying the marker.
-We provide the convenience function \code{melt_roc} to perform this transformation.
+I provide the convenience function \code{melt_roc} to perform this transformation.
 The arguments are the data frame, a name or index identifying the disease status column, and a vector of names or indices identifying the the markers.
@@
 
-Interactive versions of the plots with grouping and faceting are fully supported.
+This example is displayed in Figure \ref{multi}. Interactive versions of the plots with grouping and faceting are fully supported.
 
@@
 \pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be plotted.
-Therefore, themes and annotations can be added in the usual \pkg{ggplot2} way.
+Therefore, users can add themes and annotations in the usual \pkg{ggplot2} way.
 A figure with a new theme, title, axis label, and AUC annotation is shown in Figure \ref{annotate}.

### 2015-10-19 Paper/sachs2015plotROC.tex
Revised and resubmitted

@@
 
-\pkg{plotROC} can be installed from the Comprehensive \proglang{R}
-Archive Network, or installed from source.
+Users can install the latest version of \pkg{plotROC} from github, as
+shown below. The package will appear on the Comprehensive \proglang{R}
+Archive Network in the near future. It can also be downloaded from
+https://github.com/sachsmc/plotROC and installed from source. As of
+October 19, 2015, \pkg{plotROC} also requires the development version of
+\pkg{ggplot2} also available from github.
 
@@
 \begin{Sinput}
-R> install.packages("plotROC")
+R> devtools::install_github("hadley/ggplot2")
+R> devtools::install_github("sachsmc/plotROC")
 R> library(plotROC)
@@
 R> plot_interactive_roc(basicplot, hide.points = TRUE)
-R> plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw()))
+R> plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw))
 \end{Sinput}
@@
 
-An interactive ROC plot can be exported by using the
+Users can export an interactive ROC plot by using the
 \code{export_interactive_roc} function, which returns a character string
@@
 \proglang{JavaScript} source can be omitted by using the
-\code{omit.js = TRUE} option. You may wish to do this when there are
+\code{omit.js = TRUE} option. Users may wish to do this when there are
 multiple interactive figures in a single document; the source only needs
@@
 curves on the same plot. \pkg{plotROC} fully supports faceting and
-grouping done by \pkg{ggplot2}. In out example dataset, we have 2
+grouping as done by \pkg{ggplot2}. In my example dataset, there are 2
 markers measured in a paired manner:
@@
 columns. \pkg{ggplot2} requires long format, with the marker result in a
-single column, and a third variable identifying the marker. We provide
+single column, and a third variable identifying the marker. I provide
 the convenience function \code{melt_roc} to perform this transformation.
@@
 
-Interactive versions of the plots with grouping and faceting are fully
-supported.
+This example is displayed in Figure \ref{multi}. Interactive versions of
+the plots with grouping and faceting are fully supported.
 
@@
 \pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be
-plotted. Therefore, themes and annotations can be added in the usual
+plotted. Therefore, users can add themes and annotations in the usual
 \pkg{ggplot2} way. A figure with a new theme, title, axis label, and AUC

### 2015-10-19 Paper/sachs2015plotROC.R

@@ -10,7 +10,7 @@ knit_hooks$set(plot = function(x, options) {
 
 options(prompt = "R> ")
 
-opts_chunk$set(dev="tikz", prompt = TRUE)
+opts_chunk$set(dev="tikz", prompt = TRUE, sanitize = TRUE, comment = NA)
 
 render_sweave()
 
@@ -68,7 +68,7 @@ print(xtable(table1, label = "table1",
       sanitize.text.function = function(x) gsub("\t", "$\\quad$", x, fixed = TRUE))
 
 
-## ----figure1, fig.width = 6, fig.height = 6, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves. Panel A shows a sparse ROC curve, with no design additions inside the plotting region. The plot results in more white space than anything else. It is difficult to accurately determine values without reference lines. Panel B shows a plot comparing 2 curves, with different line types and a legend. AUCs are also given in the legend. Panels C and D add gridlines, diagonal reference lines, and direct labels. \\label{figure1}"----
+## ----figure1, fig.width = 6, fig.height = 6, message = FALSE, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves. Panel A shows a sparse ROC curve, with no design additions inside the plotting region. The plot results in more white space than anything else. It is difficult to accurately determine values without reference lines. Panel B shows a plot comparing 2 curves, with different line types and a legend. AUCs are also given in the legend. Panels C and D add gridlines, diagonal reference lines, and direct labels. \\label{figure1}"----
 library(plotROC)
 
 set.seed(520)
@@ -89,12 +89,15 @@ lines(TPF ~ FPF, data = R2, type = 's', lty = 2)
 legend("bottomright", legend = c("Test A: AUC = 0.81", "Test B: AUC = 0.64"), lty = 1:2, cex = .75)
 
 # C
-plot(TPF ~ FPF, data = R1, type = 's', lwd  = 2, xlab = "False positive fraction", ylab = "True positive fraction", main = "C")
+plot(TPF ~ FPF, data = R1, type = 's', lwd  = 2, xlab = "False positive fraction", ylab = "True positive fraction", main = "C", xlim = c(-.05, 1))
[…]

### 2016-07-15 Paper/sachs2015plotROC.Rmd
revision number two

@@
 ---
-title: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
-shorttitle: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
-plaintitle: 'plotROC: A Better Tool for Plotting ROC Curves'
+title: '\pkg{plotROC}: A Tool for Plotting ROC Curves'
+shorttitle: '\pkg{plotROC}: A Tool for Plotting ROC Curves'
+plaintitle: 'plotROC: A Tool for Plotting ROC Curves'
 author:
@@
 
-Given a sample of test and disease status pairs, $(M_1, D_1), \ldots, (M_n, D_n)$, I can estimate the ROC curve by computing proportions in the diseased and healthy subgroups separately.
+Given a sample of test and disease status pairs, $(M_1, D_1), \ldots, (M_n, D_n)$, one can estimate the ROC curve by computing proportions in the diseased and healthy subgroups separately.
 Specifically, given a fixed cutoff $c$, an estimate of the $TPF(c)$ is the proportion of test positives among the diseased subgroup:
@@
 There are also extensions that allow for estimation with time-to-event outcomes subject to censoring.
-For a more thorough reference on the methods and theory surrounding ROC curves, I refer interested readers to @pepe2003statistical.
+For a more thorough reference on the methods and theory surrounding ROC curves, interested readers may look to @pepe2003statistical.
 
@@
 Sometimes, the FPF and TPF are incorrectly referred to as rates, using the abbreviations FPR and TPR.
-These are probabilities and their estimates are proportions, therefore I prefer the use of the term fraction as opposed to rate.
+These are probabilities and their estimates are proportions, therefore the use of the term fraction as opposed to rate is preferred.
 
@@
 Viewers can identify ROC plots through context, by observing the shape of the line, and through the addition of axis labels, titles, legends, and so on.
-In my review of the oncology literature, I observed plots with the distinctive characteristics of the plotting functions from Microsoft Excel [@excel], \proglang{SAS} [@sas], \proglang{SPSS} [@spss], and the base \proglang{R} plotting functions [@arr].
+In a review of the oncology literature, I observed plots with the distinctive characteristics of the plotting functions from Microsoft Excel [@excel], \proglang{SAS} [@sas], \proglang{SPSS} [@spss], and the base \proglang{R} plotting functions [@arr].
 
@@
 The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions.
-My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.
+The \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting package to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.
 
@@
 Direct labels are preferred over legends because they omit the additional cognitive step of matching line types or colors to labels.
-My \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use.
+The \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use.
 Exact confidence regions for points on the ROC curve are optionally calculated and displayed.
-Additionally, I use axis scales adjusted to be denser near the margins 0 and 1.
-In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins.
+Additionally, the use of axis scales adjusted to be denser near the margins 0 and 1 facilitates accurate reading of FPF and TPF values.
+In medical applications, it is often necessary to have a very low FPF (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins.
 The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.
@@
 
-The vast majority of the figures that I reviewed looked more like those in panels A and B than those in C and D.
+The vast majority of the figures that were reviewed looked more like those in panels A and B than those in C and D.
 While plots like this do technically display the trade-offs between false- and true-positives, they minimize the amount of useful information that can be displayed in the space needed to plot an ROC curve.
@@
 
-I created a \pkg{shiny} application [@shiny] in order to make the features more accessible to non-\proglang{R} users.
+A \pkg{shiny} application [@shiny] was created in order to make the features more accessible to non-\proglang{R} users.
 A limited subset of the functions of \pkg{plotROC} can be performed on an example dataset or on data that users upload to the website.
@@
 
-Users can install the latest version of \pkg{plotROC} from github, as shown below.
-The package will appear on the Comprehensive \proglang{R} Archive Network in the near future.
-It can also be downloaded from https://github.com/sachsmc/plotROC and installed from source.
-As of October 19, 2015, \pkg{plotROC} also requires the development version of \pkg{ggplot2} also available from github.
+Users can install the latest version of \pkg{plotROC} from the Comprehensive \proglang{R} Archive Network (CRAN).
+It can also be downloaded or installed from https://github.com/sachsmc/plotROC, where active development will take place.
+\pkg{plotROC} also requires at least version 2.0.1 of \pkg{ggplot2} which is also available from CRAN.
 
 ```{r load, eval = FALSE}
-devtools::install_github("hadley/ggplot2")
-devtools::install_github("sachsmc/plotROC")
+install.packages(c("ggplot2", "plotROC"))
 library(plotROC)
@@
 
-As of version 1.0.1, the \pkg{ggplot2} package [@ggplot2] exports the previously internal functions \code{Geom} and \code{Stat}.
+As of version 2.0.1, the \pkg{ggplot2} package [@ggplot2] exports the previously internal functions \code{Geom} and \code{Stat}.
 This enables developers to create their own statistical tranformations (stats) and geometric layers (geoms), while enjoying all of the other features of \pkg{ggplot2}.
-We have implemented the empirical ROC curve estimate and the calculation of exact confidence regions as statistical transformations: \code{stat_roc} and \code{stat_rocci}, respectively.
-We have also defined geometric layers for the ROC curve and confidence regions for the ROC curve: \code{geom_roc} and \code{geom_rocci}, respectively.
-For further discussion and details of the grammar of graphics as implemented in \pkg{ggplot2}, we refer readers to @wickham2010layered and the \pkg{ggplot2} vignettes.
+I have implemented the empirical ROC curve estimate and the calculation of exact confidence regions as statistical transformations: \code{stat_roc} and \code{stat_rocci}, respectively.
+I have also defined geometric layers for the ROC curve and confidence regions for the ROC curve: \code{geom_roc} and \code{geom_rocci}, respectively.
+For further discussion and details of the grammar of graphics as implemented in \pkg{ggplot2}, I refer readers to @wickham2010layered and the \pkg{ggplot2} vignettes.
 
 To use the ROC geometric layer, I use the \code{ggplot} function to define the aesthetic mappings, and the \code{geom_roc} function to add an ROC curve layer.
-The \code{geom_roc} function requires that the named aesthetics \code{d} for disease status, and \code{m} for marker be present in the \code{aes} function call inside \code{ggplot}.
+The \code{geom_roc} function requires that the named aesthetics \code{d} for disease status, and \code{m} for marker, be present in the \code{aes} function call inside \code{ggplot}.
 By default, the ROC geom and stat are linked, so that when \code{geom_roc} is called, \code{stat_roc} does the computation, and when \code{stat_roc} is called, \code{geom_roc} is used to plot the layer.
@@
 The same objects like \code{basicplot} with Roc and/or Rocci layers can be treated like any other ggplot objects.
-It can be printed to display the figure, and other layers can be added to the plot.
-We provide the function `style_roc()` which is a layer containing a theme, modified gridlines, and axes.
+They can be printed to display the figure, and other layers can be added to the plot.
+I provide the function `style_roc()` which is a layer containing a theme, modified gridlines, and axes.
 Adding the `style_roc()` layer to the ggplot object creates a plot with sensible defaults for use in print.
@@
 \end{figure}
-
-Hovering over the display shows the cutoff value at the point nearest to the cursor.
-Clicking makes the cutoff label stick until the next click, and if confidence regions are available, clicks will also display those as grey rectangles.
-The confidence regions are automatically detected.
-When the user clicks on the ROC curve, the confidence region for the TPF and FPF is overlaid using a grey rectangle.
-The label and region stick until the next click.
 
@@
 By default \code{calculate_roc} computes the empirical ROC curve.
-There are other estimation methods out there, as I have summarized in the introduction.
+There are other estimation methods out there, as summarized in the introduction.
 Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned.
@@
 This function maps each element of the plot to a corresponding element of the svg markup language.
-I keep track of the names of the points and labels elements so that I can add interactivity using \pkg{d3.js} and \proglang{JavaScript}.
-The main interactive feature I wanted was to be able to display the cutoff labels at the points on the ROC curve closest to the mouse cursor.
+Interactivity is then added using \pkg{d3.js} and \proglang{JavaScript} to manipulate those svg elements in response to users' input.
+The main interactive feature is to display the cutoff labels at the points on the ROC curve closest to the mouse cursor.
 
-There are many ways to solve this with \pkg{d3.js}, but I decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve.
-The idea is that for the set of cutoff points along the ROC curve, the \code{d3.geom.voronoi} function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point.
+There are many ways to solve this with \pkg{d3.js}, but Voronoi polygons are a convenient and efficient way to map the cursor location to approximately the nearest point on the ROC curve.
+For the set of cutoff points along the ROC curve, the \code{d3.geom.voronoi} function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point.
 Hover events are bound to the polygons so that when the mouse cursor moves around the plotting region, the closest point on the ROC curve is made visible.
@@
 This approach is similar to what is done in the \pkg{gridSVG} \code{grid.animate} function, which uses the svg \code{<animate />} tags.
-However, the available features were not sufficient for my needs, which is why I used \pkg{d3.js}.
+However, \pkg{d3.js} provides a much richer set of features.
 There are several other \proglang{R} packages that aim to create interactive figures.
@@
 \pkg{qtlcharts} [@qtlcharts] uses a set of custom \proglang{JavaScript} and \pkg{d3.js} functions to visualize data from genetic experiments.
-Their general approach is to manipulate the data and create options in \proglang{R}, and then let the charting libraries or functions handle the rendering and interactivity.
+Their general approach is to manipulate the data and record plotting options in \proglang{R}, and then pass those objects to the charting libraries or functions that handle the rendering and interactivity.
 \pkg{plotROC} lets \proglang{R} do the rendering, allowing the figures to be consistent across print and web-based media, and retaining the distinctive \proglang{R} style.

### 2016-07-15 Paper/sachs2015plotROC.tex
revision number two

@@
 
-\title{\pkg{plotROC}: A Better Tool for Plotting ROC Curves}
+\title{\pkg{plotROC}: A Tool for Plotting ROC Curves}
 
@@
 \Plainauthor{ Michael C. Sachs} %% comma-separated
-\Plaintitle{plotROC: A Better Tool for Plotting ROC Curves} %% without formatting
+\Plaintitle{plotROC: A Tool for Plotting ROC Curves} %% without formatting
 
-\Shorttitle{\pkg{plotROC}: A Better Tool for Plotting ROC Curves} %% a short title (if necessary)
+\Shorttitle{\pkg{plotROC}: A Tool for Plotting ROC Curves} %% a short title (if necessary)
 %% an abstract and keywords
@@
 Given a sample of test and disease status pairs,
-\((M_1, D_1), \ldots, (M_n, D_n)\), I can estimate the ROC curve by
+\((M_1, D_1), \ldots, (M_n, D_n)\), one can estimate the ROC curve by
 computing proportions in the diseased and healthy subgroups separately.
@@
 subject to censoring. For a more thorough reference on the methods and
-theory surrounding ROC curves, I refer interested readers to
+theory surrounding ROC curves, interested readers may look to
 \citet{pepe2003statistical}.
@@
 rates, using the abbreviations FPR and TPR. These are probabilities and
-their estimates are proportions, therefore I prefer the use of the term
-fraction as opposed to rate.
+their estimates are proportions, therefore the use of the term fraction
+as opposed to rate is preferred.
 
@@
 ROC plots through context, by observing the shape of the line, and
-through the addition of axis labels, titles, legends, and so on. In my
+through the addition of axis labels, titles, legends, and so on. In a
 review of the oncology literature, I observed plots with the distinctive
@@
 functions to create nice displays of the curves along with shaded
-confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2}
-\citep{ggplot2} plotting library to create clear, informative ROC plots,
+confidence regions. The \pkg{plotROC} package uses the \pkg{ggplot2}
+\citep{ggplot2} plotting package to create clear, informative ROC plots,
 with interactive features for use on the web, and sensible defaults for
@@
 omit the additional cognitive step of matching line types or colors to
-labels. My \pkg{plotROC} package additionally provides plotting of
+labels. The \pkg{plotROC} package additionally provides plotting of
 cutoff values, which are displayed interactively with the web-based
@@
 for points on the ROC curve are optionally calculated and displayed.
-Additionally, I use axis scales adjusted to be denser near the margins 0
-and 1. In medical applications, it is often necessary to have a very low
-FPR (less than 10\%, for instance), therefore the smaller scales are
-useful for accurately determining values near the margins. The next
-section details the usage of the \pkg{plotROC} \proglang{R} package and
-these features.
+Additionally, the use of axis scales adjusted to be denser near the
+margins 0 and 1 facilitates accurate reading of FPF and TPF values. In
+medical applications, it is often necessary to have a very low FPF (less
+than 10\%, for instance), therefore the smaller scales are useful for
+accurately determining values near the margins. The next section details
+the usage of the \pkg{plotROC} \proglang{R} package and these features.
 
@@
 
-The vast majority of the figures that I reviewed looked more like those
-in panels A and B than those in C and D. While plots like this do
+The vast majority of the figures that were reviewed looked more like
+those in panels A and B than those in C and D. While plots like this do
 technically display the trade-offs between false- and true-positives,
@@
 
-I created a \pkg{shiny} application \citep{shiny} in order to make the
+A \pkg{shiny} application \citep{shiny} was created in order to make the
 features more accessible to non-\proglang{R} users. A limited subset of
@@
 
-Users can install the latest version of \pkg{plotROC} from github, as
-shown below. The package will appear on the Comprehensive \proglang{R}
-Archive Network in the near future. It can also be downloaded from
-https://github.com/sachsmc/plotROC and installed from source. As of
-October 19, 2015, \pkg{plotROC} also requires the development version of
-\pkg{ggplot2} also available from github.
+Users can install the latest version of \pkg{plotROC} from the
+Comprehensive \proglang{R} Archive Network (CRAN). It can also be
+downloaded or installed from https://github.com/sachsmc/plotROC, where
+active development will take place. \pkg{plotROC} also requires at least
+version 2.0.1 of \pkg{ggplot2} which is also available from CRAN.
 
@@
 \begin{Sinput}
-R> devtools::install_github("hadley/ggplot2")
-R> devtools::install_github("sachsmc/plotROC")
+R> install.packages(c("ggplot2", "plotROC"))
 R> library(plotROC)
@@
 
-As of version 1.0.1, the \pkg{ggplot2} package \citep{ggplot2} exports
+As of version 2.0.1, the \pkg{ggplot2} package \citep{ggplot2} exports
 the previously internal functions \code{Geom} and \code{Stat}. This
@@
 (stats) and geometric layers (geoms), while enjoying all of the other
-features of \pkg{ggplot2}. We have implemented the empirical ROC curve
+features of \pkg{ggplot2}. I have implemented the empirical ROC curve
 estimate and the calculation of exact confidence regions as statistical
-transformations: \code{stat_roc} and \code{stat_rocci}, respectively. We
+transformations: \code{stat_roc} and \code{stat_rocci}, respectively. I
 have also defined geometric layers for the ROC curve and confidence
@@
 respectively. For further discussion and details of the grammar of
-graphics as implemented in \pkg{ggplot2}, we refer readers to
+graphics as implemented in \pkg{ggplot2}, I refer readers to
 \citet{wickham2010layered} and the \pkg{ggplot2} vignettes.
@@
 an ROC curve layer. The \code{geom_roc} function requires that the named
-aesthetics \code{d} for disease status, and \code{m} for marker be
+aesthetics \code{d} for disease status, and \code{m} for marker, be
 present in the \code{aes} function call inside \code{ggplot}. By
@@
 The same objects like \code{basicplot} with Roc and/or Rocci layers can
-be treated like any other ggplot objects. It can be printed to display
-the figure, and other layers can be added to the plot. We provide the
+be treated like any other ggplot objects. They can be printed to display
+the figure, and other layers can be added to the plot. I provide the
 function \texttt{style\_roc()} which is a layer containing a theme,
@@
 \end{figure}
-
-Hovering over the display shows the cutoff value at the point nearest to
-the cursor. Clicking makes the cutoff label stick until the next click,
-and if confidence regions are available, clicks will also display those
-as grey rectangles. The confidence regions are automatically detected.
-When the user clicks on the ROC curve, the confidence region for the TPF
-and FPF is overlaid using a grey rectangle. The label and region stick
-until the next click.
 
@@
 By default \code{calculate_roc} computes the empirical ROC curve. There
-are other estimation methods out there, as I have summarized in the
+are other estimation methods out there, as summarized in the
 introduction. Any estimation method can be used, as long as the cutoff,
@@
 
-\begin{Schunk}
-\begin{Sinput}
-R> library(survivalROC)
-R> survT <- rexp(350, 1/5)
-R> cens <- rbinom(350, 1, .1)
-R> M <- -8 * sqrt(survT) + rnorm(350, sd = survT)
-R> sroc <- lapply(c(2, 5, 10), function(t){ 
-+   stroc <- survivalROC(Stime = survT, status = cens, marker = M, 
-+                        predict.time = t, method = "NNE", 
-+                        span = .25 * 350^(-.2))
-+   data.frame(TPF = stroc[["TP"]], FPF = stroc[["FP"]], 
-+              c = stroc[["cut.values"]], 
-+              time = rep(stroc[["predict.time"]], 
-+                         length(stroc[["FP"]])))
-+   })
-R> 
-R> sroclong <- do.call(rbind, sroc)
-R> 
-R> survplot <- ggplot(sroclong, aes(x = FPF, y = TPF, label = c, color = time)) + 
-+   geom_roc(labels = FALSE, stat = "identity")
-\end{Sinput}
-\end{Schunk}
+\textbackslash{}begin\{Schunk\} \textbackslash{}begin\{Sinput\}
+R\textgreater{} library(survivalROC) R\textgreater{} survT \textless{}-
+rexp(350, 1/5) R\textgreater{} cens \textless{}- rbinom(350, 1, .1)
+R\textgreater{} M \textless{}- -8 * sqrt(survT) + rnorm(350, sd = survT)
+R\textgreater{} sroc \textless{}- lapply(c(2, 5, 10), function(t)\{ +
+stroc \textless{}- survivalROC(Stime = survT, status = cens, marker = M,
++ predict.time = t, method = ``NNE'', + span = .25 * 350\^{}(-.2)) +
+data.frame(TPF = stroc{[}{[}``TP''{]}{]}, FPF = stroc{[}{[}``FP''{]}{]},
++ c = stroc{[}{[}``cut.values''{]}{]}, + time =
+rep(stroc{[}{[}``predict.time''{]}{]}, +
+length(stroc{[}{[}``FP''{]}{]}))) + \}) R\textgreater{} R\textgreater{}
+sroclong \textless{}- do.call(rbind, sroc) R\textgreater{}
+R\textgreater{} survplot \textless{}- ggplot(sroclong, aes(x = FPF, y =
+TPF, label = c, color = time)) + + geom\_roc(labels = FALSE, stat =
+``identity'') \textbackslash{}end\{Sinput\}
+\textbackslash{}end\{Schunk\}
 
@@
 element of the plot to a corresponding element of the svg markup
-language. I keep track of the names of the points and labels elements so
-that I can add interactivity using \pkg{d3.js} and
-\proglang{JavaScript}. The main interactive feature I wanted was to be
-able to display the cutoff labels at the points on the ROC curve closest
-to the mouse cursor.
+language. Interactivity is then added using \pkg{d3.js} and
+\proglang{JavaScript} to manipulate those svg elements in response to
+users' input. The main interactive feature is to display the cutoff
+labels at the points on the ROC curve closest to the mouse cursor.
 
-There are many ways to solve this with \pkg{d3.js}, but I decided to use
-Voronoi polygons to map the cursor location to the nearest point on the
-ROC curve. The idea is that for the set of cutoff points along the ROC
-curve, the \code{d3.geom.voronoi} function chain computes a set of
-polygons overlaying the plotting region such that the area of each
-polygon contains the region of the plot closest to it's corresponding
-cutoff point. Hover events are bound to the polygons so that when the
-mouse cursor moves around the plotting region, the closest point on the
-ROC curve is made visible. Similarly, click events are bound to the
-polygons so that the appropriate confidence region is made visible upon
-clicking. The svg code and all necessary \proglang{JavaScript} code is
-returned in the character string provided by
-\code{export_interactive_roc}.
+There are many ways to solve this with \pkg{d3.js}, but Voronoi polygons
+are a convenient and efficient way to map the cursor location to
+approximately the nearest point on the ROC curve. For the set of cutoff
+points along the ROC curve, the \code{d3.geom.voronoi} function chain
+computes a set of polygons overlaying the plotting region such that the
+area of each polygon contains the region of the plot closest to it's
+corresponding cutoff point. Hover events are bound to the polygons so
+that when the mouse cursor moves around the plotting region, the closest
+point on the ROC curve is made visible. Similarly, click events are
+bound to the polygons so that the appropriate confidence region is made
+visible upon clicking. The svg code and all necessary
+\proglang{JavaScript} code is returned in the character string provided
+by \code{export_interactive_roc}.
 
@@
 \code{grid.animate} function, which uses the svg \code{<animate />}
-tags. However, the available features were not sufficient for my needs,
-which is why I used \pkg{d3.js}. There are several other \proglang{R}
-packages that aim to create interactive figures. The authors of
-\pkg{animint} \citep{animint} created an extensive \proglang{JavaScript}
-library that creates plots in a similar way as \pkg{ggplot2}. A set of
-interactive features can be added to plots using \pkg{d3.js}.
-\pkg{ggvis} \citep{ggvis}, \pkg{rCharts} \citep{rcharts}, and the more
-recently released \pkg{htmlwidgets} \citep{htmlwidgets} all leverage
-existing charting libraries written in \proglang{JavaScript}.
-\pkg{qtlcharts} \citep{qtlcharts} uses a set of custom
-\proglang{JavaScript} and \pkg{d3.js} functions to visualize data from
-genetic experiments. Their general approach is to manipulate the data
-and create options in \proglang{R}, and then let the charting libraries
-or functions handle the rendering and interactivity. \pkg{plotROC} lets
-\proglang{R} do the rendering, allowing the figures to be consistent
-across print and web-based media, and retaining the distinctive
-\proglang{R} style. This also allows users to manipulate the figures
-directly in \proglang{R} to suit their needs, using tools that are more
-accessible and familiar to most \proglang{R} users. Then, the
-\proglang{JavaScript} adds a layer of interactivity to the rendered
-figures.
+tags. However, \pkg{d3.js} provides a much richer set of features. There
+are several other \proglang{R} packages that aim to create interactive
+figures. The authors of \pkg{animint} \citep{animint} created an
+extensive \proglang{JavaScript} library that creates plots in a similar
+way as \pkg{ggplot2}. A set of interactive features can be added to
+plots using \pkg{d3.js}. \pkg{ggvis} \citep{ggvis}, \pkg{rCharts}
+\citep{rcharts}, and the more recently released \pkg{htmlwidgets}
+\citep{htmlwidgets} all leverage existing charting libraries written in
+\proglang{JavaScript}. \pkg{qtlcharts} \citep{qtlcharts} uses a set of
+custom \proglang{JavaScript} and \pkg{d3.js} functions to visualize data
+from genetic experiments. Their general approach is to manipulate the
+data and record plotting options in \proglang{R}, and then pass those
+objects to the charting libraries or functions that handle the rendering
+and interactivity. \pkg{plotROC} lets \proglang{R} do the rendering,
+allowing the figures to be consistent across print and web-based media,
+and retaining the distinctive \proglang{R} style. This also allows users
+to manipulate the figures directly in \proglang{R} to suit their needs,
+using tools that are more accessible and familiar to most \proglang{R}
+users. Then, the \proglang{JavaScript} adds a layer of interactivity to
+the rendered figures.
 
@@
 \begin{Soutput}
-R version 3.2.2 (2015-08-14)
+R version 3.2.4 (2016-03-10)
 Platform: x86_64-apple-darwin13.4.0 (64-bit)
-Running under: OS X 10.10.5 (Yosemite)
+Running under: OS X 10.11.4 (El Capitan)
 
@@
 other attached packages:
-[1] survivalROC_1.0.3  tikzDevice_0.8.1   plotROC_2.0.0     
-[4] ggplot2_1.0.1.9003 xtable_1.7-4       stringr_1.0.0     
-[7] knitr_1.11
+[1] survivalROC_1.0.3 plotROC_2.0.1     ggplot2_2.1.0     xtable_1.8-2     
+[5] stringr_1.0.0     knitr_1.13
 
 loaded via a namespace (and not attached):
- [1] Rcpp_0.12.1      digest_0.6.8     grid_3.2.2       plyr_1.8.3      
- [5] gtable_0.1.2     formatR_1.2.1    magrittr_1.5     evaluate_0.8    
- [9] scales_0.3.0     highr_0.5.1      stringi_0.5-5    labeling_0.3    
-[13] tools_3.2.2      munsell_0.4.2    colorspace_1.2-6 filehash_2.3    
+ [1] Rcpp_0.12.5       digest_0.6.9      grid_3.2.4       
+ [4] plyr_1.8.3        gtable_0.2.0      formatR_1.4      
+ [7] magrittr_1.5      evaluate_0.9      scales_0.4.0     
+[10] highr_0.6         stringi_1.0-1     tikzDevice_0.10-1
+[13] labeling_0.3      tools_3.2.4       munsell_0.4.3    
+[16] colorspace_1.2-6  filehash_2.3     
 \end{Soutput}

### 2016-07-15 Paper/sachs2015plotROC.R

@@ -69,6 +69,7 @@ print(xtable(table1, label = "table1",
 
 
 ## ----figure1, fig.width = 6, fig.height = 6, message = FALSE, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves. Panel A shows a sparse ROC curve, with no design additions inside the plotting region. The plot results in more white space than anything else. It is difficult to accurately determine values without reference lines. Panel B shows a plot comparing 2 curves, with different line types and a legend. AUCs are also given in the legend. Panels C and D add gridlines, diagonal reference lines, and direct labels. \\label{figure1}"----
+
 library(plotROC)
 
 set.seed(520)
@@ -110,9 +111,8 @@ text(.2, .8, "A")
 text(.3, .4, "B")
 
 ## ----load, eval = FALSE--------------------------------------------------
-## devtools::install_github("hadley/ggplot2")
-## devtools::install_github("sachsmc/plotROC")
-## library(plotROC)
+# install.packages(c("ggplot2", "plotROC"))
+library(plotROC)
 
 ## ----shiny, eval = FALSE-------------------------------------------------
 shiny_plotROC()
